### 2022-03-05 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
initial 9

---
title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area in Canada"
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: SEES
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: SEES
    footnote: 1
address:
  - code: SEES
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes and visualises the data contained within the {TTS2016R} data-package created in `R`, the statistical computing and graphics language. In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, and the origin and destination trips. {TTS2016R} also includes the calculated car travel time from TAZ origin-destination centroid pairs. This information is useful to understand patterns of commuting in the region. We illustrate the estimation of a distance-decay curve using these data. {TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manupulation, and the final data products are detailed."

keywords: Jobs; population; travel time; impedance; Greater Toronto and Hamilton Area; Ontario, Canada; R 
classoption:
  - Royal
  - times
bibliography: bibfile
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---
```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

```{r install-data-package, include=FALSE}
if (!require("TTS2016R", character.only = TRUE)) {
      remotes::install_github("soukhova/TTS2016R",
                        build_vignettes = TRUE)
  }
```

```{r load-packages, include=FALSE, cache=FALSE}
library(TTS2016R)
library(dplyr)
library(fitdistrplus)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(sf)
library(scales)
library(stats)
library(ggspatial)
library(shadowtext)
library(tmap)
# library(ggpmisc)
# library(ggrepel)
# library(cowplot)
# library(spdep)
# library(RColorBrewer)
# library(extrafont)
# font_import()
# loadfonts(device = "win")

options(scipen = 999)
```

# Introduction

This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R). Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open]. The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data. This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018]. In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r]. The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function, showcase how it can be used to calculate accessibility to employment, and invite its use in other applications.

Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required. In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process. By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region. It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.

# Home-to-work commute data

{TTS2016R} includes counts of fully-employed population by place of residence and counts of full-time jobs by place of work aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries). Unique identifiers use the GTA06 Zoning System of TTS. The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) are organized in the form of an origin-destination table, indicative of patterns of home-to-work commute (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions). These data were retrieved from the Transportation Tomorrow Survey Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].

Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}. These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible uses. It is important to note that for simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid. The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode. Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in the descriptive statistics in Table \ref{tab:TTS-16-desc-stats}.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# grouping the planning boundaries/municipalities so they make up the 20 regions in the TTS 2016. Note: st_buffer is used as there are small existing gaps between some boundaries. st_buffer of 10 m is enough to widen all boundaries and complete the st_union without issue.
group_ggh_pd_poly <- TTS2016R::ggh_pd %>% st_buffer(10) %>% group_by(REGION) %>% 
  summarize(REGION_name = first(REGION_name),
            geometry = st_union((geometry)))

# creating an object of centroids for each region - this will be used to label polygons on the map
group_ggh_pd <- sf::st_centroid(group_ggh_pd_poly) 
points <- sf::st_coordinates(group_ggh_pd) %>% data.frame() 
group_ggh_pd <- cbind(group_ggh_pd, points)

## manually readjusting the X and Y coordinate of "County of Peterborough" and "Brant" as they overlap some cities
group_ggh_pd[group_ggh_pd$REGION_name=="Brant", "X"] <- 544000.0
group_ggh_pd[group_ggh_pd$REGION_name=="Brant", "Y"] <- 4767466

group_ggh_pd[group_ggh_pd$REGION_name=="Peterborough County", "Y"] <- 4921000
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE, out.height=400}
#plot 
ggplot() +
  geom_sf(data = TTS2016R::ggh_pd,
          fill = "gainsboro", color = "darkgray") +
  geom_sf(data = group_ggh_pd_poly, 
          color = "steelblue", fill = NA, size = 0.7) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) + 
  geom_shadowtext(data = group_ggh_pd,
                  aes(x = X, y = Y, label = REGION_name),
                  size = 2.5,
                  nudge_y = 3000,
                  nudge_x = 2000) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank())

ggsave("images/TTS16-survey-area.png")
```
```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the Greater Golden Horseshoe in Ontario, Canada.", out.width="80%", fig.align='center'}
knitr::include_graphics("images/TTS16-survey-area.png")
```

```{r creating-desc-stats-table}
#forming a complete descriptive statistic table

Statistics <- data.frame("Statistics" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))

Trips <- data.frame("OD_Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
                                   summary(od_ft_tt$trips)[[2]] %>% round(),  
                                   summary(od_ft_tt$trips)[[3]] %>% round(), 
                                   summary(od_ft_tt$trips)[[4]] %>% round(), 
                                   summary(od_ft_tt$trips)[[5]] %>% round(),
                                   summary(od_ft_tt$trips)[[6]]%>% round(),
                                   NA))

Travel_time <- data.frame("OD_Travel_time" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[2]] %>% round(),  
                                               summary(od_ft_tt$travel_time)[[3]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[4]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[5]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[6]] %>% round(),  
                                               3507))

TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(), 
                                      summary(ggh_taz$AREA)[[2]] %>% round(), 
                                      summary(ggh_taz$AREA)[[3]] %>% round(), 
                                      summary(ggh_taz$AREA)[[4]] %>% round(), 
                                      summary(ggh_taz$AREA)[[5]] %>% round(), 
                                      summary(ggh_taz$AREA)[[6]] %>% round(), 
                                      NA))

Workers <- data.frame("Workers" = c(summary(ggh_taz$workers)[[1]] %>% round(), 
                                    summary(ggh_taz$workers)[[2]] %>% round(), 
                                    summary(ggh_taz$workers)[[3]] %>% round(), 
                                    summary(ggh_taz$workers)[[4]] %>% round(), 
                                    summary(ggh_taz$workers)[[5]] %>% round(), 
                                    summary(ggh_taz$workers)[[6]] %>% round(), 
                                    NA))

Jobs <- data.frame("Jobs" = c(summary(ggh_taz$jobs)[[1]] %>% round(), 
                              summary(ggh_taz$jobs)[[2]] %>% round(), 
                              summary(ggh_taz$jobs)[[3]] %>% round(), 
                              summary(ggh_taz$jobs)[[4]] %>% round(), 
                              summary(ggh_taz$jobs)[[5]] %>% round(), 
                              summary(ggh_taz$jobs)[[6]] %>% round(), 
                              NA))

desc_stats <- cbind(Statistics, Trips, Travel_time, TAZ_Area, Workers, Jobs)

#kable tabling 
desc_stats %>%
  kable(format = "latex",
        align="lrrrrrr",
        booktabs = T,
        col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
        caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, workers, and jobs for the traffic analysis zones (TAZ) from the TTS 2016 dataset along with estimated car origin-destination travel times.") %>%
  add_header_above(c(" ", "Trips", "Car Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
  kable_styling(full_width = "T", 
                latex_options = c("scale_down"),
                position = "center")
```

## Employed individuals and jobs

The origin-destination information consists of a cross-tabulation of people who are employed full-time by place of residence (origin) and places of employment in the GGH (destination) using the GTA06 zoning system. The number of workers and jobs is not equal; the boundaries of the survey are permeable, so workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed. This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs). While the 2016 TTS survey boundaries are drawn to minimize the difference between opportunities at the destination and supply at the origins they are still not equal <!-- cite? is this true ? -->. That said, these data offer a perspective of the potential of home-based trips to places of employment and as such, the number of trips taken are equal to the number of workers in the GGH.

```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (top) and jobs (bottom) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=11}

tts_workers <- tm_shape(ggh_taz) +
  tm_fill(col = "workers", title = "Full-time employed people", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tts_jobs <- tm_shape(ggh_taz) +
  tm_fill(col = "jobs", title = "Full-time jobs", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tmap_arrange(tts_workers, tts_jobs)

```

```{r ECD-plot, fig.cap="\\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set. Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ. "}

Number <- rbind(ggh_taz$jobs %>% data.frame(),
                 ggh_taz$workers %>% data.frame())
Group <- rbind(rep("Jobs", each=length(ggh_taz$jobs)) %>% data.frame(),
                rep("Workers", each=length(ggh_taz$workers))%>% data.frame())

ecdf_data <- cbind(Number, Group)
colnames(ecdf_data) <- c("Number", "Group")

rect1 <- data.frame(xmin=115, xmax=3055, ymin=-Inf, ymax=Inf)

ggplot(ecdf_data , aes(x=Number, col=Group)) + 
  # geom_segment(aes(x = 115, y = 0, xend = 115, yend = 0.34), col = "blue", linetype=2) +
  # geom_segment(aes(x = 3055, y = 0, xend = 3055, yend = 0.94), col = "blue", linetype=2) +
  geom_rect(data = rect1, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), fill="lightblue", alpha=0.2, inherit.aes = FALSE) +
  stat_ecdf() + 
  scale_x_continuous(trans='sqrt', breaks=c(115, 3055, 10000, 20000, 30000, 40000)) + 
  scale_y_continuous(breaks=c(0, 0.25,0.34, .50, .75,0.94, 1.00)) +
  labs(x = "Opportunities per TAZ",
       y = "Cumulative probability") + 
      scale_color_manual("Opportunities", 
                         values = c("Jobs" = "Black",
                                    "Workers" = "Red"))+
  scale_linetype_manual("Opportunities",values=c(0,1)) +
  geom_point(aes(x = 115, y = 0.34), col = "blue")+
  geom_point(aes(x=3055, y=0.94), col = "blue") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5),
        legend.position = c(0.90,0.75),
        axis.line = element_line(colour = "black"),
        panel.border = element_blank()) 
```

Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ. It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of jobs-housing imbalance, which can impact accessibility in a region [@Levine1998rethinking]. Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH. The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022]. The south east border of the GGH neighbours Lake Ontario and is delineated by the urban built boundary of the Ontario Growth Plan being home to the highest density of working population in the GGH [@ontario_built_2019;@auditor_general_of_ontario_value_2021]. The northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake, and Peterborough regions which are home to lower density of worker population density population [@auditor_general_of_ontario_value_2021]. Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs.

It can also be seen that from the bottom plot in Figure \ref{fig:tts-workers-jobs-plot} that high to medium-low concentrations of jobs are often present in the same areas as workers but only when the scale is transformed. In other words, though there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers; see the following cumulative probability distribution in Figure \ref{fig:ECD-plot} in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ. This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}.

\newpage

## Calculated travel time

{TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}. This travel time corresponds to a car commute calculated using the R package {r5r} and is interpreted as the travel time for a work commute for full-time employed people in the GGH. The travel times were calculated assuming the following input parameters: a maximum travel time less than or equal to 180 mins (3 hrs) and a street network retrieved from OpenStreetMaps for the GGH area. The 3 hr threshold was selected as it captures 99% of the trip taken (see the travel times summarized in Table \ref{tab:TTS-16-desc-stats}).

It is important to note that all travel times within this data set are calculated assuming car travel and one departure time for all origins. These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies, however, it is not possible from the data retrieval system to obtain higher order tabulations that would allow us to split the data further. Thus we do not know which trips are made with non-car modes nor exact departure times in these tables. Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity, we assume that all trips are taken by one-time departure car trip.

```{r data-for-impedance}
# remove all NA trips from dataset and set all 0min travel times to 0.1 min
od_ft_tt  <- TTS2016R::od_ft_tt %>% 
  filter( !is.na(travel_time)) %>% 
  mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))

all_tt <- od_ft_tt  %>% 
  dplyr::select(trips, travel_time)

all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
```

```{r fitting-impedance}
#fitting impedance function
gamma_ <- fitdistrplus::fitdist(data=all_tt, "gamma", method="mle", optim.method="Nelder-Mead") 
```

```{r calc-for-accessibility-GGH}
# transfer calibrated impedance function values to OD matrix
od_ft_tt <- od_ft_tt %>%
  mutate(f = dgamma(travel_time, gamma_$estimate["shape"], gamma_$estimate["rate"]))

#add the number of jobs and workers to the od_ft_tt matrix
od_ft <- od_ft_tt %>% merge(ggh_taz %>% dplyr::select(GTA06, workers) %>% st_drop_geometry(),
                   by.x = "Origin", by.y="GTA06", all.x = TRUE)

od_ft <- od_ft %>% merge(ggh_taz %>% dplyr::select(GTA06, jobs) %>% st_drop_geometry(),
                   by.x = "Destination", by.y="GTA06", all.x = TRUE)

#calculate accessibility for workers from any origin to jobs in Toronto 
GGH_c_accessibility <- od_ft %>% 
  mutate(GGH_A_ij = f * jobs) %>%
  group_by(Origin) %>%
  summarise(GGH_A_i = sum(GGH_A_ij, na.rm = T),
            GGH_sum_tt_i = sum(travel_time),
            GGH_tt_trips_i = mean(travel_time),
            GGH_sum_f_i = sum(f),
            GGH_f_trips_i = mean(f))

#Merge TO accessibly calculation to the ggh_taz:
GGH_taz_acc <- ggh_taz %>% merge(GGH_c_accessibility, by.x=c("GTA06"), by.y=c("Origin"), all.x=T) 
```

```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (top) and average worker travel time (bottom) for each TAZ in the GGH.", fig.width=8, fig.height=11, message=FALSE}

tts_total_tt <- tm_shape(GGH_taz_acc) +
  tm_fill(col = "GGH_sum_tt_i", title = "Total travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tts_tt_per_trip <- tm_shape(GGH_taz_acc) +
  tm_fill(col = "GGH_tt_trips_i", title = "Avg. travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tmap_arrange(tts_total_tt, tts_tt_per_trip)
```

\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time (min) resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}). However, when the average travel time per trip in each TAZ is presented, the spatial distribution is distinct from all other plots presented so far. We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA) (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) and Niagara and Waterloo, the average travel times are moderately low. Further from these areas, travel times are higher. Interestingly, even in eastern areas (e.g., Peterborough) with high employment and high job concentration, average travel time is higher than within the GTHA.

## Calibrating an impedance function

Impedance functions are useful to understand mobility behavior and are part, for instance, of gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and accessibility analysis in many applications [@hansen_how_1959;@levinson_accessibility_1998;@talen_assessing_1998;@reggiani_accessibility_2011;@paez_jobs_2013;@kuai_examining_2017; @barboza_balancing_2021]. An origin destination matrix and a cost matrix (i.e., with travel times) can be used in combination to estimate impedance functions. An impedance function $f(\cdot)$ depends on the cost of travel between locations $i$ and $j$ $c_{ij}$; it usually is a monotonically decreasing function, although sometimes the function can increase to reflect patterns of separation between activities; for instance, separation of land uses means that very short commuting trips are relatively rare. There can be sometimes also fluctuations due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration.

A useful technique to calibrate an impedance function is to use the trip length distribution (TLD) as measured from origin-destination data [@horbachov_theoretical_2018; @batista_estimation_2019]. The TLD is the representation of the likelihood that a proportion of trips are taken at a specific travel cost. In our data set, where we assume cost is travel time, the impedance function maps low travel times to higher proportions of trips, and high travel times are mapped to low proportion of trips.

In the GGH data presented, the empirical TLD (i.e., proportion of trips taken vs. travel time in minutes) is fitted to a density distribution using maximum likelihood techniques and the Nelder-Mead method for direct optimization available within the `fitdistrplus` package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics seen in Figure \ref{fig:TLD-Gamma-plot} and Figure \ref{fig:plot-cullen-frey}, the gamma distribution is selected for the presented data.

```{r save-impedance-plot, include=FALSE}
# For some reason plot(gamma_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("images/impedance_function.png")
plot(gamma_)
dev.off()
```

```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots"}
knitr::include_graphics("images/impedance_function.png")
```

The resulting calibrated impedance function is given in the following general form where the estimated 'shape' is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for }	0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}

```{r create-cullen-frey-data}
# remove all NA trips from dataset and set all 0min travel times to 0.1min
od_ft_tt  <- od_ft_tt %>% filter( !is.na(travel_time)) %>% mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))
all_tt <- od_ft_tt  %>% dplyr::select(trips, travel_time)
all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
```

```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."}
fitdistrplus::descdist(data=all_tt) 
```

# Accessibility to employment

As noted above, impedance functions are an essential component of accessibility analysis. Equation (\ref{eq:conventional-accessibility}) shows that accessibility $A_i$ is the weighted sum of opportunities that can be reached from location $i$, given the cost of travel $c_{ij}$. Summing the opportunities in the neighborhood of $i$ as defined by the impedance function $f(\cdot)$, provides estimates of the number of opportunities that can be reached from $i$ at a certain cost.

\begin{equation}
\label{eq:conventional-accessibility}
A_i = \sum_{j=1}^JO_jf(c_{ij})
\end{equation}

\noindent where:

-   $A$ is accessibility. 
-   $i$ is a set of origin locations.
-   $j$ is a set of destination locations.
-   $O_j$ is the number of opportunities at location $j$. These are opportunities for activity and add some sort of *supply* to the area;
-   $c_{ij}$ is a measure of the cost of moving between $i$ and $j$
-   $f(\cdot)$ is an impedance function of $c_{ij}$.

```{r access-per-worker}
GGH_taz_acc <- GGH_taz_acc %>%
  mutate(avg_A_per_W = GGH_A_i/workers)
```

```{r plot-access-SA-GGH-TTS, fig.cap="\\label{fig:plot-access-SA-GGH-TTS}Accessibility (top) and normalized accessibility (bottom) to employment in the GGH area", fig.width=8, fig.height=11, message=FALSE}
## accessibility
access_TTS_GGH <- tm_shape(GGH_taz_acc) +
  tm_fill(col = "GGH_A_i", title = "Accessibility \n(A_i)", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

access_normalized_TTS_GGH <- tm_shape(GGH_taz_acc) +
  tm_fill(col = "avg_A_per_W", title = "Average accessibility \nper worker", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tmap_arrange(access_TTS_GGH, access_normalized_TTS_GGH)
```
\newpage

Accessibility is a property of the origin (i.e., origin TAZ in our data), the landscape of opportunities (i.e., the TAZ destinations in our data), and transportation infrastructure (i.e., the travel time table). Figure \ref{fig:plot-access-SA-GGH-TTS} presents the accessibility estimates that result from the impedance function calibrated in the preceding section. The top plot, that of the raw accessibility values, presents a distinct radial trend where the majority of TAZ in and around Toronto have high accessibility values and values gradually decrease in TAZ which are further from Toronto's boundary. Alternatively, the bottom plot in Figure \ref{fig:plot-access-SA-GGH-TTS} present the worker-normalized accessibility values. In other words, the accessibility value for each TAZ is divided by the number of workers in each TAZ. The patterns in this plot are similar to the Greater Toronto Area radial trend observed in the top plot but are less extreme (i.e., smaller spread).

<!-- room to critique accessibility or suggestion how we need alternatives to comprehend this measure..? -->

# Concluding remarks

The open data product introduced in this paper shares tables for worker-to-employment data from the 2016 TTS aggregated by TAZ. In addition, inter-centroid travel time tables are calculated, and the planning/municipality boundaries are included to compliment the 2016 TTS data. This open data product, {TTS2016R}, is freely available to explore in an `R` environment. One possible use of these data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.

New digital formats are increasingly complex and the explanation of the methods often do not concisely and intelligibly fit within the confines of a traditional article. With this motivation, we invite all who are interested to use {TTS2016R} to explore the worker-employment patterns contained in the {TTS2016R} package. In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own. We see this product as providing open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.

\newpage

# References {#references .unnumbered}

### 2022-03-05 Manuscript-Data-Package/Manuscript-Data-Package.tex
initial 9

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\setcitestyle{aysep={,}}

\title{TTS2016R: A data set to study population and employment patterns
from the 2016 Transportation Tomorrow Survey (TTS) in the Greater
Toronto and Hamilton Area in Canada}

\runninghead{}

\author{Anastasia Soukhov\affilnum{}, Antonio Paez\affilnum{}}

\affiliation{\affilnum{}{}}

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data-package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe area in Canada retrieved from the 2016 Transportation
Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ),
the number of people who are employed full-time per TAZ, the number of
jobs per TAZ, and the origin and destination trips. \{TTS2016R\} also
includes the calculated car travel time from TAZ origin-destination
centroid pairs. This information is useful to understand patterns of
commuting in the region. We illustrate the estimation of a
distance-decay curve using these data. \{TTS2016R\} can be freely
downloaded and explored at: \url{https://github.com/soukhova/TTS2016R}
where the documentation and code involved in data creation,
manupulation, and the final data products are detailed.
\end{abstract}

\keywords{Jobs; population; travel time; impedance; Greater Toronto and
Hamilton Area; Ontario, Canada; R}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is
an \texttt{R} data package which currently consists of three objects
which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or
are curated to facilitate the use and analysis of TTS data. This package
includes person-to-jobs origin-destinations, traffic analysis zone (TAZ)
boundaries, and planning/municipality boundaries for the Greater Golden
Horse area (GGH) located in southern Ontario, Canada
\citep{data_management_group_tts_2018}. In addition, the package
includes TAZ centroid-to-centroid travel times by car computed using
package \{r5r\} \citep{Pereira2021r5r}. The aim of this paper is to walk
readers through the empirical home-based work commute data set,
illustrate the calculation of an impedance function, showcase how it can
be used to calculate accessibility to employment, and invite its use in
other applications.

Data from the TTS are in principle available to the public but are not
fully open, since permission to access the data retrieval system is
required. In addition, the raw data can be technically demanding,
cumbersome to work with, and requires multiple software to process. By
pre-processing the data in a \texttt{R} environment, \{TTS2016R\} offers
a slice of the TTS data useful to understand patterns of commuting to
work in the region. It also provides open infrastructure for additional
TTS or complimentary data sets to be amended by the authors or wider
open-source community in the future.

\hypertarget{home-to-work-commute-data}{%
\section{Home-to-work commute data}\label{home-to-work-commute-data}}

\{TTS2016R\} includes counts of fully-employed population by place of
residence and counts of full-time jobs by place of work aggregated by
TAZ (n=3,764 within the survey boundaries). Unique identifiers use the
GTA06 Zoning System of TTS. The number of jobs (3,081,900) and workers
(3,446,957) are organized in the form of an origin-destination table,
indicative of patterns of home-to-work commute (there are 3,446,957
potential interactions). These data were retrieved from the
Transportation Tomorrow Survey Data Retrieval System on October 28, 2021
and reflect the potential interaction of full-time employed people and
jobs within the GGH survey boundaries shown in Figure
\ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology
\citep{data_management_group_tts_2018}.

Also included in \{TTS2016R\} are travel times and cost of travel from
origin to destination by car; travel times are calculated using the
\texttt{R} package \{r5r\}. These travel times are useful to estimate
the cost of travel and to calculate impedance functions, among other
possible uses. It is important to note that for simplicity, all
interactions within \{TTS2016R\} are assumed to be taken by car, and the
travel time is calculated from an origin TAZ centroid to a destination
TAZ centroid. The centroid is snapped to the nearest street line by
\texttt{r5r} and the travel time is calculated for all trips assuming a
car travel mode. Additionally, only travel times less than or equal to
180 mins (3 hrs) are calculated; this threshold represents 99\% of
trip's travel times which are summarized in the descriptive statistics
in Table \ref{tab:TTS-16-desc-stats}.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{images/TTS16-survey-area}

}

\caption{\label{fig:TTS-16-survey-area}The TTS 2016 study area within the Greater Golden Horseshoe in Ontario, Canada.}\label{fig:TTS-16-survey-area}
\end{figure}

\begin{table}

\caption{\label{tab:creating-desc-stats-table}\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, workers, and jobs for the traffic analysis zones (TAZ) from the TTS 2016 dataset along with estimated car origin-destination travel times.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
\multicolumn{1}{r}{ } & \multicolumn{1}{r}{Trips} & \multicolumn{1}{r}{Car Travel Time} & \multicolumn{1}{r}{Area} & \multicolumn{1}{r}{Workers} & \multicolumn{1}{r}{Jobs} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
  & (\#) & (min) & (km\textasciicircum{}2) & (\#) & (\#)\\
\midrule
Min. & 1 & 0 & 0 & 0 & 0\\
1st Qu. & 14 & 13 & 1 & 25 & 64\\
Median & 22 & 20 & 1 & 464 & 244\\
Mean & 33 & 23 & 7 & 916 & 819\\
3rd Qu. & 38 & 30 & 3 & 1378 & 700\\
\addlinespace
Max. & 1129 & 179 & 879 & 8491 & 41821\\
NA's & NA & 3507 & NA & NA & NA\\
\bottomrule
\end{tabu}}
\end{table}

\hypertarget{employed-individuals-and-jobs}{%
\subsection{Employed individuals and
jobs}\label{employed-individuals-and-jobs}}

The origin-destination information consists of a cross-tabulation of
people who are employed full-time by place of residence (origin) and
places of employment in the GGH (destination) using the GTA06 zoning
system. The number of workers and jobs is not equal; the boundaries of
the survey are permeable, so workers who reside within the boundaries
but travel outside of the boundaries are counted as workers within an
origin TAZ, while jobs in TAZ that are filled by workers who reside
outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
3,081,900 jobs). While the 2016 TTS survey boundaries are drawn to
minimize the difference between opportunities at the destination and
supply at the origins they are still not equal . That said, these data
offer a perspective of the potential of home-based trips to places of
employment and as such, the number of trips taken are equal to the
number of workers in the GGH.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (top) and jobs (bottom) in each TAZ in the GGH area as specified in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/ECD-plot-1} \caption{\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set. Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ. }\label{fig:ECD-plot}
\end{figure}

Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers
and jobs per TAZ. It can be observed that the spatial distribution of
jobs and workers is unequal, which is indicative of jobs-housing
imbalance, which can impact accessibility in a region
\citep{Levine1998rethinking}. Workers are concentrated in TAZ within the
center of the GGH and along the south-east and northern boarder of the
GGH. The center of the GGH corresponds to the Greater Toronto Area (GTA)
which is the most densely populated area in southern Ontario
\citep{statistics_canada_daily_2022}. The south east border of the GGH
neighbours Lake Ontario and is delineated by the urban built boundary of
the Ontario Growth Plan being home to the highest density of working
population in the GGH
\citep{ontario_built_2019, auditor_general_of_ontario_value_2021}. The
northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake,
and Peterborough regions which are home to lower density of worker
population density population
\citep{auditor_general_of_ontario_value_2021}. Conversely, the spread of
jobs in the GGH is lower than the number of workers indicating
population is more spatially dispersed than jobs.

It can also be seen that from the bottom plot in Figure
\ref{fig:tts-workers-jobs-plot} that high to medium-low concentrations
of jobs are often present in the same areas as workers but only when the
scale is transformed. In other words, though there is a higher number of
TAZ with no workers than zones with no jobs (i.e., 791 TAZ with no
workers : 396 TAZ with no jobs) and the mean of workers per TAZ is
higher than the mean of jobs (i.e., 916 workers : 819 jobs) the number
TAZ with an extreme number of jobs at the highest and lowest percentiles
is significantly higher than the number of workers; see the following
cumulative probability distribution in Figure \ref{fig:ECD-plot} in
which the 94th to 100th percentile and the 0th to 34th percentile of
jobs in TAZ is higher than the number of workers in TAZ. This means that
between these ranges, TAZ have a higher number of workers than they do
jobs, echoing the more even spatial distribution of workers observed in
Figure \ref{fig:tts-workers-jobs-plot}.

\newpage

\hypertarget{calculated-travel-time}{%
\subsection{Calculated travel time}\label{calculated-travel-time}}

\{TTS2016R\} also includes travel time data for each home-to-work trip
as displayed in Figure \ref{fig:plot-tt-ttpertrip}. This travel time
corresponds to a car commute calculated using the R package \{r5r\} and
is interpreted as the travel time for a work commute for full-time
employed people in the GGH. The travel times were calculated assuming
the following input parameters: a maximum travel time less than or equal
to 180 mins (3 hrs) and a street network retrieved from OpenStreetMaps
for the GGH area. The 3 hr threshold was selected as it captures 99\% of
the trip taken (see the travel times summarized in Table
\ref{tab:TTS-16-desc-stats}).

It is important to note that all travel times within this data set are
calculated assuming car travel and one departure time for all origins.
These assumptions are not completely realistic since we know a small
proportion of trips are taken by non-car modes and travel time
departures varies, however, it is not possible from the data retrieval
system to obtain higher order tabulations that would allow us to split
the data further. Thus we do not know which trips are made with non-car
modes nor exact departure times in these tables. Though modal split and
travel times can be estimated through other methods
\citep[e.g.,][]{allen_suburbanization_2021, higgins2021changes}, for
simplicity, we assume that all trips are taken by one-time departure car
trip.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (top) and average worker travel time (bottom) for each TAZ in the GGH.}\label{fig:plot-tt-ttpertrip}
\end{figure}

\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time (min) resembles the spatial trend distribution in the number
of employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}). However, when the average travel time
per trip in each TAZ is presented, the spatial distribution is distinct
from all other plots presented so far. We can see that in areas around
the south-eastern border that make up the Greater Toronto and Hamilton
Area (GTHA) (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) and
Niagara and Waterloo, the average travel times are moderately low.
Further from these areas, travel times are higher. Interestingly, even
in eastern areas (e.g., Peterborough) with high employment and high job
concentration, average travel time is higher than within the GTHA.

\hypertarget{calibrating-an-impedance-function}{%
\subsection{Calibrating an impedance
function}\label{calibrating-an-impedance-function}}

Impedance functions are useful to understand mobility behavior and are
part, for instance, of gravity models of spatial interaction
\citep{wilson1971, haynes_gravity_1985} and accessibility analysis in
many applications
\citep{hansen_how_1959, levinson_accessibility_1998, talen_assessing_1998, reggiani_accessibility_2011, paez_jobs_2013, kuai_examining_2017, barboza_balancing_2021}.
An origin destination matrix and a cost matrix (i.e., with travel times)
can be used in combination to estimate impedance functions. An impedance
function \(f(\cdot)\) depends on the cost of travel between locations
\(i\) and \(j\) \(c_{ij}\); it usually is a monotonically decreasing
function, although sometimes the function can increase to reflect
patterns of separation between activities; for instance, separation of
land uses means that very short commuting trips are relatively rare.
There can be sometimes also fluctuations due to hierarchical patterns,
where travelers bypass opportunities in favor of more distant
destinations that offer economies of agglomeration.

A useful technique to calibrate an impedance function is to use the trip
length distribution (TLD) as measured from origin-destination data
\citep{horbachov_theoretical_2018, batista_estimation_2019}. The TLD is
the representation of the likelihood that a proportion of trips are
taken at a specific travel cost. In our data set, where we assume cost
is travel time, the impedance function maps low travel times to higher
proportions of trips, and high travel times are mapped to low proportion
of trips.

In the GGH data presented, the empirical TLD (i.e., proportion of trips
taken vs.~travel time in minutes) is fitted to a density distribution
using maximum likelihood techniques and the Nelder-Mead method for
direct optimization available within the \texttt{fitdistrplus} package
\citep{fitdistrplus_2015}. Based on goodness-of-fit criteria and
diagnostics seen in Figure \ref{fig:TLD-Gamma-plot} and Figure
\ref{fig:plot-cullen-frey}, the gamma distribution is selected for the
presented data.

\begin{figure}
\includegraphics[width=1\linewidth]{images/impedance_function} \caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
\end{figure}

The resulting calibrated impedance function is given in the following
general form where the estimated `shape' is \(\alpha\) = 2.019, the
estimated `rate' is \(\beta\) = 0.094 , and \(\Gamma(\alpha)\) is
defined in Equation (\ref{gamma-dist}).

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-cullen-frey-1} \caption{\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times.}\label{fig:plot-cullen-frey}
\end{figure}

\begin{verbatim}
summary statistics
------
min:  0.1   max:  179 
median:  18 
mean:  21.4344 
estimated sd:  14.61254 
estimated skewness:  1.703326 
estimated kurtosis:  8.353363 
\end{verbatim}

\hypertarget{accessibility-to-employment}{%
\section{Accessibility to
employment}\label{accessibility-to-employment}}

As noted above, impedance functions are an essential component of
accessibility analysis. Equation (\ref{eq:conventional-accessibility})
shows that accessibility \(A_i\) is the weighted sum of opportunities
that can be reached from location \(i\), given the cost of travel
\(c_{ij}\). Summing the opportunities in the neighborhood of \(i\) as
defined by the impedance function \(f(\cdot)\), provides estimates of
the number of opportunities that can be reached from \(i\) at a certain
cost.

\begin{equation}
\label{eq:conventional-accessibility}
A_i = \sum_{j=1}^JO_jf(c_{ij})
\end{equation}

\noindent where:

\begin{itemize}
\tightlist
\item
  \(A\) is accessibility.
\item
  \(i\) is a set of origin locations.
\item
  \(j\) is a set of destination locations.
\item
  \(O_j\) is the number of opportunities at location \(j\). These are
  opportunities for activity and add some sort of \emph{supply} to the
  area;
\item
  \(c_{ij}\) is a measure of the cost of moving between \(i\) and \(j\)
\item
  \(f(\cdot)\) is an impedance function of \(c_{ij}\).
\end{itemize}

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-access-SA-GGH-TTS-1} \caption{\label{fig:plot-access-SA-GGH-TTS}Accessibility (top) and normalized accessibility (bottom) to employment in the GGH area}\label{fig:plot-access-SA-GGH-TTS}
\end{figure}
\newpage

Accessibility is a property of the origin (i.e., origin TAZ in our
data), the landscape of opportunities (i.e., the TAZ destinations in our
data), and transportation infrastructure (i.e., the travel time table).
Figure \ref{fig:plot-access-SA-GGH-TTS} presents the accessibility
estimates that result from the impedance function calibrated in the
preceding section. The top plot, that of the raw accessibility values,
presents a distinct radial trend where the majority of TAZ in and around
Toronto have high accessibility values and values gradually decrease in
TAZ which are further from Toronto's boundary. Alternatively, the bottom
plot in Figure \ref{fig:plot-access-SA-GGH-TTS} present the
worker-normalized accessibility values. In other words, the
accessibility value for each TAZ is divided by the number of workers in
each TAZ. The patterns in this plot are similar to the Greater Toronto
Area radial trend observed in the top plot but are less extreme (i.e.,
smaller spread).

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

The open data product introduced in this paper shares tables for
worker-to-employment data from the 2016 TTS aggregated by TAZ. In
addition, inter-centroid travel time tables are calculated, and the
planning/municipality boundaries are included to compliment the 2016 TTS
data. This open data product, \{TTS2016R\}, is freely available to
explore in an \texttt{R} environment. One possible use of these data, as
showcased in this paper, is the calibration of impedance functions which
in turn can be used for accessibility analysis.

New digital formats are increasingly complex and the explanation of the
methods often do not concisely and intelligibly fit within the confines
of a traditional article. With this motivation, we invite all who are
interested to use \{TTS2016R\} to explore the worker-employment patterns
contained in the \{TTS2016R\} package. In the spirit of novel and
original research, we hope readers value the efforts made to detail the
data in order to improve transparency in our work and encourage others
to replicate and, hopefully, inspire research of their own. We see this
product as providing open infrastructure for additional TTS or
complimentary data sets to be amended by the authors or wider
open-source community in the future.

\newpage

\bibliographystyle{sageh}
\bibliography{bibfile}

\end{document}

### 2022-03-05 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
initial 9

---
title: "TTS Origin-Destination Data by Full-Time Employment"
output: 
---

In this notebook we process a cross-tabulation of origin and destinations by people employed full-time in the Greater Golden Horseshoe (GGH) in 2016.

The source data is in file `tts-2016-OD-FT-Employment.txt`, which was retrieved from the Transportation Tomorrow Survey Data Retrieval System on October 28, 2021.

The centroids of all work zones (origins) and full-time (FT) employment zones (destinations) are calculated.

## Preliminaries

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
```

## Traffic Analysis Zones

Read the boundaries of the traffic analysis zones:
```{r}
ggh_taz <- st_read("TTS16-data-inputs/OD-by-FT-Employment/tts06_83_region.shp")
```

Project object:
```{r}
ggh_taz <- ggh_taz %>%
  st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
```

Extract the IDs for all TAZ:
```{r}
ghta_taz_id <- ggh_taz$GTA06
```

## Planning boundaries

Read the boundaries of the traffic analysis zones:
```{r}
ggh_pd <- st_read("TTS16-data-inputs/Boundaries/tts06_pd_83.shp")
```

Project object:
```{r}
ggh_pd <- ggh_pd %>%
  st_transform(crs = 32617)
```

Adding region names to the object as they are not included in the origin data:
(http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf):
```{r}
ggh_pd <- ggh_pd %>%
  mutate(REGION_name = 
           ifelse(REGION == 1, "Toronto", 
                  ifelse(REGION == 2, "Durham", 
                         ifelse(REGION == 3, "York", 
                                ifelse(REGION == 4, "Peel", 
                                       ifelse(REGION == 5, "Halton", 
                                              ifelse(REGION == 6, "Hamilton", 
                                                     ifelse(REGION == 11, "Niagara", 
                                                            ifelse(REGION == 12, "Waterloo", 
                                                                   ifelse(REGION == 13, "Guelph", 
                                                                          ifelse(REGION == 14, "Wellington", 
                                                                                 ifelse(REGION == 15, "Orangeville", 
                                                                                        ifelse(REGION == 16, "Barrier", 
                                                                                               ifelse(REGION == 17, "Simcoe", 
                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
                                                                                                             ifelse(REGION == 19, "Peterborough City", 
                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
                                                                                                                           ifelse(REGION == 21, "Orillia", 
                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
                                                                                                                                         ifelse(REGION == 23, "Brantford", 
                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
```

## Origin destination matrices for work and shop

Read trip tables. The query was for "trips" by zone of residence, zone of work, and trip purpose (Work):
```{r}
od_ft <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

Find the position in the table where the information on work trips begins:
```{r}
idx <- which(od_ft$X1 == "TABLE    : emp_stat (Full time)")
```

Slice the table to obtain the work trips (remove heading information):
```{r}
od_ft <- od_ft %>% 
  slice((idx+2):n())
```

Separate the zone identifiers and the trips, and convert to numeric:
```{r}
od_ft <- od_ft %>%
  separate(X1, into = c("Zones", "Persons"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Persons = as.numeric(Persons))
```

The resulting object `od_ft` has the number of individuals who are employed full-time by place of residence (Origin) and place of work (Destination) using the GTA06 zoning system.

## Workers and Jobs

The od table can be used to calculate the workers and jobs by traffic analysis zone:
```{r}
workers_ft <- od_ft %>%
  group_by(Origin) %>%
  summarize(workers = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Origin,
            workers)

jobs_ft <- od_ft %>%
  group_by(Destination) %>%
  summarize(jobs = sum(Persons),
            .groups = "drop") %>%
  transmute(GTA06 = Destination,
            jobs)

```

Join workers and jobs to simple features object:
```{r}
ggh_taz <- ggh_taz %>%
  # Join workers
  left_join(workers_ft,
            by = "GTA06") %>%
  # Join jobs
  left_join(jobs_ft,
            by = "GTA06") %>%
  # Replace NAs with zeros
  mutate(workers = replace_na(workers, 0),
         jobs = replace_na(jobs, 0))
```

Plot worker population in the ggh:
```{r}
ggplot() +
  geom_sf(data = ggh_taz,
          aes(fill = workers),
          color = NA) +
  scale_fill_fermenter(direction = 1)
```

Plot job distribution in the ggh:
```{r}
ggplot() +
  geom_sf(data = ggh_taz,
          aes(fill = jobs),
          color = NA) +
  scale_fill_fermenter(direction = 1)
```

## Creating input for travel-time OD calculations

Create centroids for each feature in ggh_taz:
```{r}
ggh_taz_cent <- st_centroid(ggh_taz) 
```

Attach the centroid point geometry to the origin object (workers) and the destination object (jobs)
```{r}
#workers (origins)
workers_ft_cent <- workers_ft %>% merge(ggh_taz_cent %>% select(GTA06), by="GTA06") %>% st_sf

work_origins <- cbind(workers_ft_cent, st_coordinates(st_transform(workers_ft_cent, crs = 4326))) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
  select(id, lon, lat) %>% as.data.frame()

```

```{r}
#jobs (destinations)
jobs_ft_cent <- jobs_ft %>% merge(ggh_taz_cent %>% select(GTA06), by="GTA06") %>% st_sf

job_destinations <- cbind(jobs_ft_cent, st_coordinates(st_transform(jobs_ft_cent, crs = 4326))) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
  select(id, lon, lat)

```

## saving inputs for OD travel time calculations
```{r}
save("work_origins", file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/work_origins.Rdata")
save("job_destinations", file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/job_destinations.Rdata")
```

## saving inputs for later processing
```{r}
save("ggh_taz", file = "TTS16-data-inputs/OD-by-FT-Employment/ggh_taz.Rdata")
save("ggh_pd", file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
save("od_ft", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
```

### 2022-03-05 data-raw/00-Toy-Example-Data-Preparation.Rmd
initial 9

---
title: "Toy example dataset"
output: 
---
In this notebook we provide a toy dataset that is used to calculate both conventional accessibility and _spatial available_ accessibility. This is a more smaller and theoretical alternative dataset that can be used in place the the real-world TTS-16.

Libraries:
```{r message=FALSE}
library(sf)
library(tidyverse)
```

Load data for example:
```{r}
load("toy-example-data-inputs/od_table.rda")
toy_od_table <- od_table
load("toy-example-data-inputs/simulated_data.rda")
toy_sim_zones <- simulated_data
load("toy-example-data-inputs/trips.rda")
toy_trips <- trips
```

Remove unneeded columns and add trips to OD:
```{r}
toy_od_table <- toy_od_table %>% 
  dplyr::select(-c("f", "V_ij", "W_ij", "V_ij_r")) %>%
  merge(toy_trips %>%
          dplyr::select(Origin, Destination, trips),
               by = c("Origin", "Destination"), all.x=TRUE)
```

Save it to /data folder:
```{r}
usethis::use_data(toy_od_table, overwrite = TRUE)
usethis::use_data(toy_sim_zones, overwrite = TRUE)
```

### 2022-03-05 data-raw/01-TTS16-Calculating-Travel-Time-OD.Rmd
initial 9

---
title: "01 Calculating Travel Time for TTS16 OD"
output: 
---

In this notebook, we calculate the car time from origins (centroids of work zones) to destinations (centroids of full-time employment zones) for a one-way morning commute for the golden greater horse area (yellow) as specified by the Transportation Tomorrow Survey (TTS) 2016 dataset. This calculation is performed using [`r5r`](https://github.com/ipeaGIT/r5r) package.

![](TTS16-data-inputs/OD-by-FT-Employment/Greater-Golden-Horseshoe-Map.png)
We then we save all objects (the travel time OD matrix, the geometries, IDs, and number of workers and jobs within each traffic analysis zones (TAZ), and trips from work to job) as an intermediate object for further data processing in the next notebook '02-TTS-16-Imp-Function-Trip-Length-Distribution.Rmd'.

Libraries:
```{r setup, include=FALSE}
library(disk.frame)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(httr)
library(knitr)
#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
library(kableExtra)
# install.packages("devtools")
# install.packages("Rtools")
# devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
library(lodes)
library(progress)
library(purrr)
library(r5r)
library(sf)
library(stplanr)
library(tidycensus)
library(tidyr)
library(tmap)
library(zoo) # for rollapplyr

# setup options
setup_disk.frame() # setup disk.frame parallel computing
options(future.globals.maxSize = Inf) # allow disk.frame threads to talk to eachother without limits
options(java.parameters = "-Xmx6G") # assign memory for r5r to work with
tmap_mode("plot")
```

##--- Prepare Input Data
loading the destination points and origins points:
```{r}
load(file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/work_origins.Rdata")
load(file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/job_destinations.Rdata")
```

##--- Download OSM Network Data

```{r set up r5 path, include=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("TTS16-data-inputs/Travel-Time-Calculations/r5_graph")
r5_path <- file.path("TTS16-data-inputs/Travel-Time-Calculations/r5_graph")

# create data directories
dir.create("TTS16-data-inputs/Travel-Time-Calculations/data") # for storing data
dir.create("TTS16-data-inputs/Travel-Time-Calculations/df") # for storing od matrix as a disk.frame
dir.create("TTS16-data-inputs/Travel-Time-Calculations/r5_graph") # for the r5 network graph
r5_path <- file.path("TTS16-data-inputs/Travel-Time-Calculations/r5_graph")
```

```{r download data, include=FALSE, eval=FALSE}
#downloading ontario osm in the correct format
download.file(url = paste0("https://download.geofabrik.de/north-america/canada/ontario-latest.osm.pbf"),
              destfile = file.path(r5_path, "osm.pbf"),
              mode = "wb")

```

##--- Set Up R5 Routing

```{r build graph, include = FALSE}
r5_ONT <- setup_r5(data_path = r5_path, verbose = FALSE)
```

##--- Calculate OD Matrices

Car travel time - from all jobs to all work locations
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(work_origins)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(work_origins,
                          outdir = "TTS16-data-inputs/Travel-Time-Calculations/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_ONT,
                          origins = Orig_chunk,
                          destinations = job_destinations,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-10-20 07:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 180)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "TTS16-data-inputs/Travel-Time-Calculations/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car1 <- as.data.frame(output_df)

save("output_OD_car1", file = "TTS16-data-inputs/Travel-Time-Calculations/results/output_OD_car1.Rdata")
#I made the cut off 180 minutes, so I anticipate a small proportion of trips will not have a travel time calculated. As a note, this code took me 260 mins to execute.
```

Java garbage collector:
```{r}
stop_r5(r5_ONT)
rJava::.jgc(R.gc = TRUE)
```

##--- Joining OD Travel time matrix with taken OD trips

Load:
```{r}
#Trips taken from origin to destination and associated people
load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
```

```{r}
od_ft_tt <- od_ft %>% 
  merge(output_OD_car1, by.x=c("Origin", "Destination"), by.y=c("fromId", "toId"), all.x=T) %>%
  rename(trips = Persons)

rm(output_OD_car1)
summary(od_ft_tt)
```

Travel time is between 0-179 min with 3507 OD pairs (3.5%) having NA travel time (presumably a travel time greater than 180). This is not a significant number of OD pairs but.. how many trips? Let's check:
```{r}
junk1 <- od_ft_tt %>% filter(is.na(travel_time)) %>% summarize(trips = sum(trips))
junk2 <- od_ft_tt %>% summarize(trips = sum(trips))

junk1/junk2
```

10.9% of trips are NA travel time, that is pretty large. This seems a bit large but we will check the travel time length distribution

##--- Saving inputs for processing in the next notebook

See '02-TTS-16-Imp-Function-Trip-Length-Distribution.Rmd' for further analysis:
```{r}
save("od_ft_tt", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
```

### 2022-03-05 data-raw/02-TTS16-Imp-Function-Trip-Length-Distribution.Rmd
initial 9

---
title: "R Notebook"
output:
---

Load packages:
```{r}
library(ggplot2)
library(dplyr)
library(MASS)

# install.packages("fitdistrplus")
# install.packages("logspline")
library(fitdistrplus)
library(logspline)

options(scipen = 999)
```

##-- Exploring the OD travel time data

```{r}
load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
summary(od_ft_tt)
```
Let's explore all the NA values...

All NA travel times are 180 min or greater OR belong to a destination which is "external undefined" or "no usual location" or "unknown" (IDs 9998, 8888, 9999) according to the TTS data codes (source?); for the time being, lets set the trips which we know are 180 min or greater to "180min" and the other unknown destination trips to "300". We will see how it impacts the trip length distribution next chunk.
```{r}
od_ft_tt <- od_ft_tt %>% mutate(travel_time = ifelse( is.na(travel_time) & Destination != "9998" & Destination != "8888" & Destination != "9999", 180, travel_time),
                                travel_time = ifelse( is.na(travel_time) & (Destination == "9998" | Destination == "8888" | Destination == "9999"), 300, travel_time))
summary(od_ft_tt)
```

Trip length distribution (cutting the travel_time into 150 intervals and summing all trips within each interval). The TLD is a probability density function for likelihood of travel informed by travel time:
```{r}
tld <- od_ft_tt %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(trips = sum(trips),
            travel_time = mean(travel_time))
```

Plot the TLD:
```{r}
ggplot(data = tld, 
       aes(x =travel_time, y = trips)) +
  geom_point()
```
So, the artificially assigned '180'min travel time appears to be an outlier (visually) of the distribution trend but there are signficantly more 'unknown' destination trips with an undetermined travel time (assigned to "300min"). So since this is the case, we will not re-run the OD travel time calculations for trip which are longer than 180 min as the number of trips are relatively insignificant; i.e. ~35,000 trips which is only...

```{r}
null_dest_trips_numb <- od_ft_tt %>% filter(travel_time=="300") %>% summarize(trips = sum(trips))
totaltrip_trips_numb <- od_ft_tt %>% summarize(trips = sum(trips))
dest_180_trips_numb <- od_ft_tt %>% filter(travel_time=="180") %>% summarize(trips = sum(trips))

dest_180_trips_numb/(totaltrip_trips_numb-null_dest_trips_numb) * 100
```

0.95% of trips with known destinations! Compared to the number of unknown destination trips:
```{r}
null_dest_trips_numb/(totaltrip_trips_numb) * 100
```
which account for 10.1% of total trips.

Let's plot a guess: assigning a travel time for the trips with known destinations but uncalculated travel times; set a random travel time between 180min and 240min (3hr to 4hr!) for all travel times which have been artifically set to 180 min and plot.
```{r}
test <- od_ft_tt %>% mutate(travel_time = ifelse(travel_time==180, seq(from = 180, to = 240), travel_time))
```

Trip length distribution and plot:
```{r}
 test <- test %>%
  mutate(tt_classes = cut(travel_time, 
                          1000,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(trips = sum(trips),
            travel_time = mean(travel_time))

ggplot(data = test, 
       aes(x =travel_time, y = trips)) +
  geom_point()

```

It may be worth recalculating these longer travel times for completeness... but for now we move on. We will keep uncalculated travel times (greater than 180min) and thoses with unknown destinations NA.

```{r}
load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")

tld <- od_ft_tt %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(trips = sum(trips),
            travel_time = mean(travel_time))
```

## Saving TTS-16 objects to folder
```{r}
#car travel times for all origins to all destinations trips in TTS-16, number of associated trips, and the travel time.
usethis::use_data(od_ft_tt, overwrite = TRUE)

# all, unique, traffic analysis zones with associated number of workers and jobs per zone
load(file = "TTS16-data-inputs/OD-by-FT-Employment/ggh_taz.Rdata")
usethis::use_data(ggh_taz, overwrite = TRUE)
```
```{r}
#boundaries of interest
# load(file = "TTS16-data-inputs/Boundaries/hamilton_cma.Rdata")
# usethis::use_data(hamilton_cma, overwrite = TRUE)
# 
# load(file = "TTS16-data-inputs/Boundaries/toronto_muni_bound.Rdata")
# usethis::use_data(toronto_muni_bound, overwrite = TRUE)

load(file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
usethis::use_data(ggh_pd, overwrite = TRUE)
```

### 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
Final edits to manuscript and a gramatical fix to package contents.

@@
 In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS).
-Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, and the origin and destination trips.
-{TTS2016R} also includes the calculated car travel time from TAZ origin-destination centroid pairs.
-This information is useful to understand patterns of commuting in the region.
-We illustrate the estimation of a distance-decay curve using these data.
-{TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manupulation, and the final data products are detailed."
+Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, origin-destination trips, and calculated car travel time from TAZ origin-destination centroid pairs.
+To illustrate how this information can be analysed to understand patterns in commuting, we estimate a distance-decay curve (i.e., impedance function) for the region.
+{TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manipulation, and the final data products are detailed."
 
@@
 library(tmap)
+library(ggpmisc)
 # library(ggpmisc)
@@
 In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r].
-The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function, showcase how it can be used to calculate accessibility to employment, and invite its use in other applications.
+The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.
 
-Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
+<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
 In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
@@
 It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
+-->
 
@@
 
-{TTS2016R} includes counts of fully-employed population by place of residence and counts of full-time jobs by place of work aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries).
-Unique identifiers use the GTA06 Zoning System of TTS.
-The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) are organized in the form of an origin-destination table, indicative of patterns of home-to-work commute (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions).
-These data were retrieved from the Transportation Tomorrow Survey Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
+{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries).
+TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
+As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).
+
+The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions).
+These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
 
 Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}.
-These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible uses.
-It is important to note that for simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid.
+These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications.
+For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid.
 The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in the descriptive statistics in Table \ref{tab:TTS-16-desc-stats}.
+Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
 
@@
 ```
-```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE, out.height=400}
+```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
 #plot 
-ggplot() +
-  geom_sf(data = TTS2016R::ggh_pd,
-          fill = "gainsboro", color = "darkgray") +
+TTS_area_plot <- ggplot() +
+  geom_sf(data = TTS2016R::ggh_pd, color = "darkgray",
+          aes(fill = REGION_name)) +
+  scale_fill_manual(values = viridis::viridis(20)) +
   geom_sf(data = group_ggh_pd_poly, 
-          color = "steelblue", fill = NA, size = 0.7) +
-  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+          color = "black", fill = NA, size = 0.7) +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot
                          height = unit(0.8, "cm"), 
@@
                                                           fill = c("white","dimgrey"))) +
-  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot
                    height = unit(0.15, "cm")) + 
@@
                   aes(x = X, y = Y, label = REGION_name),
-                  size = 2.5,
+                  size = 3,
                   nudge_y = 3000,
                   nudge_x = 2000) +
-  theme_minimal() +
+  theme_void() +
   theme(legend.position = "none",
@@
 
-ggsave("images/TTS16-survey-area.png")
+# ggsave("images/TTS16-survey-area.png")
 ```
-```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the Greater Golden Horseshoe in Ontario, Canada.", out.width="80%", fig.align='center'}
-knitr::include_graphics("images/TTS16-survey-area.png")
-```
-
 ```{r creating-desc-stats-table}
@@
 
-Statistics <- data.frame("Statistics" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))
+Statistics <- data.frame("Statistic" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))
 
-Trips <- data.frame("OD_Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
+Trips <- data.frame("Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
                                    summary(od_ft_tt$trips)[[2]] %>% round(),  
@@
 
-Travel_time <- data.frame("OD_Travel_time" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
+Travel_time <- data.frame("TT" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
                                                summary(od_ft_tt$travel_time)[[2]] %>% round(),  
@@
 
-TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(), 
-                                      summary(ggh_taz$AREA)[[2]] %>% round(), 
-                                      summary(ggh_taz$AREA)[[3]] %>% round(), 
-                                      summary(ggh_taz$AREA)[[4]] %>% round(), 
-                                      summary(ggh_taz$AREA)[[5]] %>% round(), 
-                                      summary(ggh_taz$AREA)[[6]] %>% round(), 
-                                      NA))
+# TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(1), 
+#                                       summary(ggh_taz$AREA)[[2]] %>% round(1), 
+#                                       summary(ggh_taz$AREA)[[3]] %>% round(1), 
+#                                       summary(ggh_taz$AREA)[[4]] %>% round(1), 
+#                                       summary(ggh_taz$AREA)[[5]] %>% round(1), 
+#                                       summary(ggh_taz$AREA)[[6]] %>% round(1), 
+#                                       NA))
 
@@
 
-desc_stats <- cbind(Statistics, Trips, Travel_time, TAZ_Area, Workers, Jobs)
+desc_stats <- cbind(Statistics, Trips, Travel_time, Workers, Jobs)
 
-#kable tabling 
-desc_stats %>%
-  kable(format = "latex",
-        align="lrrrrrr",
-        booktabs = T,
-        col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
-        caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, workers, and jobs for the traffic analysis zones (TAZ) from the TTS 2016 dataset along with estimated car origin-destination travel times.") %>%
-  add_header_above(c(" ", "Trips", "Car Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
-  kable_styling(full_width = "T", 
-                latex_options = c("scale_down"),
-                position = "center")
+# #kable tabling 
+# desc_stats %>%
+#   kable(format = "latex",
+#         align="lrrrrrr",
+#         booktabs = T,
+#         col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
+#         caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, calculated origin-destination car travel time, TAZ area, workers per TAZ, and jobs per TAZ.") %>%
+#   add_header_above(c(" ", "Trips", "Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
+#   kable_styling(full_width = "T", 
+#                 latex_options = c("scale_down"),
+#                 position = "center")
+```
+
+```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
+Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align='center'}
+# knitr::include_graphics("images/TTS16-survey-area.png")
+
+TTS_area_plot + annotate(geom = "table",
+            x = 921500, y = 4890000, 
+           label = list(desc_stats))
 ```
@@
 
-The origin-destination information consists of a cross-tabulation of people who are employed full-time by place of residence (origin) and places of employment in the GGH (destination) using the GTA06 zoning system.
-The number of workers and jobs is not equal; the boundaries of the survey are permeable, so workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
+The origin-destination table (i.e., trips) consists of a cross-tabulation of people who are employed full-time by place of GGH residence (origin) and places of GGH employment (destination) using the GTA06 TAZ zoning system.
+It is important to note that the total number of workers and jobs is the TTS 2016 region are not equal but the number of trips taken are equal to the number of workers.
+Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
 This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs).
-While the 2016 TTS survey boundaries are drawn to minimize the difference between opportunities at the destination and supply at the origins they are still not equal <!-- cite?
-is this true ?
--->.
-That said, these data offer a perspective of the potential of home-based trips to places of employment and as such, the number of trips taken are equal to the number of workers in the GGH.
+As such, the origin-destination table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
 
-```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (top) and jobs (bottom) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=11}
+```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=5}
 
 tts_workers <- tm_shape(ggh_taz) +
-  tm_fill(col = "workers", title = "Full-time employed people", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "workers", title = "Full-time \nemployed people", palette = "-Spectral", style = "jenks") +
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 tts_jobs <- tm_shape(ggh_taz) +
-  tm_fill(col = "jobs", title = "Full-time jobs", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "jobs", title = "Full-time \njobs", palette = "-Spectral", style = "jenks") +
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 
+<!--
 ```{r ECD-plot, fig.cap="\\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set.
-Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ.
-"}
+Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ.", fig.width=8, fig.height=4}
 
@@
 ```
+-->
+Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ.
+It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs-housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
+It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
+<!-- see the following cumulative probability distribution in Figure \ref{fig:ECD-plot}.
+-->
 
-Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ.
-It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of jobs-housing imbalance, which can impact accessibility in a region [@Levine1998rethinking].
-Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH.
+<!-- Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH.
 The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022].
@@
 Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs.
+-->
 
-It can also be seen that from the bottom plot in Figure \ref{fig:tts-workers-jobs-plot} that high to medium-low concentrations of jobs are often present in the same areas as workers but only when the scale is transformed.
-In other words, though there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers; see the following cumulative probability distribution in Figure \ref{fig:ECD-plot} in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ.
+<!--- in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ.
 This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}.
+-->
 
@@
 
-{TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R package {r5r} and is interpreted as the travel time for a work commute for full-time employed people in the GGH.
-The travel times were calculated assuming the following input parameters: a maximum travel time less than or equal to 180 mins (3 hrs) and a street network retrieved from OpenStreetMaps for the GGH area.
-The 3 hr threshold was selected as it captures 99% of the trip taken (see the travel times summarized in Table \ref{tab:TTS-16-desc-stats}).
-
-It is important to note that all travel times within this data set are calculated assuming car travel and one departure time for all origins.
-These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies, however, it is not possible from the data retrieval system to obtain higher order tabulations that would allow us to split the data further.
-Thus we do not know which trips are made with non-car modes nor exact departure times in these tables.
-Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity, we assume that all trips are taken by one-time departure car trip.
+As mentioned, {TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
+This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Table \ref{tab:TTS-16-desc-stats}).
+It is important to note that travel times within this data set are calculated assuming only car travel and one departure time for all origins.
+These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies.
+However, it is not possible from the data retrieval system to obtain higher order tabulations so <!--  we do not know which trips are made with non-car modes nor exact departure times in these tables.
+Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity,-->  we carry on with the assume that all trips are taken by one-time departure car trip.
 
@@
 
-```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (top) and average worker travel time (bottom) for each TAZ in the GGH.", fig.width=8, fig.height=11, message=FALSE}
+```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (left) and average worker travel time (right) for each TAZ in the GGH.", fig.width=8, fig.height=5, message=FALSE}
 
@@
 
-As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time (min) resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}).
-However, when the average travel time per trip in each TAZ is presented, the spatial distribution is distinct from all other plots presented so far.
-We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA) (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) and Niagara and Waterloo, the average travel times are moderately low.
+As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the average travel time is distinct from other plots presented so far.
+We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA)<!-- (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) -->, Niagara and Waterloo, the average travel times are moderately low.
 Further from these areas, travel times are higher.
@@
 
-Impedance functions are useful to understand mobility behavior and are part, for instance, of gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and accessibility analysis in many applications [@hansen_how_1959;@levinson_accessibility_1998;@talen_assessing_1998;@reggiani_accessibility_2011;@paez_jobs_2013;@kuai_examining_2017; @barboza_balancing_2021].
-An origin destination matrix and a cost matrix (i.e., with travel times) can be used in combination to estimate impedance functions.
-An impedance function $f(\cdot)$ depends on the cost of travel between locations $i$ and $j$ $c_{ij}$; it usually is a monotonically decreasing function, although sometimes the function can increase to reflect patterns of separation between activities; for instance, separation of land uses means that very short commuting trips are relatively rare.
-There can be sometimes also fluctuations due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration.
+Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@paez_jobs_2013;@barboza_balancing_2021].
+An impedance function $f(\cdot)$ depends on the cost of travel $c_{ij}$ between locations $i$ and $j$ (all which is supplied in the travel time and origin-destination table within {TTS2016R}).
+<!-- The impedance function is usually a monotonically decreasing function since the separation of land uses means that very short commuting trips are relatively rare.
+However, sometimes the function can increase to reflect patterns of separation between activities.
+This could be needed to reflect fluctuations in travel cost due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration.
+-->
 
@@
 
-In the GGH data presented, the empirical TLD (i.e., proportion of trips taken vs.
-travel time in minutes) is fitted to a density distribution using maximum likelihood techniques and the Nelder-Mead method for direct optimization available within the `fitdistrplus` package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics seen in Figure \ref{fig:TLD-Gamma-plot} and Figure \ref{fig:plot-cullen-frey}, the gamma distribution is selected for the presented data.
+Using the data contained in {TTS2016R}, we fit the empirical TLD to a density distribution using maximum likelihood techniques and the Nelder-Mead method for direct optimization available within the `R` package {fitdistrplus} [@fitdistrplus_2015].
+Based on goodness-of-fit criteria and diagnostics seen in Figure \ref{fig:TLD-Gamma-plot}, the gamma distribution is selected.
+The 'shape' parameter is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).
 
@@
 
-```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots"}
+```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots", out.width="75%", fig.align='center'}
 knitr::include_graphics("images/impedance_function.png")
 ```
-
-The resulting calibrated impedance function is given in the following general form where the estimated 'shape' is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).
 
@@
 
-```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."}
-fitdistrplus::descdist(data=all_tt) 
-```
+<!-- # ```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."} -->
+<!-- # fitdistrplus::descdist(data=all_tt)  -->
+<!-- # ``` -->
 
-# Accessibility to employment
-
-As noted above, impedance functions are an essential component of accessibility analysis.
-Equation (\ref{eq:conventional-accessibility}) shows that accessibility $A_i$ is the weighted sum of opportunities that can be reached from location $i$, given the cost of travel $c_{ij}$.
-Summing the opportunities in the neighborhood of $i$ as defined by the impedance function $f(\cdot)$, provides estimates of the number of opportunities that can be reached from $i$ at a certain cost.
-
-\begin{equation}
-\label{eq:conventional-accessibility}
-A_i = \sum_{j=1}^JO_jf(c_{ij})
-\end{equation}
-
-\noindent where:
-
--   $A$ is accessibility. 
--   $i$ is a set of origin locations.
--   $j$ is a set of destination locations.
--   $O_j$ is the number of opportunities at location $j$. These are opportunities for activity and add some sort of *supply* to the area;
--   $c_{ij}$ is a measure of the cost of moving between $i$ and $j$
--   $f(\cdot)$ is an impedance function of $c_{ij}$.
-
-```{r access-per-worker}
-GGH_taz_acc <- GGH_taz_acc %>%
-  mutate(avg_A_per_W = GGH_A_i/workers)
-```
-
-```{r plot-access-SA-GGH-TTS, fig.cap="\\label{fig:plot-access-SA-GGH-TTS}Accessibility (top) and normalized accessibility (bottom) to employment in the GGH area", fig.width=8, fig.height=11, message=FALSE}
-## accessibility
-access_TTS_GGH <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "GGH_A_i", title = "Accessibility \n(A_i)", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-access_normalized_TTS_GGH <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "avg_A_per_W", title = "Average accessibility \nper worker", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-tmap_arrange(access_TTS_GGH, access_normalized_TTS_GGH)
-```
 \newpage
-
-Accessibility is a property of the origin (i.e., origin TAZ in our data), the landscape of opportunities (i.e., the TAZ destinations in our data), and transportation infrastructure (i.e., the travel time table).
-Figure \ref{fig:plot-access-SA-GGH-TTS} presents the accessibility estimates that result from the impedance function calibrated in the preceding section.
-The top plot, that of the raw accessibility values, presents a distinct radial trend where the majority of TAZ in and around Toronto have high accessibility values and values gradually decrease in TAZ which are further from Toronto's boundary.
-Alternatively, the bottom plot in Figure \ref{fig:plot-access-SA-GGH-TTS} present the worker-normalized accessibility values.
-In other words, the accessibility value for each TAZ is divided by the number of workers in each TAZ.
-The patterns in this plot are similar to the Greater Toronto Area radial trend observed in the top plot but are less extreme (i.e., smaller spread).
-
-<!-- room to critique accessibility or suggestion how we need alternatives to comprehend this measure..? -->
 
@@
 
-The open data product introduced in this paper shares tables for worker-to-employment data from the 2016 TTS aggregated by TAZ.
-In addition, inter-centroid travel time tables are calculated, and the planning/municipality boundaries are included to compliment the 2016 TTS data.
+The open data product introduced in this paper shares tables for home-to-work related data from the 2016 TTS.
+In addition, inter-centroid travel time tables are calculated, and the planning/municipality boundaries are included as a compliment.
 This open data product, {TTS2016R}, is freely available to explore in an `R` environment.
-One possible use of these data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.
+One possible use of this data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.
 
-New digital formats are increasingly complex and the explanation of the methods often do not concisely and intelligibly fit within the confines of a traditional article.
+<!-- New digital formats are increasingly complex and the explanation of the methods often do not concisely and intelligibly fit within the confines of a traditional article.
 With this motivation, we invite all who are interested to use {TTS2016R} to explore the worker-employment patterns contained in the {TTS2016R} package.
-In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own.
+--> In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own.
 We see this product as providing open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
 
-\newpage
-
 # References {#references .unnumbered}

### 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.tex
Final edits to manuscript and a gramatical fix to package contents.

@@
 the number of people who are employed full-time per TAZ, the number of
-jobs per TAZ, and the origin and destination trips. \{TTS2016R\} also
-includes the calculated car travel time from TAZ origin-destination
-centroid pairs. This information is useful to understand patterns of
-commuting in the region. We illustrate the estimation of a
-distance-decay curve using these data. \{TTS2016R\} can be freely
-downloaded and explored at: \url{https://github.com/soukhova/TTS2016R}
-where the documentation and code involved in data creation,
-manupulation, and the final data products are detailed.
+jobs per TAZ, origin-destination trips, and calculated car travel time
+from TAZ origin-destination centroid pairs. To illustrate how this
+information can be analysed to understand patterns in commuting, we
+estimate a distance-decay curve (i.e., impedance function) for the
+region. \{TTS2016R\} can be freely downloaded and explored at:
+\url{https://github.com/soukhova/TTS2016R} where the documentation and
+code involved in data creation, manipulation, and the final data
+products are detailed.
 \end{abstract}
@@
 readers through the empirical home-based work commute data set,
-illustrate the calculation of an impedance function, showcase how it can
-be used to calculate accessibility to employment, and invite its use in
-other applications.
-
-Data from the TTS are in principle available to the public but are not
-fully open, since permission to access the data retrieval system is
-required. In addition, the raw data can be technically demanding,
-cumbersome to work with, and requires multiple software to process. By
-pre-processing the data in a \texttt{R} environment, \{TTS2016R\} offers
-a slice of the TTS data useful to understand patterns of commuting to
-work in the region. It also provides open infrastructure for additional
-TTS or complimentary data sets to be amended by the authors or wider
-open-source community in the future.
+illustrate the calculation of an impedance function that can be used to
+calculate accessibility to employment, and invite its use in other
+applications.
 
@@
 \{TTS2016R\} includes counts of fully-employed population by place of
-residence and counts of full-time jobs by place of work aggregated by
-TAZ (n=3,764 within the survey boundaries). Unique identifiers use the
-GTA06 Zoning System of TTS. The number of jobs (3,081,900) and workers
-(3,446,957) are organized in the form of an origin-destination table,
-indicative of patterns of home-to-work commute (there are 3,446,957
-potential interactions). These data were retrieved from the
-Transportation Tomorrow Survey Data Retrieval System on October 28, 2021
-and reflect the potential interaction of full-time employed people and
-jobs within the GGH survey boundaries shown in Figure
-\ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology
-\citep{data_management_group_tts_2018}.
+residence (origin) and counts of full-time jobs by place of work
+(destination) aggregated by TAZ (n=3,764 within the survey boundaries).
+TAZ typically are defined based on land-use and population demographics
+in order to estimate the number of trips produced and attracted to each
+zone \citep{meyer_urban_2001}. As such, each TAZ is uniquely identified
+using the GTA06 Zoning System which can be used to join to the
+origin-destination table (i.e., trips taken).
+
+The number of jobs (3,081,885) and workers (3,446,957) in this package
+are organized in the form of an origin-destination table which is
+indicative of home-to-work commute patterns (there are 3,446,957
+potential interactions). These data were retrieved from the TTS Data
+Retrieval System on October 28, 2021 and reflect the potential
+interaction of full-time employed people and jobs within the GGH survey
+boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by
+the 2016 TTS methodology \citep{data_management_group_tts_2018}.
 
@@
 the cost of travel and to calculate impedance functions, among other
-possible uses. It is important to note that for simplicity, all
-interactions within \{TTS2016R\} are assumed to be taken by car, and the
-travel time is calculated from an origin TAZ centroid to a destination
-TAZ centroid. The centroid is snapped to the nearest street line by
-\texttt{r5r} and the travel time is calculated for all trips assuming a
-car travel mode. Additionally, only travel times less than or equal to
-180 mins (3 hrs) are calculated; this threshold represents 99\% of
-trip's travel times which are summarized in the descriptive statistics
-in Table \ref{tab:TTS-16-desc-stats}.
+possible applications. For simplicity, all interactions within
+\{TTS2016R\} are assumed to be taken by car, and the travel time is
+calculated from an origin TAZ centroid to a destination TAZ centroid.
+The centroid is snapped to the nearest street line by \texttt{r5r} and
+the travel time is calculated for all trips assuming a car travel mode.
+Additionally, only travel times less than or equal to 180 mins (3 hrs)
+are calculated; this threshold represents 99\% of trip's travel times
+which are summarized in Figure \ref{fig:TTS-16-survey-area}.
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{images/TTS16-survey-area}
+{\centering \includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/TTS-16-survey-area-1}
 
@@
 
-\caption{\label{fig:TTS-16-survey-area}The TTS 2016 study area within the Greater Golden Horseshoe in Ontario, Canada.}\label{fig:TTS-16-survey-area}
+\caption{\label{fig:TTS-16-survey-area}The TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
+Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
 \end{figure}
-
-\begin{table}
-
-\caption{\label{tab:creating-desc-stats-table}\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, workers, and jobs for the traffic analysis zones (TAZ) from the TTS 2016 dataset along with estimated car origin-destination travel times.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
-\toprule
-\multicolumn{1}{r}{ } & \multicolumn{1}{r}{Trips} & \multicolumn{1}{r}{Car Travel Time} & \multicolumn{1}{r}{Area} & \multicolumn{1}{r}{Workers} & \multicolumn{1}{r}{Jobs} \\
-\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
-  & (\#) & (min) & (km\textasciicircum{}2) & (\#) & (\#)\\
-\midrule
-Min. & 1 & 0 & 0 & 0 & 0\\
-1st Qu. & 14 & 13 & 1 & 25 & 64\\
-Median & 22 & 20 & 1 & 464 & 244\\
-Mean & 33 & 23 & 7 & 916 & 819\\
-3rd Qu. & 38 & 30 & 3 & 1378 & 700\\
-\addlinespace
-Max. & 1129 & 179 & 879 & 8491 & 41821\\
-NA's & NA & 3507 & NA & NA & NA\\
-\bottomrule
-\end{tabu}}
-\end{table}
 
@@
 
-The origin-destination information consists of a cross-tabulation of
-people who are employed full-time by place of residence (origin) and
-places of employment in the GGH (destination) using the GTA06 zoning
-system. The number of workers and jobs is not equal; the boundaries of
-the survey are permeable, so workers who reside within the boundaries
-but travel outside of the boundaries are counted as workers within an
-origin TAZ, while jobs in TAZ that are filled by workers who reside
-outside the GGH boundaries are \emph{unknown} since they were not
+The origin-destination table (i.e., trips) consists of a
+cross-tabulation of people who are employed full-time by place of GGH
+residence (origin) and places of GGH employment (destination) using the
+GTA06 TAZ zoning system. It is important to note that the total number
+of workers and jobs is the TTS 2016 region are not equal but the number
+of trips taken are equal to the number of workers. Since the outer
+boundaries of the TTS are permeable, workers who reside within the
+boundaries but travel outside of the boundaries are counted as workers
+within an origin TAZ, while jobs in TAZ that are filled by workers who
+reside outside the GGH boundaries are \emph{unknown} since they were not
 surveyed. This mismatch results in the total number of workers being
 1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
-3,081,900 jobs). While the 2016 TTS survey boundaries are drawn to
-minimize the difference between opportunities at the destination and
-supply at the origins they are still not equal . That said, these data
-offer a perspective of the potential of home-based trips to places of
-employment and as such, the number of trips taken are equal to the
-number of workers in the GGH.
+3,081,885 jobs). As such, the origin-destination table contained in
+\{TTS2016R\} offers a perspective on all workers in the GGH and their
+home-based trips to places of GGH employment.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (top) and jobs (bottom) in each TAZ in the GGH area as specified in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/ECD-plot-1} \caption{\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set.
-Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ.
-}\label{fig:ECD-plot}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the GGH area as specified in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
 \end{figure}
@@
 and jobs per TAZ. It can be observed that the spatial distribution of
-jobs and workers is unequal, which is indicative of jobs-housing
-imbalance, which can impact accessibility in a region
-\citep{Levine1998rethinking}. Workers are concentrated in TAZ within the
-center of the GGH and along the south-east and northern boarder of the
-GGH. The center of the GGH corresponds to the Greater Toronto Area (GTA)
-which is the most densely populated area in southern Ontario
-\citep{statistics_canada_daily_2022}. The south east border of the GGH
-neighbours Lake Ontario and is delineated by the urban built boundary of
-the Ontario Growth Plan being home to the highest density of working
-population in the GGH
-\citep{ontario_built_2019, auditor_general_of_ontario_value_2021}. The
-northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake,
-and Peterborough regions which are home to lower density of worker
-population density population
-\citep{auditor_general_of_ontario_value_2021}. Conversely, the spread of
-jobs in the GGH is lower than the number of workers indicating
-population is more spatially dispersed than jobs.
-
-It can also be seen that from the bottom plot in Figure
-\ref{fig:tts-workers-jobs-plot} that high to medium-low concentrations
-of jobs are often present in the same areas as workers but only when the
-scale is transformed. In other words, though there is a higher number of
-TAZ with no workers than zones with no jobs (i.e., 791 TAZ with no
-workers : 396 TAZ with no jobs) and the mean of workers per TAZ is
-higher than the mean of jobs (i.e., 916 workers : 819 jobs) the number
-TAZ with an extreme number of jobs at the highest and lowest percentiles
-is significantly higher than the number of workers; see the following
-cumulative probability distribution in Figure \ref{fig:ECD-plot} in
-which the 94th to 100th percentile and the 0th to 34th percentile of
-jobs in TAZ is higher than the number of workers in TAZ. This means that
-between these ranges, TAZ have a higher number of workers than they do
-jobs, echoing the more even spatial distribution of workers observed in
-Figure \ref{fig:tts-workers-jobs-plot}.
+jobs and workers is unequal, which is indicative of a jobs-housing
+imbalance that can impact accessibility in a region
+\citep{Levine1998rethinking}. It can also be seen that there is a higher
+number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
+with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
+is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
+number TAZ with an extreme number of jobs at the highest and lowest
+percentiles is significantly higher than the number of workers.
 
@@
 
-\{TTS2016R\} also includes travel time data for each home-to-work trip
-as displayed in Figure \ref{fig:plot-tt-ttpertrip}. This travel time
-corresponds to a car commute calculated using the R package \{r5r\} and
-is interpreted as the travel time for a work commute for full-time
-employed people in the GGH. The travel times were calculated assuming
-the following input parameters: a maximum travel time less than or equal
-to 180 mins (3 hrs) and a street network retrieved from OpenStreetMaps
-for the GGH area. The 3 hr threshold was selected as it captures 99\% of
-the trip taken (see the travel times summarized in Table
-\ref{tab:TTS-16-desc-stats}).
-
-It is important to note that all travel times within this data set are
-calculated assuming car travel and one departure time for all origins.
-These assumptions are not completely realistic since we know a small
-proportion of trips are taken by non-car modes and travel time
-departures varies, however, it is not possible from the data retrieval
-system to obtain higher order tabulations that would allow us to split
-the data further. Thus we do not know which trips are made with non-car
-modes nor exact departure times in these tables. Though modal split and
-travel times can be estimated through other methods
-\citep[e.g.,][]{allen_suburbanization_2021, higgins2021changes}, for
-simplicity, we assume that all trips are taken by one-time departure car
-trip.
+As mentioned, \{TTS2016R\} also includes travel time data for each
+home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
+This travel time corresponds to a car commute calculated using the R
+package \{r5r\} (see descriptive statistics in Table
+\ref{tab:TTS-16-desc-stats}). It is important to note that travel times
+within this data set are calculated assuming only car travel and one
+departure time for all origins. These assumptions are not completely
+realistic since we know a small proportion of trips are taken by non-car
+modes and travel time departures varies. However, it is not possible
+from the data retrieval system to obtain higher order tabulations so we
+carry on with the assume that all trips are taken by one-time departure
+car trip.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (top) and average worker travel time (bottom) for each TAZ in the GGH.}\label{fig:plot-tt-ttpertrip}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (left) and average worker travel time (right) for each TAZ in the GGH.}\label{fig:plot-tt-ttpertrip}
 \end{figure}
@@
 As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
-travel time (min) resembles the spatial trend distribution in the number
-of employed people in the previous plot (Figure
-\ref{fig:tts-workers-jobs-plot}). However, when the average travel time
-per trip in each TAZ is presented, the spatial distribution is distinct
-from all other plots presented so far. We can see that in areas around
-the south-eastern border that make up the Greater Toronto and Hamilton
-Area (GTHA) (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) and
-Niagara and Waterloo, the average travel times are moderately low.
-Further from these areas, travel times are higher. Interestingly, even
-in eastern areas (e.g., Peterborough) with high employment and high job
-concentration, average travel time is higher than within the GTHA.
+travel time resembles the spatial trend distribution in the number of
+employed people in the previous plot (Figure
+\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
+average travel time is distinct from other plots presented so far. We
+can see that in areas around the south-eastern border that make up the
+Greater Toronto and Hamilton Area (GTHA), Niagara and Waterloo, the
+average travel times are moderately low. Further from these areas,
+travel times are higher. Interestingly, even in eastern areas (e.g.,
+Peterborough) with high employment and high job concentration, average
+travel time is higher than within the GTHA.
 
@@
 
-Impedance functions are useful to understand mobility behavior and are
-part, for instance, of gravity models of spatial interaction
-\citep{wilson1971, haynes_gravity_1985} and accessibility analysis in
-many applications
-\citep{hansen_how_1959, levinson_accessibility_1998, talen_assessing_1998, reggiani_accessibility_2011, paez_jobs_2013, kuai_examining_2017, barboza_balancing_2021}.
-An origin destination matrix and a cost matrix (i.e., with travel times)
-can be used in combination to estimate impedance functions. An impedance
-function \(f(\cdot)\) depends on the cost of travel between locations
-\(i\) and \(j\) \(c_{ij}\); it usually is a monotonically decreasing
-function, although sometimes the function can increase to reflect
-patterns of separation between activities; for instance, separation of
-land uses means that very short commuting trips are relatively rare.
-There can be sometimes also fluctuations due to hierarchical patterns,
-where travelers bypass opportunities in favor of more distant
-destinations that offer economies of agglomeration.
+Impedance functions are useful to understand mobility behaviour and are
+used to estimate gravity models of spatial interaction
+\citep{wilson1971, haynes_gravity_1985} and applied in accessibility
+analysis
+\citep{hansen_how_1959, paez_jobs_2013, barboza_balancing_2021}. An
+impedance function \(f(\cdot)\) depends on the cost of travel \(c_{ij}\)
+between locations \(i\) and \(j\) (all which is supplied in the travel
+time and origin-destination table within \{TTS2016R\}).
 
@@
 
-In the GGH data presented, the empirical TLD (i.e., proportion of trips
-taken vs.~travel time in minutes) is fitted to a density distribution
-using maximum likelihood techniques and the Nelder-Mead method for
-direct optimization available within the \texttt{fitdistrplus} package
-\citep{fitdistrplus_2015}. Based on goodness-of-fit criteria and
-diagnostics seen in Figure \ref{fig:TLD-Gamma-plot} and Figure
-\ref{fig:plot-cullen-frey}, the gamma distribution is selected for the
-presented data.
+Using the data contained in \{TTS2016R\}, we fit the empirical TLD to a
+density distribution using maximum likelihood techniques and the
+Nelder-Mead method for direct optimization available within the
+\texttt{R} package \{fitdistrplus\} \citep{fitdistrplus_2015}. Based on
+goodness-of-fit criteria and diagnostics seen in Figure
+\ref{fig:TLD-Gamma-plot}, the gamma distribution is selected. The
+`shape' parameter is \(\alpha\) = 2.019, the estimated `rate' is
+\(\beta\) = 0.094 , and \(\Gamma(\alpha)\) is defined in Equation
+(\ref{gamma-dist}).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{images/impedance_function} \caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
+
+{\centering \includegraphics[width=0.75\linewidth]{images/impedance_function}
+
+}
+
+\caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
 \end{figure}
-
-The resulting calibrated impedance function is given in the following
-general form where the estimated `shape' is \(\alpha\) = 2.019, the
-estimated `rate' is \(\beta\) = 0.094 , and \(\Gamma(\alpha)\) is
-defined in Equation (\ref{gamma-dist}).
 
@@
 
-\begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-cullen-frey-1} \caption{\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times.}\label{fig:plot-cullen-frey}
-\end{figure}
-
-\begin{verbatim}
-summary statistics
-------
-min:  0.1   max:  179 
-median:  18 
-mean:  21.4344 
-estimated sd:  14.61254 
-estimated skewness:  1.703326 
-estimated kurtosis:  8.353363 
-\end{verbatim}
-
-\hypertarget{accessibility-to-employment}{%
-\section{Accessibility to
-employment}\label{accessibility-to-employment}}
-
-As noted above, impedance functions are an essential component of
-accessibility analysis. Equation (\ref{eq:conventional-accessibility})
-shows that accessibility \(A_i\) is the weighted sum of opportunities
-that can be reached from location \(i\), given the cost of travel
-\(c_{ij}\). Summing the opportunities in the neighborhood of \(i\) as
-defined by the impedance function \(f(\cdot)\), provides estimates of
-the number of opportunities that can be reached from \(i\) at a certain
-cost.
-
-\begin{equation}
-\label{eq:conventional-accessibility}
-A_i = \sum_{j=1}^JO_jf(c_{ij})
-\end{equation}
-
-\noindent where:
-
-\begin{itemize}
-\tightlist
-\item
-  \(A\) is accessibility.
-\item
-  \(i\) is a set of origin locations.
-\item
-  \(j\) is a set of destination locations.
-\item
-  \(O_j\) is the number of opportunities at location \(j\). These are
-  opportunities for activity and add some sort of \emph{supply} to the
-  area;
-\item
-  \(c_{ij}\) is a measure of the cost of moving between \(i\) and \(j\)
-\item
-  \(f(\cdot)\) is an impedance function of \(c_{ij}\).
-\end{itemize}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-access-SA-GGH-TTS-1} \caption{\label{fig:plot-access-SA-GGH-TTS}Accessibility (top) and normalized accessibility (bottom) to employment in the GGH area}\label{fig:plot-access-SA-GGH-TTS}
-\end{figure}
 \newpage
-
-Accessibility is a property of the origin (i.e., origin TAZ in our
-data), the landscape of opportunities (i.e., the TAZ destinations in our
-data), and transportation infrastructure (i.e., the travel time table).
-Figure \ref{fig:plot-access-SA-GGH-TTS} presents the accessibility
-estimates that result from the impedance function calibrated in the
-preceding section. The top plot, that of the raw accessibility values,
-presents a distinct radial trend where the majority of TAZ in and around
-Toronto have high accessibility values and values gradually decrease in
-TAZ which are further from Toronto's boundary. Alternatively, the bottom
-plot in Figure \ref{fig:plot-access-SA-GGH-TTS} present the
-worker-normalized accessibility values. In other words, the
-accessibility value for each TAZ is divided by the number of workers in
-each TAZ. The patterns in this plot are similar to the Greater Toronto
-Area radial trend observed in the top plot but are less extreme (i.e.,
-smaller spread).
 
@@
 The open data product introduced in this paper shares tables for
-worker-to-employment data from the 2016 TTS aggregated by TAZ. In
-addition, inter-centroid travel time tables are calculated, and the
-planning/municipality boundaries are included to compliment the 2016 TTS
-data. This open data product, \{TTS2016R\}, is freely available to
-explore in an \texttt{R} environment. One possible use of these data, as
-showcased in this paper, is the calibration of impedance functions which
-in turn can be used for accessibility analysis.
+home-to-work related data from the 2016 TTS. In addition, inter-centroid
+travel time tables are calculated, and the planning/municipality
+boundaries are included as a compliment. This open data product,
+\{TTS2016R\}, is freely available to explore in an \texttt{R}
+environment. One possible use of this data, as showcased in this paper,
+is the calibration of impedance functions which in turn can be used for
+accessibility analysis.
 
-New digital formats are increasingly complex and the explanation of the
-methods often do not concisely and intelligibly fit within the confines
-of a traditional article. With this motivation, we invite all who are
-interested to use \{TTS2016R\} to explore the worker-employment patterns
-contained in the \{TTS2016R\} package. In the spirit of novel and
-original research, we hope readers value the efforts made to detail the
-data in order to improve transparency in our work and encourage others
-to replicate and, hopefully, inspire research of their own. We see this
-product as providing open infrastructure for additional TTS or
-complimentary data sets to be amended by the authors or wider
-open-source community in the future.
-
-\newpage
+In the spirit of novel and original research, we hope readers value the
+efforts made to detail the data in order to improve transparency in our
+work and encourage others to replicate and, hopefully, inspire research
+of their own. We see this product as providing open infrastructure for
+additional TTS or complimentary data sets to be amended by the authors
+or wider open-source community in the future.
 

### 2022-03-11 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
Final edits to manuscript and a gramatical fix to package contents.

@@
                                                                                  ifelse(REGION == 15, "Orangeville", 
-                                                                                        ifelse(REGION == 16, "Barrier", 
+                                                                                        ifelse(REGION == 16, "Barrie", 
                                                                                                ifelse(REGION == 17, "Simcoe", 
@@
 #workers (origins)
-workers_ft_cent <- workers_ft %>% merge(ggh_taz_cent %>% select(GTA06), by="GTA06") %>% st_sf
+workers_ft_cent <- workers_ft %>% merge(ggh_taz_cent %>% dplyr::select(GTA06), by="GTA06") %>% st_sf()
 
 work_origins <- cbind(workers_ft_cent, st_coordinates(st_transform(workers_ft_cent, crs = 4326))) %>%
-  rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
-  select(id, lon, lat) %>% as.data.frame()
+  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
+  dplyr::select(id, lon, lat) %>% as.data.frame()
 
@@
 #jobs (destinations)
-jobs_ft_cent <- jobs_ft %>% merge(ggh_taz_cent %>% select(GTA06), by="GTA06") %>% st_sf
+jobs_ft_cent <- jobs_ft %>% merge(ggh_taz_cent %>% dplyr::select(GTA06), by="GTA06") %>% st_sf
 
 job_destinations <- cbind(jobs_ft_cent, st_coordinates(st_transform(jobs_ft_cent, crs = 4326))) %>%
-  rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
-  select(id, lon, lat)
+  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
+  dplyr::select(id, lon, lat)
 

### 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
small edits to manuscript

@@
 Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
-
-```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
+```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
 Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align='center'}
@@
 TTS_area_plot + annotate(geom = "table",
-            x = 921500, y = 4890000, 
+            x = 915000, y = 4880000, 
            label = list(desc_stats))
 ```
-
 ## Employed individuals and jobs
@@
 
-```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=5}
+```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.", fig.width=8, fig.height=5}
 
@@
 
-\newpage
-
 ## Calculated travel time
@@
 
-```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (left) and average worker travel time (right) for each TAZ in the GGH.", fig.width=8, fig.height=5, message=FALSE}
+```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.", fig.width=8, fig.height=5, message=FALSE}
 
@@
 
-Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@paez_jobs_2013;@barboza_balancing_2021].
+Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@talen_assessing_1998;@paez_jobs_2013;@barboza_balancing_2021].
 An impedance function $f(\cdot)$ depends on the cost of travel $c_{ij}$ between locations $i$ and $j$ (all which is supplied in the travel time and origin-destination table within {TTS2016R}).
@@
 
-```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots", out.width="75%", fig.align='center'}
+```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car TLD (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots", out.width="75%", fig.align='center'}
 knitr::include_graphics("images/impedance_function.png")

### 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.tex
small edits to manuscript

@@
 
-\caption{\label{fig:TTS-16-survey-area}The TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
+\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
 Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the GGH area as specified in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
 \end{figure}
@@
 percentiles is significantly higher than the number of workers.
-
-\newpage
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (left) and average worker travel time (right) for each TAZ in the GGH.}\label{fig:plot-tt-ttpertrip}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.}\label{fig:plot-tt-ttpertrip}
 \end{figure}
@@
 analysis
-\citep{hansen_how_1959, paez_jobs_2013, barboza_balancing_2021}. An
-impedance function \(f(\cdot)\) depends on the cost of travel \(c_{ij}\)
-between locations \(i\) and \(j\) (all which is supplied in the travel
-time and origin-destination table within \{TTS2016R\}).
+\citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
+An impedance function \(f(\cdot)\) depends on the cost of travel
+\(c_{ij}\) between locations \(i\) and \(j\) (all which is supplied in
+the travel time and origin-destination table within \{TTS2016R\}).
 
@@
 
-\caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car trip length distribution (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
+\caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car TLD (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
 \end{figure}

### 2022-03-17 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.Rmd
additional changes to manuscript - final

--- 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
@@
 ---
-title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area in Canada"
+title: "TTS2016R: A dataset to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area, Canada"
 author:
@@
     affiliation: SEES
-  - name: "Antonio Paez"
+  - name: "Antonio Páez"
     email: "paezha@mcmaster.ca"
@@
   - times
-bibliography: bibfile
+bibliography: bibfile.bib
 bibliographystyle: sageh
@@
 ---
-```{r knitr-setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  cache = TRUE,
-  warning = FALSE,
-  comment = '', 
-  out.width = "1\\linewidth")
-```
-
-```{r install-data-package, include=FALSE}
-if (!require("TTS2016R", character.only = TRUE)) {
-      remotes::install_github("soukhova/TTS2016R",
-                        build_vignettes = TRUE)
-  }
-```
-
-```{r load-packages, include=FALSE, cache=FALSE}
-library(TTS2016R)
-library(dplyr)
-library(fitdistrplus)
-library(ggplot2)
-library(kableExtra)
-library(patchwork)
-library(sf)
-library(scales)
-library(stats)
-library(ggspatial)
-library(shadowtext)
-library(tmap)
-library(ggpmisc)
-# library(ggpmisc)
-# library(ggrepel)
-# library(cowplot)
-# library(spdep)
-# library(RColorBrewer)
-# library(extrafont)
-# font_import()
-# loadfonts(device = "win")
-
-options(scipen = 999)
-```
-
-# Introduction
-
-This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R).
-Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open].
-The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data.
-This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018].
-In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r].
-The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.
-
-<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
-In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
-By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region.
-It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
--->
-
-# Home-to-work commute data
-
-{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries).
-TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
-As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).
-
-The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions).
-These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
-
-Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}.
-These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications.
-For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid.
-The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# grouping the planning boundaries/municipalities so they make up the 20 regions in the TTS 2016.
-Note: st_buffer is used as there are small existing gaps between some boundaries.
-st_buffer of 10 m is enough to widen all boundaries and complete the st_union without issue.
-group_ggh_pd_poly <- TTS2016R::ggh_pd %>% st_buffer(10) %>% group_by(REGION) %>% 
-  summarize(REGION_name = first(REGION_name),
-            geometry = st_union((geometry)))
-
-# creating an object of centroids for each region - this will be used to label polygons on the map
-group_ggh_pd <- sf::st_centroid(group_ggh_pd_poly) 
-points <- sf::st_coordinates(group_ggh_pd) %>% data.frame() 
-group_ggh_pd <- cbind(group_ggh_pd, points)
-
-## manually readjusting the X and Y coordinate of "County of Peterborough" and "Brant" as they overlap some cities
-group_ggh_pd[group_ggh_pd$REGION_name=="Brant", "X"] <- 544000.0
-group_ggh_pd[group_ggh_pd$REGION_name=="Brant", "Y"] <- 4767466
-
-group_ggh_pd[group_ggh_pd$REGION_name=="Peterborough County", "Y"] <- 4921000
-```
-```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
-#plot 
-TTS_area_plot <- ggplot() +
-  geom_sf(data = TTS2016R::ggh_pd, color = "darkgray",
-          aes(fill = REGION_name)) +
-  scale_fill_manual(values = viridis::viridis(20)) +
-  geom_sf(data = group_ggh_pd_poly, 
-          color = "black", fill = NA, size = 0.7) +
-  annotation_north_arrow(location = "tl", # north arrow for both the main plot
-                         height = unit(0.8, "cm"), 
-                         width = unit(0.8, "cm"),
-                         style = north_arrow_orienteering(line_width = 0.25,
-                                                          line_col = "dimgrey", 
-                                                          fill = c("white","dimgrey"))) +
-  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot
-                   height = unit(0.15, "cm")) + 
-  geom_shadowtext(data = group_ggh_pd,
-                  aes(x = X, y = Y, label = REGION_name),
-                  size = 3,
-                  nudge_y = 3000,
-                  nudge_x = 2000) +
-  theme_void() +
-  theme(legend.position = "none",
-        axis.title = element_blank())
-
-# ggsave("images/TTS16-survey-area.png")
-```
-```{r creating-desc-stats-table}
-#forming a complete descriptive statistic table
-
-Statistics <- data.frame("Statistic" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))
-
-Trips <- data.frame("Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[2]] %>% round(),  
-                                   summary(od_ft_tt$trips)[[3]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[4]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[5]] %>% round(),
-                                   summary(od_ft_tt$trips)[[6]]%>% round(),
-                                   NA))
-
-Travel_time <- data.frame("TT" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[2]] %>% round(),  
-                                               summary(od_ft_tt$travel_time)[[3]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[4]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[5]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[6]] %>% round(),  
-                                               3507))
-
-# TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(1), 
-#                                       summary(ggh_taz$AREA)[[2]] %>% round(1), 
-#                                       summary(ggh_taz$AREA)[[3]] %>% round(1), 
-#                                       summary(ggh_taz$AREA)[[4]] %>% round(1), 
-#                                       summary(ggh_taz$AREA)[[5]] %>% round(1), 
-#                                       summary(ggh_taz$AREA)[[6]] %>% round(1), 
-#                                       NA))
-
-Workers <- data.frame("Workers" = c(summary(ggh_taz$workers)[[1]] %>% round(), 
-                                    summary(ggh_taz$workers)[[2]] %>% round(), 
-                                    summary(ggh_taz$workers)[[3]] %>% round(), 
-                                    summary(ggh_taz$workers)[[4]] %>% round(), 
-                                    summary(ggh_taz$workers)[[5]] %>% round(), 
-                                    summary(ggh_taz$workers)[[6]] %>% round(), 
-                                    NA))
-
-Jobs <- data.frame("Jobs" = c(summary(ggh_taz$jobs)[[1]] %>% round(), 
-                              summary(ggh_taz$jobs)[[2]] %>% round(), 
-                              summary(ggh_taz$jobs)[[3]] %>% round(), 
-                              summary(ggh_taz$jobs)[[4]] %>% round(), 
-                              summary(ggh_taz$jobs)[[5]] %>% round(), 
-                              summary(ggh_taz$jobs)[[6]] %>% round(), 
-                              NA))
-
-desc_stats <- cbind(Statistics, Trips, Travel_time, Workers, Jobs)
-
-# #kable tabling 
-# desc_stats %>%
-#   kable(format = "latex",
-#         align="lrrrrrr",
-#         booktabs = T,
-#         col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
-#         caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, calculated origin-destination car travel time, TAZ area, workers per TAZ, and jobs per TAZ.") %>%
-#   add_header_above(c(" ", "Trips", "Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
-#   kable_styling(full_width = "T", 
-#                 latex_options = c("scale_down"),
-#                 position = "center")
-```
-```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align='center'}
-# knitr::include_graphics("images/TTS16-survey-area.png")
-
-TTS_area_plot + annotate(geom = "table",
-            x = 915000, y = 4880000, 
-           label = list(desc_stats))
-```
-## Employed individuals and jobs
-
-The origin-destination table (i.e., trips) consists of a cross-tabulation of people who are employed full-time by place of GGH residence (origin) and places of GGH employment (destination) using the GTA06 TAZ zoning system.
-It is important to note that the total number of workers and jobs is the TTS 2016 region are not equal but the number of trips taken are equal to the number of workers.
-Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
-This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs).
-As such, the origin-destination table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
-
-```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.", fig.width=8, fig.height=5}
-
-tts_workers <- tm_shape(ggh_taz) +
-  tm_fill(col = "workers", title = "Full-time \nemployed people", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-tts_jobs <- tm_shape(ggh_taz) +
-  tm_fill(col = "jobs", title = "Full-time \njobs", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-tmap_arrange(tts_workers, tts_jobs)
-
-```
-
-<!--
-```{r ECD-plot, fig.cap="\\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set.
-Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ.", fig.width=8, fig.height=4}
-
-Number <- rbind(ggh_taz$jobs %>% data.frame(),
-                 ggh_taz$workers %>% data.frame())
-Group <- rbind(rep("Jobs", each=length(ggh_taz$jobs)) %>% data.frame(),
-                rep("Workers", each=length(ggh_taz$workers))%>% data.frame())
-
-ecdf_data <- cbind(Number, Group)
-colnames(ecdf_data) <- c("Number", "Group")
-
-rect1 <- data.frame(xmin=115, xmax=3055, ymin=-Inf, ymax=Inf)
-
-ggplot(ecdf_data , aes(x=Number, col=Group)) + 
-  # geom_segment(aes(x = 115, y = 0, xend = 115, yend = 0.34), col = "blue", linetype=2) +
-  # geom_segment(aes(x = 3055, y = 0, xend = 3055, yend = 0.94), col = "blue", linetype=2) +
-  geom_rect(data = rect1, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), fill="lightblue", alpha=0.2, inherit.aes = FALSE) +
-  stat_ecdf() + 
-  scale_x_continuous(trans='sqrt', breaks=c(115, 3055, 10000, 20000, 30000, 40000)) + 
-  scale_y_continuous(breaks=c(0, 0.25,0.34, .50, .75,0.94, 1.00)) +
-  labs(x = "Opportunities per TAZ",
-       y = "Cumulative probability") + 
-      scale_color_manual("Opportunities", 
-                         values = c("Jobs" = "Black",
-                                    "Workers" = "Red"))+
-  scale_linetype_manual("Opportunities",values=c(0,1)) +
-  geom_point(aes(x = 115, y = 0.34), col = "blue")+
-  geom_point(aes(x=3055, y=0.94), col = "blue") +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust=0.5),
-        legend.position = c(0.90,0.75),
-        axis.line = element_line(colour = "black"),
-        panel.border = element_blank()) 
-```
--->
-Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ.
-It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs-housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
-It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
-<!-- see the following cumulative probability distribution in Figure \ref{fig:ECD-plot}.
--->
-
-<!-- Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH.
-The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022].
-The south east border of the GGH neighbours Lake Ontario and is delineated by the urban built boundary of the Ontario Growth Plan being home to the highest density of working population in the GGH [@ontario_built_2019;@auditor_general_of_ontario_value_2021].
-The northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake, and Peterborough regions which are home to lower density of worker population density population [@auditor_general_of_ontario_value_2021].
-Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs.
--->
-
-<!--- in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ.
-This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}.
--->
-
-## Calculated travel time
-
-As mentioned, {TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Table \ref{tab:TTS-16-desc-stats}).
-It is important to note that travel times within this data set are calculated assuming only car travel and one departure time for all origins.
-These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies.
-However, it is not possible from the data retrieval system to obtain higher order tabulations so <!--  we do not know which trips are made with non-car modes nor exact departure times in these tables.
-Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity,-->  we carry on with the assume that all trips are taken by one-time departure car trip.
-
-```{r data-for-impedance}
-# remove all NA trips from dataset and set all 0min travel times to 0.1 min
-od_ft_tt  <- TTS2016R::od_ft_tt %>% 
-  filter( !is.na(travel_time)) %>% 
-  mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))
-
-all_tt <- od_ft_tt  %>% 
-  dplyr::select(trips, travel_time)
-
-all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
-```
-
-```{r fitting-impedance}
-#fitting impedance function
-gamma_ <- fitdistrplus::fitdist(data=all_tt, "gamma", method="mle", optim.method="Nelder-Mead") 
-```
-
-```{r calc-for-accessibility-GGH}
-# transfer calibrated impedance function values to OD matrix
-od_ft_tt <- od_ft_tt %>%
-  mutate(f = dgamma(travel_time, gamma_$estimate["shape"], gamma_$estimate["rate"]))
-
-#add the number of jobs and workers to the od_ft_tt matrix
-od_ft <- od_ft_tt %>% merge(ggh_taz %>% dplyr::select(GTA06, workers) %>% st_drop_geometry(),
-                   by.x = "Origin", by.y="GTA06", all.x = TRUE)
-
-od_ft <- od_ft %>% merge(ggh_taz %>% dplyr::select(GTA06, jobs) %>% st_drop_geometry(),
-                   by.x = "Destination", by.y="GTA06", all.x = TRUE)
-
-#calculate accessibility for workers from any origin to jobs in Toronto 
-GGH_c_accessibility <- od_ft %>% 
-  mutate(GGH_A_ij = f * jobs) %>%
-  group_by(Origin) %>%
-  summarise(GGH_A_i = sum(GGH_A_ij, na.rm = T),
-            GGH_sum_tt_i = sum(travel_time),
-            GGH_tt_trips_i = mean(travel_time),
-            GGH_sum_f_i = sum(f),
-            GGH_f_trips_i = mean(f))
-
-#Merge TO accessibly calculation to the ggh_taz:
-GGH_taz_acc <- ggh_taz %>% merge(GGH_c_accessibility, by.x=c("GTA06"), by.y=c("Origin"), all.x=T) 
-```
-
-```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.", fig.width=8, fig.height=5, message=FALSE}
-
-tts_total_tt <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "GGH_sum_tt_i", title = "Total travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-tts_tt_per_trip <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "GGH_tt_trips_i", title = "Avg. travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
-  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
-  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
-  tm_layout(legend.position = c("right", "bottom"))
-
-tmap_arrange(tts_total_tt, tts_tt_per_trip)
-```
-
-\newpage
-
-As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the average travel time is distinct from other plots presented so far.
-We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA)<!-- (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) -->, Niagara and Waterloo, the average travel times are moderately low.
-Further from these areas, travel times are higher.
-Interestingly, even in eastern areas (e.g., Peterborough) with high employment and high job concentration, average travel time is higher than within the GTHA.
-
-## Calibrating an impedance function
-
-Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@talen_assessing_1998;@paez_jobs_2013;@barboza_balancing_2021].
-An impedance function $f(\cdot)$ depends on the cost of travel $c_{ij}$ between locations $i$ and $j$ (all which is supplied in the travel time and origin-destination table within {TTS2016R}).
-<!-- The impedance function is usually a monotonically decreasing function since the separation of land uses means that very short commuting trips are relatively rare.
-However, sometimes the function can increase to reflect patterns of separation between activities.
-This could be needed to reflect fluctuations in travel cost due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration.
--->
-
-A useful technique to calibrate an impedance function is to use the trip length distribution (TLD) as measured from origin-destination data [@horbachov_theoretical_2018; @batista_estimation_2019].
-The TLD is the representation of the likelihood that a proportion of trips are taken at a specific travel cost.
-In our data set, where we assume cost is travel time, the impedance function maps low travel times to higher proportions of trips, and high travel times are mapped to low proportion of trips.
-
-Using the data contained in {TTS2016R}, we fit the empirical TLD to a density distribution using maximum likelihood techniques and the Nelder-Mead method for direct optimization available within the `R` package {fitdistrplus} [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics seen in Figure \ref{fig:TLD-Gamma-plot}, the gamma distribution is selected.
-The 'shape' parameter is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).
-
-```{r save-impedance-plot, include=FALSE}
-# For some reason plot(gamma_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
-png("images/impedance_function.png")
-plot(gamma_)
-dev.off()
-```
-
-```{r TLD-Gamma-plot, fig.cap="\\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car TLD (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots", out.width="75%", fig.align='center'}
-knitr::include_graphics("images/impedance_function.png")
-```
-
-\begin{equation}
-\label{gamma-dist}
-\begin{array}{l}\ 
-f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for }	0 \leq x \leq \infty\\
-\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
-\end{array}
-\end{equation}
-
-```{r create-cullen-frey-data}
-# remove all NA trips from dataset and set all 0min travel times to 0.1min
-od_ft_tt  <- od_ft_tt %>% filter( !is.na(travel_time)) %>% mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))
-all_tt <- od_ft_tt  %>% dplyr::select(trips, travel_time)
-all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
-```
-
-<!-- # ```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."} -->
-<!-- # fitdistrplus::descdist(data=all_tt)  -->
-<!-- # ``` -->
-
-\newpage
-
-# Concluding remarks
-
-The open data product introduced in this paper shares tables for home-to-work related data from the 2016 TTS.
-In addition, inter-centroid travel time tables are calculated, and the planning/municipality boundaries are included as a compliment.
-This open data product, {TTS2016R}, is freely available to explore in an `R` environment.
-One possible use of this data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.
-
-<!-- New digital formats are increasingly complex and the explanation of the methods often do not concisely and intelligibly fit within the confines of a traditional article.
-With this motivation, we invite all who are interested to use {TTS2016R} to explore the worker-employment patterns contained in the {TTS2016R} package.
---> In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own.
-We see this product as providing open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
-
-# References {#references .unnumbered}

### 2022-03-17 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.tex
additional changes to manuscript - final

--- 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.tex
@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-
 \begin{document}
@@
 
-\title{TTS2016R: A data set to study population and employment patterns
+\title{TTS2016R: A dataset to study population and employment patterns
 from the 2016 Transportation Tomorrow Survey (TTS) in the Greater
-Toronto and Hamilton Area in Canada}
+Toronto and Hamilton Area, Canada}
 
@@
 
-\author{Anastasia Soukhov\affilnum{}, Antonio Paez\affilnum{}}
+\author{Anastasia Soukhov\affilnum{}, Antonio Páez\affilnum{}}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-This manuscript presents the open data product
-\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
-products are the result of turning source data (open or otherwise) into
-accessible information that adds value to the original inputs
-\citep[see][]{Arribas2021open}. The product presented in this paper is
-an \texttt{R} data package which currently consists of three objects
-which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or
-are curated to facilitate the use and analysis of TTS data. This package
-includes person-to-jobs origin-destinations, traffic analysis zone (TAZ)
-boundaries, and planning/municipality boundaries for the Greater Golden
-Horse area (GGH) located in southern Ontario, Canada
-\citep{data_management_group_tts_2018}. In addition, the package
-includes TAZ centroid-to-centroid travel times by car computed using
-package \{r5r\} \citep{Pereira2021r5r}. The aim of this paper is to walk
-readers through the empirical home-based work commute data set,
-illustrate the calculation of an impedance function that can be used to
-calculate accessibility to employment, and invite its use in other
-applications.
-
-\hypertarget{home-to-work-commute-data}{%
-\section{Home-to-work commute data}\label{home-to-work-commute-data}}
-
-\{TTS2016R\} includes counts of fully-employed population by place of
-residence (origin) and counts of full-time jobs by place of work
-(destination) aggregated by TAZ (n=3,764 within the survey boundaries).
-TAZ typically are defined based on land-use and population demographics
-in order to estimate the number of trips produced and attracted to each
-zone \citep{meyer_urban_2001}. As such, each TAZ is uniquely identified
-using the GTA06 Zoning System which can be used to join to the
-origin-destination table (i.e., trips taken).
-
-The number of jobs (3,081,885) and workers (3,446,957) in this package
-are organized in the form of an origin-destination table which is
-indicative of home-to-work commute patterns (there are 3,446,957
-potential interactions). These data were retrieved from the TTS Data
-Retrieval System on October 28, 2021 and reflect the potential
-interaction of full-time employed people and jobs within the GGH survey
-boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by
-the 2016 TTS methodology \citep{data_management_group_tts_2018}.
-
-Also included in \{TTS2016R\} are travel times and cost of travel from
-origin to destination by car; travel times are calculated using the
-\texttt{R} package \{r5r\}. These travel times are useful to estimate
-the cost of travel and to calculate impedance functions, among other
-possible applications. For simplicity, all interactions within
-\{TTS2016R\} are assumed to be taken by car, and the travel time is
-calculated from an origin TAZ centroid to a destination TAZ centroid.
-The centroid is snapped to the nearest street line by \texttt{r5r} and
-the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs)
-are calculated; this threshold represents 99\% of trip's travel times
-which are summarized in Figure \ref{fig:TTS-16-survey-area}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/TTS-16-survey-area-1}
-
-}
-
-\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
-\end{figure}
-
-\hypertarget{employed-individuals-and-jobs}{%
-\subsection{Employed individuals and
-jobs}\label{employed-individuals-and-jobs}}
-
-The origin-destination table (i.e., trips) consists of a
-cross-tabulation of people who are employed full-time by place of GGH
-residence (origin) and places of GGH employment (destination) using the
-GTA06 TAZ zoning system. It is important to note that the total number
-of workers and jobs is the TTS 2016 region are not equal but the number
-of trips taken are equal to the number of workers. Since the outer
-boundaries of the TTS are permeable, workers who reside within the
-boundaries but travel outside of the boundaries are counted as workers
-within an origin TAZ, while jobs in TAZ that are filled by workers who
-reside outside the GGH boundaries are \emph{unknown} since they were not
-surveyed. This mismatch results in the total number of workers being
-1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
-3,081,885 jobs). As such, the origin-destination table contained in
-\{TTS2016R\} offers a perspective on all workers in the GGH and their
-home-based trips to places of GGH employment.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
-\end{figure}
-
-Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers
-and jobs per TAZ. It can be observed that the spatial distribution of
-jobs and workers is unequal, which is indicative of a jobs-housing
-imbalance that can impact accessibility in a region
-\citep{Levine1998rethinking}. It can also be seen that there is a higher
-number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
-with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
-is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
-number TAZ with an extreme number of jobs at the highest and lowest
-percentiles is significantly higher than the number of workers.
-
-\hypertarget{calculated-travel-time}{%
-\subsection{Calculated travel time}\label{calculated-travel-time}}
-
-As mentioned, \{TTS2016R\} also includes travel time data for each
-home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R
-package \{r5r\} (see descriptive statistics in Table
-\ref{tab:TTS-16-desc-stats}). It is important to note that travel times
-within this data set are calculated assuming only car travel and one
-departure time for all origins. These assumptions are not completely
-realistic since we know a small proportion of trips are taken by non-car
-modes and travel time departures varies. However, it is not possible
-from the data retrieval system to obtain higher order tabulations so we
-carry on with the assume that all trips are taken by one-time departure
-car trip.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.}\label{fig:plot-tt-ttpertrip}
-\end{figure}
-
-\newpage
-
-As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
-travel time resembles the spatial trend distribution in the number of
-employed people in the previous plot (Figure
-\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
-average travel time is distinct from other plots presented so far. We
-can see that in areas around the south-eastern border that make up the
-Greater Toronto and Hamilton Area (GTHA), Niagara and Waterloo, the
-average travel times are moderately low. Further from these areas,
-travel times are higher. Interestingly, even in eastern areas (e.g.,
-Peterborough) with high employment and high job concentration, average
-travel time is higher than within the GTHA.
-
-\hypertarget{calibrating-an-impedance-function}{%
-\subsection{Calibrating an impedance
-function}\label{calibrating-an-impedance-function}}
-
-Impedance functions are useful to understand mobility behaviour and are
-used to estimate gravity models of spatial interaction
-\citep{wilson1971, haynes_gravity_1985} and applied in accessibility
-analysis
-\citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
-An impedance function \(f(\cdot)\) depends on the cost of travel
-\(c_{ij}\) between locations \(i\) and \(j\) (all which is supplied in
-the travel time and origin-destination table within \{TTS2016R\}).
-
-A useful technique to calibrate an impedance function is to use the trip
-length distribution (TLD) as measured from origin-destination data
-\citep{horbachov_theoretical_2018, batista_estimation_2019}. The TLD is
-the representation of the likelihood that a proportion of trips are
-taken at a specific travel cost. In our data set, where we assume cost
-is travel time, the impedance function maps low travel times to higher
-proportions of trips, and high travel times are mapped to low proportion
-of trips.
-
-Using the data contained in \{TTS2016R\}, we fit the empirical TLD to a
-density distribution using maximum likelihood techniques and the
-Nelder-Mead method for direct optimization available within the
-\texttt{R} package \{fitdistrplus\} \citep{fitdistrplus_2015}. Based on
-goodness-of-fit criteria and diagnostics seen in Figure
-\ref{fig:TLD-Gamma-plot}, the gamma distribution is selected. The
-`shape' parameter is \(\alpha\) = 2.019, the estimated `rate' is
-\(\beta\) = 0.094 , and \(\Gamma(\alpha)\) is defined in Equation
-(\ref{gamma-dist}).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{images/impedance_function}
-
-}
-
-\caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car TLD (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
-\end{figure}
-
-\begin{equation}
-\label{gamma-dist}
-\begin{array}{l}\ 
-f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
-\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
-\end{array}
-\end{equation}
-
-\newpage
-
-\hypertarget{concluding-remarks}{%
-\section{Concluding remarks}\label{concluding-remarks}}
-
-The open data product introduced in this paper shares tables for
-home-to-work related data from the 2016 TTS. In addition, inter-centroid
-travel time tables are calculated, and the planning/municipality
-boundaries are included as a compliment. This open data product,
-\{TTS2016R\}, is freely available to explore in an \texttt{R}
-environment. One possible use of this data, as showcased in this paper,
-is the calibration of impedance functions which in turn can be used for
-accessibility analysis.
-
-In the spirit of novel and original research, we hope readers value the
-efforts made to detail the data in order to improve transparency in our
-work and encourage others to replicate and, hopefully, inspire research
-of their own. We see this product as providing open infrastructure for
-additional TTS or complimentary data sets to be amended by the authors
-or wider open-source community in the future.
-
 \bibliographystyle{sageh}

### 2022-03-17 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
additional changes to manuscript - final

@@
 ```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align='center'}
-# knitr::include_graphics("images/TTS16-survey-area.png")
-
-TTS_area_plot + annotate(geom = "table",
+Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align = 'center', message=FALSE, warning=FALSE}
+TTS_area_plot <- TTS_area_plot + annotate(geom = "table",
             x = 915000, y = 4880000, 
            label = list(desc_stats))
+ggsave("images/TTS16-survey-area.png")
+knitr::include_graphics("images/TTS16-survey-area.png")
 ```
+
 ## Employed individuals and jobs
@@
 As mentioned, {TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Table \ref{tab:TTS-16-desc-stats}).
+This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Figure \ref{fig:TTS-16-survey-area}).
 It is important to note that travel times within this data set are calculated assuming only car travel and one departure time for all origins.

### 2022-03-17 Manuscript-Data-Package/Manuscript-Data-Package.tex
additional changes to manuscript - final

@@
 
-{\centering \includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/TTS-16-survey-area-1}
+{\centering \includegraphics[width=1\linewidth]{images/TTS16-survey-area}
 
@@
 This travel time corresponds to a car commute calculated using the R
-package \{r5r\} (see descriptive statistics in Table
-\ref{tab:TTS-16-desc-stats}). It is important to note that travel times
+package \{r5r\} (see descriptive statistics in Figure
+\ref{fig:TTS-16-survey-area}). It is important to note that travel times
 within this data set are calculated assuming only car travel and one

### 2022-03-19 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.Rmd
submitted this version to EPB

@@
   - times
-bibliography: bibfile.bib
 bibliographystyle: sageh

### 2022-03-19 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
submitted this version to EPB

@@
 In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r].
-The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.
-
-<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
+The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
 In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
@@
 ```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align = 'center', message=FALSE, warning=FALSE}
+Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align = 'center', fig.pos='H', message=FALSE, warning=FALSE}
 TTS_area_plot <- TTS_area_plot + annotate(geom = "table",
@@
 As such, the origin-destination table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
-
 ```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.", fig.width=8, fig.height=5}
@@
 ```
-
 <!--
@@
 <!-- see the following cumulative probability distribution in Figure \ref{fig:ECD-plot}.
--->
-
-<!-- Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH.
+--> <!-- Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH.
 The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022].
@@
 Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs.
--->
-
-<!--- in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ.
+--> <!--- in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ.
 This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}.
@@
 ```
-
 \newpage
@@
 The 'shape' parameter is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).
-
 ```{r save-impedance-plot, include=FALSE}
@@
 ```
-
 <!-- # ```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."} -->
 <!-- # fitdistrplus::descdist(data=all_tt)  -->
-<!-- # ``` -->
-
-\newpage
+<!-- # ``` -->\newpage
 

### 2022-10-22 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

@@
 The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data.
-This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018].
-In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r].
-The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
-In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
-By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region.
-It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
--->
+This package includes full-time home-based work-to-job origin destinations counts and mode-specific trip numbers, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018].
+In addition, the package includes TAZ centroid-to-centroid travel times by car, transit, cyling, and walking mode computed using package {r5r} [@Pereira2021r5r].
+
+The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.
+Though data from the TTS is freely available to the public (URL <!-- -->), the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
+By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region and amends calculated travel times to the data set.
+This package also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
 
 # Home-to-work commute data
+<!-- NOTE: When extracting trip vs.
+person data in the TTS database.
+For 'by person' extraction: Trips made by person include all trips.
+For 'by trip' extraction: trips made do not include walking trips to non-school or non-work purposes, so the total number of trips extracted through this method is NOT equal to the total number of trips per person (the total number of trips per person is always higher than the total number of trips).
+You can't filter the 'by person' extraction by trip purpose.
+You can filter the 'by trip' extraction by trip purpose, so we will be using the 'by trip' extraction - filtered by 'home-based work' - to get the primary mode trip numbers.
+-->
+{TTS2016R} includes counts of full-time employed population by place of residence (origin), counts of full-time usual place of work (destination), number of trips to work by mode, average estimated travel time of trip, and the number of vehicles by mode.
+This data is aggregated and available at the level of TAZ: TAZ are a spatial unit of analysis typically used to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
+They are thus defined by transportation planners for a region based on intra-similarity and inter-dissimilarity between land-use and population demographics.
+Within the GGH boundaries, `r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` TAZ are specified.
+The TAZ range between ≥`r 0.019` km^2 in spatial area to a maximum of `r max(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 (median: `r median(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 and 3rd quantile: `r quantile(TTS2016R::ggh_taz$AREA, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")` km^2).
 
-{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries).
-TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
-As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).
+The number of full-time working population (will be referred to as 'workers' throughout) and full-time places of employment (will be referred to as 'jobs' throughout) is also quite variable between TAZ.
+In terms of workers, TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`).
+In terms of jobs, TAZ contain between `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`).
 
+Each TAZ is uniquely identified using the GTA06 Zoning System which can be used to match the worker residential location to the job location in the origin-destination table.
 The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions).
-These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
+This data was retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the interaction of full-time employed people to their places of employment as shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
 
-Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}.
-These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications.
-For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid.
-The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
+Also included in {TTS2016R} are the estimated travel times from origin to destination by various modes; travel times are calculated using the package {r5r} and the notebook for this calculation is availalbe in the documentation of {TTS2016R}.
+These travel times are useful addition to {TTS2016R} since they are not included in the TTS Data Retrieval System but they are vitally important to estimate the cost of travel and associated impedance functions, among other possible applications.
+The included travel times are by car, walking, cyling, and public transit modes and are calculated from each origin TAZ centroid to each destination TAZ centroid where a home-based work trip is made (`r TTS2016R::od_ft_tt$Origin %>% length() %>% prettyNum(big.mark = ",") ` OD pairs).
+The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all pairs.
+Additionally, for car and transit modes only travel times less than or equal to 180 mins are calculated.
+For cycling and walking modes only travel times less than or equal to 120 mins are calculated.
+
+These threshold represents 90% or greater of trip's travel times, per mode, which are summarized in Figure \ref{fig:TTS-16-survey-area}.
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}

### 2022-10-22 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

@@
 library(units)
+library(tmap)
+library(ggplot2)
 ```
@@
 
-Read the boundaries of the traffic analysis zones:
+Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
 ```{r}
@@
 
-## Origin destination matrices for work and shop
+add 'REGION' from the planning distract object to the `ggh_taz`. 
+```{r}
+#buffering to clean up the edges of these objects -> there wre issues using st_join. 
+ggh_taz <- st_buffer(ggh_taz, dist=0)
+ggh_pd <- st_buffer(ggh_pd, dist=0)
 
-Read trip tables. The query was for "trips" by zone of residence, zone of work, and trip purpose (Work):
+ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
+```
+
+## CMA and CA boundaries
+If there's interest in tying the TAZ to the Canadian Census, the boundaries and IDs for census metropolitan areas (CMA) and census ares (CA) are provided in this package as well.
+These areas are determined by Statistics Canada and
+
+Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
+```{r}
+ggh_cma <- st_read("TTS16-data-inputs/Boundaries/lcma000b16a_e.shp")
+```
+
+Project object and retain specific columns:
+```{r}
+ggh_cma <- ggh_cma %>%
+  st_transform(crs = 32617) %>%
+  transmute(CMAUID,
+            CMANAME,
+            CMATYPE)
+```
+
+add 'CMAUID' from the cma object to the `ggh_taz`. 
+```{r}
+#doing a st_join doesn't provide a clean solution, so I find the centroids of each TAZ, st_join them to the CMA/CA, and then join the resulting CMAUID property to the ggh_taz object
+test_pts <-st_centroid(ggh_taz)
+test<-st_join(test_pts, ggh_cma %>% transmute(CMAUID), left = TRUE, join=st_within)
+ggh_taz <-merge(ggh_taz, test %>% transmute(GTA06, CMAUID)%>% st_drop_geometry(), by=c("GTA06"))
+```
+
+## Origin destination matrices for place of residence and typical place work (by person)
+Read exported TTS 2016 tables.
+The query was a cross-tabulation of "persons" by zone of residence and place of employment.
+The data was filtered to only include full-time not-remote workers and was retrieved from the TTS 2016:
 ```{r}
@@
 
+## Trips taken (by mode) by worker from place of residence to job at place of work
+Read exported TTs 2016 tables.
+The query was a cross-tablulation of "trips" by zone of residence and place of employment.
+The data was filtered to only include full-time not-remote working trip purposes by primary mode and was retrieved from the TTS 2016:
+```{r}
+od_ft_trips <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment-by_W_R_trip-byprimmode.txt", 
+                      delim = "\t", 
+                      col_names = FALSE)
+```
+
+Find the position in the table where the information on work trips begins:
+```{r}
+id_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Transit excluding GO rail)")
+id_cycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Cycle)")
+id_car_driver <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto driver)")
+id_GO <- which(od_ft_trips$X1 == "TABLE    : mode_prime (GO rail only)")
+id_GO_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Joint GO rail and local transit)")
+id_motorcycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Motorcycle)")
+id_other <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Other)")
+id_car_pass <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto passenger)")
+id_schoolbus <- which(od_ft_trips$X1 == "TABLE    : mode_prime (School bus)")
+id_taxi <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Taxi passenger)")
+id_rideshare <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Paid rideshare)")
+id_walk <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Walk)")
+id_end <- od_ft_trips$X1 %>% length()
+```
+
+Slice the table to obtain the work trips by primary mode (there are 12 possible modes, so 12 new tables) (remove heading information and other modes):
+```{r}
+od_ft_trips_transit <- od_ft_trips %>% 
+  slice((id_transit+3):id_cycle-1)
+
+od_ft_trips_cycle <- od_ft_trips %>% 
+  slice((id_cycle+3):id_car_driver-1)
+
+od_ft_trips_car_driver <- od_ft_trips %>% 
+  slice((id_car_driver+3):id_GO-1)
+
+od_ft_trips_GO <- od_ft_trips %>% 
+  slice((id_GO+3):id_GO_transit-1)
+
+od_ft_trips_GO_transit <- od_ft_trips %>% 
+  slice((id_GO_transit+3):id_motorcycle-1)
+
+od_ft_trips_motorcycle <- od_ft_trips %>% 
+  slice((id_motorcycle+3):id_other-1)
+
+od_ft_trips_other <- od_ft_trips %>% 
+  slice((id_other+3):id_car_pass-1)
+
+od_ft_trips_car_pass <- od_ft_trips %>% 
+  slice((id_car_pass+3):id_schoolbus-1)
+
+od_ft_trips_schoolbus <- od_ft_trips %>% 
+  slice((id_schoolbus+3):id_taxi-1)
+
+od_ft_trips_taxi<- od_ft_trips %>% 
+  slice((id_taxi+3):id_rideshare-1)
+
+od_ft_trips_rideshare <- od_ft_trips %>% 
+  slice((id_rideshare+3):id_walk-1)
+
+od_ft_trips_walk <- od_ft_trips %>% 
+  slice((id_walk+2):id_end)
+```
+
+Let's consolidate the 12 tables into 1 table, based on origin - destination and clean up the table. If there's no trip using that mode, a 0 is added. The following are the codes:
+
+- T_LT - Local transit excluding GO rail
+- T_LTGO - Local transit and GO
+- T_GO - GO rail excluding local transit
+```{r}
+od_ft_trips_transit <- od_ft_trips_transit %>%
+  separate(X1, into = c("Zones", "T_LT"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_LT = as.numeric(T_LT))
+
+od_ft_trips_GO_transit <- od_ft_trips_GO_transit %>%
+  separate(X1, into = c("Zones", "T_LTGO"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_LTGO = as.numeric(T_LTGO))
+
+od_ft_trips_GO <- od_ft_trips_GO %>%
+  separate(X1, into = c("Zones", "T_GO"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_GO = as.numeric(T_GO))
+```
+
+- BC - Cycle
+- W - Walk
+```{r}
+od_ft_trips_cycle <- od_ft_trips_cycle %>%
+  separate(X1, into = c("Zones", "BC"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         BC = as.numeric(BC))
+
+od_ft_trips_walk <- od_ft_trips_walk %>%
+  separate(X1, into = c("Zones", "W"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         W = as.numeric(W))
+```
+
+- C_DPC - Driver private car 
+- C_PPC - Passenger private car
+- C_TA - Taxi
+- C_RS - Rideshare
+- C_SB - School bus
+```{r}
+od_ft_trips_car_driver <- od_ft_trips_car_driver %>%
+  separate(X1, into = c("Zones", "C_DPC"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_DPC = as.numeric(C_DPC))
+
+od_ft_trips_car_pass <- od_ft_trips_car_pass %>%
+  separate(X1, into = c("Zones", "C_PPC"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_PPC = as.numeric(C_PPC))
+
+od_ft_trips_taxi <- od_ft_trips_taxi %>%
+  separate(X1, into = c("Zones", "C_TA"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_TA = as.numeric(C_TA))
+
+od_ft_trips_rideshare <- od_ft_trips_rideshare %>%
+  separate(X1, into = c("Zones", "C_RS"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_RS = as.numeric(C_RS))
+
+od_ft_trips_schoolbus <- od_ft_trips_schoolbus %>%
+  separate(X1, into = c("Zones", "C_SB"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_SB = as.numeric(C_SB))
+```
+
+- M - Motorcycle
+- O - Other
+```{r}
+od_ft_trips_motorcycle<- od_ft_trips_motorcycle %>%
+  separate(X1, into = c("Zones", "M"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         M = as.numeric(M))
+
+od_ft_trips_other <- od_ft_trips_other %>%
+  separate(X1, into = c("Zones", "O"), sep = " (?=[^ ]+$)") %>%
+  mutate(Zones = str_trim(Zones, side = "both")) %>%
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         O = as.numeric(O))
+```
+
+And now, consolidate all 12 tables into 1 table! with 12 columns for mode trips numbers + the origin and destination columns (those are GTA06 TAZ IDs):
+```{r}
+od_ft_trips <- full_join(od_ft_trips_transit, od_ft_trips_GO_transit, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_GO, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_cycle, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_walk, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_car_driver, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_car_pass, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_taxi, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_rideshare, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_schoolbus, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_motorcycle, by=c("Origin", "Destination")) %>%
+  full_join(od_ft_trips_other, by=c("Origin", "Destination")) 
+```
+
+Replace all NAs with 0s:
+```{r}
+od_ft_trips[is.na(od_ft_trips)] <- 0
+```
+
+Combine the `od_ft` person data frame with the `od_ft_trips` data frame.
+Notice that there are more OD pairs in the `od_ft` data frame (`r od_ft$Origin %>% length()`) than `od_ft_trips` (`r od_ft_trips$Origin %>% length()`) because some OD pairs which have workers and/or are a location for jobs do not have any trips taken that survey day.
+The TTS 2016 is a travel survey after all.
+```{r}
+od_ft_p_t <- full_join(od_ft, od_ft_trips, by=c("Origin", "Destination")) 
+```
+
+Check to see if the total number of trips adds up to the total available from TTS 2016 (to ensure we separated and merged all the tables correctly)
+```{r}
+od_ft_p_t <- od_ft_p_t %>% rowwise() %>%
+  mutate(sum_trips = sum(T_LT,T_LTGO,T_GO,BC,W,C_DPC,C_PPC,C_TA,C_RS,C_SB,M,O, na.rm=T))
+
+#the number of total trips.
+This value is slightly higher than the value for total trips (not disagg.
+by mode, which is equal to 3,282,539 - I checked in excel) as available from the TTS 2016.
+However, I checked the summed up total trip, disaggregated by mode, in excel, and it yields this number of 3,282,611 as well.
+So there must just be some rounding up that occurred in the TTS 2016 when disaggregating by mode...
+regardless the numerical difference is insignificant.
+od_ft_p_t$sum_trips %>% sum()
+```
+
+Replace all NAs with 0s, again:
+```{r}
+od_ft_p_t[is.na(od_ft_p_t)] <- 0
+```
+
 ## Workers and Jobs
 
-The od table can be used to calculate the workers and jobs by traffic analysis zone:
+The od table can be used to calculate the count of workers, count of jobs, and count of % transit, bike, and walk slit, by traffic analysis zone:
 ```{r}
-workers_ft <- od_ft %>%
+workers_ft <- od_ft_p_t %>%
   group_by(Origin) %>%
@@
 
-jobs_ft <- od_ft %>%
+jobs_ft <- od_ft_p_t %>%
   group_by(Destination) %>%
@@
             jobs)
+```
 
+Active travel modal splits:
+```{r}
+LTGO_split_ft <- od_ft_p_t %>%
+  group_by(Origin) %>%
+  summarize(LTGO_split = sum(T_LT, T_LTGO, T_GO)/sum(sum_trips)) %>% 
+  transmute(GTA06 = Origin,
+            LTGO_split)
+
+W_split_ft <- od_ft_p_t %>%
+  group_by(Origin) %>%
+  summarize(W_split = sum(W)/sum(sum_trips)) %>%
+  transmute(GTA06 = Origin,
+            W_split)
+
+BC_split_ft <- od_ft_p_t %>%
+  group_by(Origin) %>%
+  summarize(BC_split = sum(BC)/sum(sum_trips)) %>%
+  transmute(GTA06 = Origin,
+            BC_split)
 ```
@@
             by = "GTA06") %>%
-  # Replace NAs with zeros
-  mutate(workers = replace_na(workers, 0),
-         jobs = replace_na(jobs, 0))
+  left_join(LTGO_split_ft,
+            by = "GTA06") %>%
+  left_join(W_split_ft,
+            by = "GTA06") %>%
+  left_join(BC_split_ft,
+            by = "GTA06")
 ```
+
+Replace NAs with 0s for the new columns:
+```{r}
+ggh_taz <- ggh_taz %>%
+  mutate_at(vars(workers, jobs, LTGO_split, W_split, BC_split), ~replace_na(., 0))
+```
+
+## Visualize
 
@@
   scale_fill_fermenter(direction = 1)
+```
+Plot transit split (0 to 1, 1 being 100% of trips made using transit) distribution in the ggh. Aggregating by CMA could be interesting:
+```{r}
+ggplot() +
+  geom_sf(data = ggh_taz,
+          aes(fill = LTGO_split),
+          color = NA) +
+  scale_fill_distiller(direction = 1)
+```
+Let's zoom into Toronto CMA, and plot the transit split
+```{r}
+#rotating the WGS84 crs so it displays Toronto aesthetically on a page; 17degree shift using +gamma
+# lat_0 and lonc are the approximate centroid of Toronto
+crs_string <- "+proj=omerc +lat_0=43.6532 +lonc=-79.3832 +alpha=0 +k_0=.7 +datum=WGS84 +units=m +no_defs +gamma=17"
+
+#the bounding box, so we focus on Toronto
+toronto_bb <- st_bbox(ggh_cma %>% filter(CMANAME %in% c("Toronto")))
+
+tm_shape(ggh_taz, bbox = toronto_bb, projection=crs_string) +
+  tm_polygons(col = "LTGO_split", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_cma %>% filter(CMANAME %in% c("Toronto")), bbox = toronto_bb, projection=crs_string) +
+  tm_borders()
+```
+The CMA/CAs overlapping all the TAZ:
+```{r}
+tm_shape(ggh_taz, projection=crs_string) +
+  tm_polygons(col = "CMAUID", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_cma, projection=crs_string) +
+  tm_borders()
+```
+
+The planning boundaries overlapping all the TAZ:
+```{r}
+tm_shape(ggh_taz, projection=crs_string) +
+  tm_polygons(col = "REGION", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_pd, projection=crs_string) +
+  tm_borders()
 ```
@@
 save("ggh_pd", file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
-save("od_ft", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
+save("ggh_cma", file = "TTS16-data-inputs/Boundaries/ggh_cma.Rdata")
+save("od_ft_p_t", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_p_t.Rdata")
 ```

### 2022-10-22 data-raw/01-TTS16-Calculating-Travel-Time-OD.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

@@
 ```{r}
+# load(file = "TTS16-data-inputs/Travel-Time-Calculations/results/output_OD_car1.Rdata")
 #Trips taken from origin to destination and associated people
-load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft.Rdata")
+load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_p_t.Rdata")
 ```
@@
 ```{r}
-od_ft_tt <- od_ft %>% 
-  merge(output_OD_car1, by.x=c("Origin", "Destination"), by.y=c("fromId", "toId"), all.x=T) %>%
-  rename(trips = Persons)
+od_ft_p_t_tt <- od_ft_p_t %>% 
+  merge(output_OD_car1, by.x=c("Origin", "Destination"), by.y=c("fromId", "toId"), all.x=T)
 
 rm(output_OD_car1)
-summary(od_ft_tt)
+summary(od_ft_p_t_tt)
 ```
@@
 ```{r}
-save("od_ft_tt", file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
+od <- od_ft_p_t_tt
+rm(od_ft_p_t_tt)
+save("od", file = "TTS16-data-inputs/OD-by-FT-Employment/od.Rdata")
 ```

### 2022-10-22 data-raw/02-TTS16-Imp-Function-Trip-Length-Distribution.Rmd
additions to the TTS2016R package and minor edits to the manuscript (more work to be done).

@@
 ```{r}
-load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
-summary(od_ft_tt)
+load(file = "TTS16-data-inputs/OD-by-FT-Employment/od.Rdata")
+summary(od)
 ```
@@
 ```{r}
-od_ft_tt <- od_ft_tt %>% mutate(travel_time = ifelse( is.na(travel_time) & Destination != "9998" & Destination != "8888" & Destination != "9999", 180, travel_time),
+od <- od %>% mutate(travel_time = ifelse( is.na(travel_time) & Destination != "9998" & Destination != "8888" & Destination != "9999", 180, travel_time),
                                 travel_time = ifelse( is.na(travel_time) & (Destination == "9998" | Destination == "8888" | Destination == "9999"), 300, travel_time))
-summary(od_ft_tt)
+summary(od)
 ```
@@
 ```{r}
-tld <- od_ft_tt %>%
+tld <- od %>%
   mutate(tt_classes = cut(travel_time, 
@@
   group_by(tt_classes) %>%
-  summarize(trips = sum(trips),
+  summarize(Persons = sum(Persons),
             travel_time = mean(travel_time))
@@
 ggplot(data = tld, 
-       aes(x =travel_time, y = trips)) +
+       aes(x =travel_time, y = Persons)) +
   geom_point()
@@
 ```{r}
-null_dest_trips_numb <- od_ft_tt %>% filter(travel_time=="300") %>% summarize(trips = sum(trips))
-totaltrip_trips_numb <- od_ft_tt %>% summarize(trips = sum(trips))
-dest_180_trips_numb <- od_ft_tt %>% filter(travel_time=="180") %>% summarize(trips = sum(trips))
+null_dest_trips_numb <- od %>% filter(travel_time=="300") %>% summarize(Persons = sum(Persons))
+totaltrip_trips_numb <- od %>% summarize(Persons = sum(Persons))
+dest_180_trips_numb <- od %>% filter(travel_time=="180") %>% summarize(Persons = sum(Persons))
 
@@
 ```{r}
-test <- od_ft_tt %>% mutate(travel_time = ifelse(travel_time==180, seq(from = 180, to = 240), travel_time))
+test <- od %>% mutate(travel_time = ifelse(travel_time==180, seq(from = 180, to = 240), travel_time))
 ```
@@
 ggplot(data = test, 
-       aes(x =travel_time, y = trips)) +
+       aes(x =travel_time, y = Persons)) +
   geom_point()
@@
 ```{r}
-load(file = "TTS16-data-inputs/OD-by-FT-Employment/od_ft_tt.Rdata")
+load(file = "TTS16-data-inputs/OD-by-FT-Employment/od.Rdata")
 
-tld <- od_ft_tt %>%
+tld <- od %>%
   mutate(tt_classes = cut(travel_time, 
@@
   group_by(tt_classes) %>%
-  summarize(trips = sum(trips),
+  summarize(Persons = sum(Persons),
             travel_time = mean(travel_time))
+```
+
+Replace all NAs with 0s:
+```{r}
+od[is.na(od)] <- 0
 ```
@@
 #car travel times for all origins to all destinations trips in TTS-16, number of associated trips, and the travel time.
-usethis::use_data(od_ft_tt, overwrite = TRUE)
+usethis::use_data(od, overwrite = TRUE)
 
@@
 ```{r}
-#boundaries of interest
-# load(file = "TTS16-data-inputs/Boundaries/hamilton_cma.Rdata")
-# usethis::use_data(hamilton_cma, overwrite = TRUE)
-# 
-# load(file = "TTS16-data-inputs/Boundaries/toronto_muni_bound.Rdata")
-# usethis::use_data(toronto_muni_bound, overwrite = TRUE)
-
 load(file = "TTS16-data-inputs/Boundaries/ggh_pd.Rdata")
 usethis::use_data(ggh_pd, overwrite = TRUE)
+
+load(file = "TTS16-data-inputs/Boundaries/ggh_cma.Rdata")
+usethis::use_data(ggh_cma, overwrite = TRUE)
 ```

### 2022-10-22 R/TTS2016R-package.R

@@ -7,12 +7,13 @@
 ## usethis namespace: end
 NULL
 
-#' TTS2016R: A data package which currently consists of three objects which are sourced from the [2016
+#' TTS2016R
+#'
+#' A data package which consists of objects sourced from the [2016
 #' Transportation Tomorrow Survey (TTS)](http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction)
-#' or are curated to facilitate the use and analysis of TTS data. This package includes person-to-jobs origin
+#' or are curated to facilitate the use and analysis of TTS data. This package includes worker-to-jobs origin
 #' -destinations and traffic analysis zone (TAZ) boundaries for the Greater Golden Horse area (GGH) located in
-#' southern Ontario, Canada. The package also includes boundary files for all municipalities included in the GGH
-#' area sourced from various municipal open data portals.
+#' southern Ontario, Canada. The package also includes associated spatial boundary files.
 #'
 #' @docType package
 #' @name TTS2016R-package
@@ -21,16 +22,21 @@ NULL
 #' @references \url{https://github.com/soukhova/TTS2016R}
 NULL
 
-#' ggh_taz
+#' Traffic analysis zones boundaries in the Greater Golden Horseshoe (GGH) area and associated data.
 #'
 #' This object contains traffic analysis zones (TAZ) sourced from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.
 #'
-#' @format A simple feature class (sf) polygon object containing 3764 rows and 5 variables; each row represents a unique TAZ with associated features.
+#' @format A simple feature class (sf) polygon object containing 3764 rows and 10 variables; each row represents a unique TAZ with associated features.
 #' \describe{
 #'   \item{GTA06}{Unique ID of the traffic analysis zone (TAZ).}
 #'   \item{AREA}{Area of TAZ in units of km^2.}
-#'   \item{workers}{The number of full-time workers within each TAZ.}
-#'   \item{jobs}{The number of jobs within each TAZd. }
+#'   \item{REGION}{Retrieved from the ggh_pd object. It is a unique ID corresponding with the greater region that each planning district (PD) is part of so filtering and analysis, by PD, is simplified.}
+#'   \item{CMAUID}{Retrieved from the ggh_cma object. It is a unique identifier for each census metropolitan area (CMA)/census agglomeration (CA). Within, filtering and analysis that can be easily linked to the census data, is simplified.}
[…]

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
draft of revised manuscript

--- 2022-03-19 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
@@
 Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open].
-The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data.
-This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018].
-In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r].
-The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required.
-In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
-By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region.
-It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.
--->
+The product presented in this paper is a `R` data package which currently consists of a fusion of objects from a variety of sources: home-to-work flows and trips by mode are sourced from the 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018], estimated travel times are calculated, and boundary files from the TTS and from the Canadian Census.
+
+What is a `R` data package?
+A `R` data package contains code, data, and documentation in a standardised collection format that can be installed by `R` users through a centralized software repository such as CRAN (the Comprehensive R Archive Network) and GitHub.
+{TTS2016R} is freely available on GitHub for all to install and freely use in the spirit of open and reproducible research.
+Currently, {TTS2016R} includes full-time home-based work-to-job origin destinations counts and mode-specific trip numbers retrieved from the 2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality, planning, and census metropolitan area boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada.
+In addition, the package includes TAZ centroid-to-centroid travel times by car, transit, cycling, and walking mode computed using package {r5r} [@Pereira2021r5r].
+
+The aim of this paper is to walk readers through the data sets, illustrate a use case (i.e., the calculation of an impedance function that can be used to calculate accessibility to employment), and invite others to experiment in its uses and applications.
+Though data from the TTS is freely available to the public through the [TTS Data Retrieval System](https://dmg.utoronto.ca/idrs/index), the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
+By pre-processing the data in a `R` environment and packaging it alongside complimentary data, {TTS2016R} offers a slice of the TTS data which can be immediately used by `R` users to analysis patterns of commuting to work in the region.
+Anticipate this package to grow in the future: it currently provides an open infrastructure for additional TTS or complimentary data sets to be amended by the authors and the open-source community in the future by request.
 
 # Home-to-work commute data
+<!-- NOTE: When extracting trip vs.
+person data in the TTS database.
+For 'by person' extraction: Trips made by person include all trips.
+For 'by trip' extraction: trips made do not include walking trips to non-school or non-work purposes, so the total number of trips extracted through this method is NOT equal to the total number of trips per person (the total number of trips per person is always higher than the total number of trips).
+You can't filter the 'by person' extraction by trip purpose.
+You can filter the 'by trip' extraction by trip purpose, so we will be using the 'by trip' extraction - filtered by 'home-based work' - to get the primary mode trip numbers.
+-->
+Currently, {TTS2016R} includes counts of full-time employed population by place of residence (origin), counts of full-time usual place of work (destination), number of trips to work by mode, and the calculated potential travel time of the trips in the GGH.
+The GGH (and hence the TTS survey area) is displayed in Figure \ref{fig:TTS-16-survey-area}.
 
-{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries).
-TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
-As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).
+This data is aggregated and available at the level of TAZ: TAZ are a spatial unit of analysis typically used to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001].
+They are thus defined by transportation planners for a region based on intra-similarity and inter-dissimilarity between land-use and population demographics.
+Within the GGH boundaries, `r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` TAZ are specified and each TAZ is uniquely identified using the GTA06 Zoning System: the survey boundary is discussed in the 2016 TTS methodology and defined by the TTS [@data_management_group_tts_2018].
+The TAZ range between $\ge$`0.019` km^2 in spatial area to a maximum of `r max(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 (median: `r median(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 and 3rd quantile: `r quantile(TTS2016R::ggh_taz$AREA, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")` km^2).
 
-The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions).
-These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].
-
-Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}.
-These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications.
-For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid.
-The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-## manually readjusting the X and Y coordinate of "County of Peterborough" and "Brant" as they overlap some cities
+## manually readjusting the X and Y coordinate of "County of Peterborough" and "Brant" as the resulting label overlap some city's names on the map.
 group_ggh_pd[group_ggh_pd$REGION_name=="Brant", "X"] <- 544000.0
@@
         axis.title = element_blank()) 
-
 # ggsave("images/TTS16-survey-area.png")
+```
+```{r creating-desc-stats-table-edits}
+od <- od %>% mutate(travel_time = ifelse(Origin == Destination, 10, travel_time),
+              travel_time = ifelse(travel_time == 0, NA, travel_time))
 ```
@@
 #forming a complete descriptive statistic table
-
 Statistics <- data.frame("Statistic" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))
 
-Trips <- data.frame("Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[2]] %>% round(),  
-                                   summary(od_ft_tt$trips)[[3]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[4]] %>% round(), 
-                                   summary(od_ft_tt$trips)[[5]] %>% round(),
-                                   summary(od_ft_tt$trips)[[6]]%>% round(),
+Links <- data.frame("Links" = c(summary(od$Persons)[[1]] %>% round(), 
+                                   summary(od$Persons)[[2]] %>% round(),  
+                                   summary(od$Persons)[[3]] %>% round(), 
+                                   summary(od$Persons)[[4]] %>% round(), 
+                                   summary(od$Persons)[[5]] %>% round(),
+                                   summary(od$Persons)[[6]]%>% round(),
                                    NA))
 
-Travel_time <- data.frame("TT" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[2]] %>% round(),  
-                                               summary(od_ft_tt$travel_time)[[3]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[4]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[5]] %>% round(), 
-                                               summary(od_ft_tt$travel_time)[[6]] %>% round(),  
+Trips <- data.frame("Trips" = c(summary(od$sum_trips)[[1]] %>% round(), 
+                                   summary(od$sum_trips)[[2]] %>% round(),  
+                                   summary(od$sum_trips)[[3]] %>% round(), 
+                                   summary(od$sum_trips)[[4]] %>% round(), 
+                                   summary(od$sum_trips)[[5]] %>% round(),
+                                   summary(od$sum_trips)[[6]]%>% round(),
+                                   NA))
+
+Travel_time <- data.frame("TT" = c(summary(od$travel_time)[[1]] %>% round(), 
+                                               summary(od$travel_time)[[2]] %>% round(),  
+                                               summary(od$travel_time)[[3]] %>% round(), 
+                                               summary(od$travel_time)[[4]] %>% round(), 
+                                               summary(od$travel_time)[[5]] %>% round(), 
+                                               summary(od$travel_time)[[6]] %>% round(),  
                                                3507))
@@
 
-desc_stats <- cbind(Statistics, Trips, Travel_time, Workers, Jobs)
+desc_stats <- cbind(Statistics, Workers, Jobs, Links, Trips, Travel_time)
 
@@
 ```
-```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align = 'center', fig.pos='H', message=FALSE, warning=FALSE}
+```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
+3,507 trips were not assigned travel times as they are longer than 180 mins.
+Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.", out.width="100%", fig.align = 'center', fig.pos='H', message=FALSE, warning=FALSE}
 TTS_area_plot <- TTS_area_plot + annotate(geom = "table",
@@
 
-## Employed individuals and jobs
+## Full-time employed people and associated places of employment
 
-The origin-destination table (i.e., trips) consists of a cross-tabulation of people who are employed full-time by place of GGH residence (origin) and places of GGH employment (destination) using the GTA06 TAZ zoning system.
-It is important to note that the total number of workers and jobs is the TTS 2016 region are not equal but the number of trips taken are equal to the number of workers.
-Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
+In the GGH, there are `r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")` workers, `r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")` jobs, and `r TTS2016R::od$sum_trips %>% sum()%>% prettyNum(big.mark = ",")` work-related trips (for the 2016 TTS survey day).
+The values are organized within the origin destination (OD) table in the {TTS2016R} package and are derived from the cross-tabulation by person and by trip for the full-time employed population and associated places of employment.
+The TTS is a proportionally representative survey, hence the values included in {TTS2016R} are adjusted to reflect the GGH population.
+
+It is important to note that the total number of full-time workers and jobs in the TTS 2016 region are not equal.
+Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but have workplaces that are outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
 This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs).
-As such, the origin-destination table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
+As such, the OD table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
+
+The count of links and trips made by the full-time working population and associated full-time place of employment per unique OD pair are quite variable.
+TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), and generate between `r 0` to `r 241` trips (median: `r 15`, 3rd quantile: `r 42`).
+```{r, include=FALSE, eval=FALSE, echo=FALSE}
+#counting the number of trips per origin TAZ.
+t <- od %>% group_by(Origin) %>% summarize(count = n()) %>%
+  merge(ggh_taz %>% st_drop_geometry(), by.x=c("Origin"), by.y =c("GTA06"), all.y=TRUE) %>%
+  mutate(count = ifelse(is.na(count), 0, count))
+t$count %>% summary()
+```
+
 ```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.", fig.width=8, fig.height=5}
@@
 tts_workers <- tm_shape(ggh_taz) +
-  tm_fill(col = "workers", title = "Full-time \nemployed people", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "workers", title = "Full-time \nemployed people", palette = "YlGnBu", alpha = 0.8, style = "jenks") +
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 tts_jobs <- tm_shape(ggh_taz) +
-  tm_fill(col = "jobs", title = "Full-time \njobs", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "jobs", title = "Full-time \njobs", palette = "YlGnBu", alpha = 0.8, style = "jenks") +
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 -->
-Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ.
-It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs-housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
+Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed people and associated jobs per TAZ.
+It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs -housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
 It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
@@
 
-As mentioned, {TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Figure \ref{fig:TTS-16-survey-area}).
-It is important to note that travel times within this data set are calculated assuming only car travel and one departure time for all origins.
-These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies.
-However, it is not possible from the data retrieval system to obtain higher order tabulations so <!--  we do not know which trips are made with non-car modes nor exact departure times in these tables.
-Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity,-->  we carry on with the assume that all trips are taken by one-time departure car trip.
+Also included in {TTS2016R} are the estimated travel times between OD as summarized in descriptive statistics table in Figure \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the package {r5r}.
+{r5r} interfaces with the java-based {r5} routing engine developed separately by [Conveyal](https://conveyal.com/).
+The inputs to {r5r} for this data package were the desired mode, a maximum travel time threshold of 180 minutes, the geo-coded origin destination pairs based on the centroids of the TAZ, and the static Open Street Map road network of Ontario.
+A travel time threshold of 180 minutes was selected since it captures almost all potential OD interactions.
 
+Additionally, the car mode was included since it is a critically important commute mode in the GGH.
+`r sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) %>% prettyNum(big.mark = ",")` of the trips are made using a car mode out of the total `r sum(od$sum_trips) %>% prettyNum(big.mark = ",")` work-related trips according to the TTS 2016 data  (i.e., `r (sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) / sum(od$sum_trips)) %>% percent()` of trips are taken by car).
+
+These travel times are useful addition to {TTS2016R} since they are not included in the TTS Data Retrieval System but they are vitally important to estimate the cost of travel and associated impedance functions, among other possible applications.
+If the readership is interested in additional information regarding the travel time computation, please see the calculation notebook in the documentation of {TTS2016R} and details about {r5r} at the [r5r package website](https://ipeagit.github.io/r5r/index.html).
 ```{r data-for-impedance}
 # remove all NA trips from dataset and set all 0min travel times to 0.1 min
-od_ft_tt  <- TTS2016R::od_ft_tt %>% 
+od  <- TTS2016R::od %>% 
   filter( !is.na(travel_time)) %>% 
@@
 
-all_tt <- od_ft_tt  %>% 
-  dplyr::select(trips, travel_time)
+all_tt <- od  %>% 
+  dplyr::select(Persons, travel_time)
 
-all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
+all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$Persons), 2]
 ```
@@
 # transfer calibrated impedance function values to OD matrix
-od_ft_tt <- od_ft_tt %>%
+od <- od %>%
   mutate(f = dgamma(travel_time, gamma_$estimate["shape"], gamma_$estimate["rate"]))
 
-#add the number of jobs and workers to the od_ft_tt matrix
-od_ft <- od_ft_tt %>% merge(ggh_taz %>% dplyr::select(GTA06, workers) %>% st_drop_geometry(),
+#add the number of jobs and workers to the od matrix
+od_ft <- od %>% merge(ggh_taz %>% dplyr::select(GTA06, workers) %>% st_drop_geometry(),
                    by.x = "Origin", by.y="GTA06", all.x = TRUE)
@@
 
-```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.", fig.width=8, fig.height=5, message=FALSE}
-
+```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
+Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed.
+", fig.width=8, fig.height=5, message=FALSE}
 tts_total_tt <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "GGH_sum_tt_i", title = "Total travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "GGH_sum_tt_i", title = "Total travel time\n per TAZ (min)", palette = "-Spectral", alpha=0.85, style = "jenks") +
+  tm_shape(ggh_pd %>% filter(REGION_name == "Niagara")) +
+  tm_borders(col="Purple")+
+  tm_shape(ggh_pd %>% filter(REGION_name == "Waterloo")) +
+  tm_borders(col="Brown")+
+  tm_shape(ggh_cma %>% filter(CMANAME == "Toronto")) +
+  tm_borders(col="Blue")+
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 tts_tt_per_trip <- tm_shape(GGH_taz_acc) +
-  tm_fill(col = "GGH_tt_trips_i", title = "Avg. travel time\n per TAZ (min)", palette = "-Spectral", style = "jenks") +
+  tm_fill(col = "GGH_tt_trips_i", title = "Avg. travel time\n per TAZ (min)", palette = "-Spectral", alpha = 0.85, style = "jenks") +
+  tm_shape(ggh_pd %>% filter(REGION_name == "Niagara")) +
+  tm_borders(col="Purple")+
+  tm_shape(ggh_pd %>% filter(REGION_name == "Waterloo")) +
+  tm_borders(col="Brown")+
+  tm_shape(ggh_cma %>% filter(CMANAME == "Toronto")) +
+  tm_borders(col="Blue")+
   tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
@@
 As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the average travel time is distinct from other plots presented so far.
-We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA)<!-- (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) -->, Niagara and Waterloo, the average travel times are moderately low.
+For instance, we can see that in areas around the south-eastern border such as Niagara and Waterloo (purple and brown borders), the average travel times are moderately low.
+Additionally, travel times (by car) within the core of the Toronto census metropolitan area (CMA) (blue) is also moderately since traffic congestion is not reflected in the travel time estimations.
 Further from these areas, travel times are higher.
-Interestingly, even in eastern areas (e.g., Peterborough) with high employment and high job concentration, average travel time is higher than within the GTHA.
 
@@
 # remove all NA trips from dataset and set all 0min travel times to 0.1min
-od_ft_tt  <- od_ft_tt %>% filter( !is.na(travel_time)) %>% mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))
-all_tt <- od_ft_tt  %>% dplyr::select(trips, travel_time)
-all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$trips), 2]
+od  <- od %>% filter( !is.na(travel_time)) %>% mutate(travel_time = ifelse(travel_time == 0, 0.1, travel_time))
+all_tt <- od  %>% dplyr::select(Persons, travel_time)
+all_tt <- all_tt[rep(seq_len(dim(all_tt)[1]), all_tt$Persons), 2]
 ```
@@
 
-The open data product introduced in this paper shares tables for home-to-work related data from the 2016 TTS.
-In addition, inter-centroid travel time tables are calculated, and the planning/municipality boundaries are included as a compliment.
-This open data product, {TTS2016R}, is freely available to explore in an `R` environment.
+{TTS2016R}, the open data package introduced in this paper fuses multiple sources of data.
+It includes an OD cross-tabulation by person and by trip mode table for home-to-work commute data from the 2016 TTS alongside complimentary boundaries and estimated travel times.
+The value of this data package is in its transparency, easy of access, and its open infrastructure for the addition of complimentary data sets in the future.
+Using `R` users can immediately and easily explore GGH commute flow trends as well as suggestion further amendments to the package by request.
 One possible use of this data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.tex
submitted this version to EPB
draft of revised manuscript

@@
 accessible information that adds value to the original inputs
-\citep[see][]{Arribas2021open}. The product presented in this paper is
-an \texttt{R} data package which currently consists of three objects
-which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or
-are curated to facilitate the use and analysis of TTS data. This package
-includes person-to-jobs origin-destinations, traffic analysis zone (TAZ)
-boundaries, and planning/municipality boundaries for the Greater Golden
-Horse area (GGH) located in southern Ontario, Canada
-\citep{data_management_group_tts_2018}. In addition, the package
-includes TAZ centroid-to-centroid travel times by car computed using
-package \{r5r\} \citep{Pereira2021r5r}. The aim of this paper is to walk
-readers through the empirical home-based work commute data set,
-illustrate the calculation of an impedance function that can be used to
-calculate accessibility to employment, and invite its use in other
-applications.
+\citep[see][]{Arribas2021open}. The product presented in this paper is a
+\texttt{R} data package which currently consists of a fusion of objects
+from a variety of sources: home-to-work flows and trips by mode are
+sourced from the 2016 Transportation Tomorrow Survey (TTS)
+\citep{data_management_group_tts_2018}, estimated travel times are
+calculated, and boundary files from the TTS and from the Canadian
+Census.
+
+What is a \texttt{R} data package? A \texttt{R} data package contains
+code, data, and documentation in a standardised collection format that
+can be installed by \texttt{R} users through a centralized software
+repository such as CRAN (the Comprehensive R Archive Network) and
+GitHub. \{TTS2016R\} is freely available on GitHub for all to install
+and freely use in the spirit of open and reproducible research.
+Currently, \{TTS2016R\} includes full-time home-based work-to-job origin
+destinations counts and mode-specific trip numbers retrieved from the
+2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality,
+planning, and census metropolitan area boundaries for the Greater Golden
+Horse area (GGH) located in southern Ontario, Canada. In addition, the
+package includes TAZ centroid-to-centroid travel times by car, transit,
+cycling, and walking mode computed using package \{r5r\}
+\citep{Pereira2021r5r}.
+
+The aim of this paper is to walk readers through the data sets,
+illustrate a use case (i.e., the calculation of an impedance function
+that can be used to calculate accessibility to employment), and invite
+others to experiment in its uses and applications. Though data from the
+TTS is freely available to the public through the
+\href{https://dmg.utoronto.ca/idrs/index}{TTS Data Retrieval System},
+the raw data can be technically demanding, cumbersome to work with, and
+requires multiple software to process. By pre-processing the data in a
+\texttt{R} environment and packaging it alongside complimentary data,
+\{TTS2016R\} offers a slice of the TTS data which can be immediately
+used by \texttt{R} users to analysis patterns of commuting to work in
+the region. Anticipate this package to grow in the future: it currently
+provides an open infrastructure for additional TTS or complimentary data
+sets to be amended by the authors and the open-source community in the
+future by request.
 
@@
 
-\{TTS2016R\} includes counts of fully-employed population by place of
-residence (origin) and counts of full-time jobs by place of work
-(destination) aggregated by TAZ (n=3,764 within the survey boundaries).
-TAZ typically are defined based on land-use and population demographics
-in order to estimate the number of trips produced and attracted to each
-zone \citep{meyer_urban_2001}. As such, each TAZ is uniquely identified
-using the GTA06 Zoning System which can be used to join to the
-origin-destination table (i.e., trips taken).
+Currently, \{TTS2016R\} includes counts of full-time employed population
+by place of residence (origin), counts of full-time usual place of work
+(destination), number of trips to work by mode, and the calculated
+potential travel time of the trips in the GGH. The GGH (and hence the
+TTS survey area) is displayed in Figure \ref{fig:TTS-16-survey-area}.
 
-The number of jobs (3,081,885) and workers (3,446,957) in this package
-are organized in the form of an origin-destination table which is
-indicative of home-to-work commute patterns (there are 3,446,957
-potential interactions). These data were retrieved from the TTS Data
-Retrieval System on October 28, 2021 and reflect the potential
-interaction of full-time employed people and jobs within the GGH survey
-boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by
-the 2016 TTS methodology \citep{data_management_group_tts_2018}.
+This data is aggregated and available at the level of TAZ: TAZ are a
+spatial unit of analysis typically used to estimate the number of trips
+produced and attracted to each zone \citep{meyer_urban_2001}. They are
+thus defined by transportation planners for a region based on
+intra-similarity and inter-dissimilarity between land-use and population
+demographics. Within the GGH boundaries, 3,764 TAZ are specified and
+each TAZ is uniquely identified using the GTA06 Zoning System: the
+survey boundary is discussed in the 2016 TTS methodology and defined by
+the TTS \citep{data_management_group_tts_2018}. The TAZ range between
+\(\ge\)\texttt{0.019} km\^{}2 in spatial area to a maximum of 879
+km\^{}2 (median: 1.3 km\^{}2 and 3rd quantile: 2.8 km\^{}2).
 
-Also included in \{TTS2016R\} are travel times and cost of travel from
-origin to destination by car; travel times are calculated using the
-\texttt{R} package \{r5r\}. These travel times are useful to estimate
-the cost of travel and to calculate impedance functions, among other
-possible applications. For simplicity, all interactions within
-\{TTS2016R\} are assumed to be taken by car, and the travel time is
-calculated from an origin TAZ centroid to a destination TAZ centroid.
-The centroid is snapped to the nearest street line by \texttt{r5r} and
-the travel time is calculated for all trips assuming a car travel mode.
-Additionally, only travel times less than or equal to 180 mins (3 hrs)
-are calculated; this threshold represents 99\% of trip's travel times
-which are summarized in Figure \ref{fig:TTS-16-survey-area}.
-
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ.
-Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
+\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
+3,507 trips were not assigned travel times as they are longer than 180 mins.
+Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
 \end{figure}
 
-\hypertarget{employed-individuals-and-jobs}{%
-\subsection{Employed individuals and
-jobs}\label{employed-individuals-and-jobs}}
+\hypertarget{full-time-employed-people-and-associated-places-of-employment}{%
+\subsection{Full-time employed people and associated places of
+employment}\label{full-time-employed-people-and-associated-places-of-employment}}
 
-The origin-destination table (i.e., trips) consists of a
-cross-tabulation of people who are employed full-time by place of GGH
-residence (origin) and places of GGH employment (destination) using the
-GTA06 TAZ zoning system. It is important to note that the total number
-of workers and jobs is the TTS 2016 region are not equal but the number
-of trips taken are equal to the number of workers. Since the outer
-boundaries of the TTS are permeable, workers who reside within the
-boundaries but travel outside of the boundaries are counted as workers
+In the GGH, there are 3,446,957 workers, 3,081,900 jobs, and 3,282,611
+work-related trips (for the 2016 TTS survey day). The values are
+organized within the origin destination (OD) table in the \{TTS2016R\}
+package and are derived from the cross-tabulation by person and by trip
+for the full-time employed population and associated places of
+employment. The TTS is a proportionally representative survey, hence the
+values included in \{TTS2016R\} are adjusted to reflect the GGH
+population.
+
+It is important to note that the total number of full-time workers and
+jobs in the TTS 2016 region are not equal. Since the outer boundaries of
+the TTS are permeable, workers who reside within the boundaries but have
+workplaces that are outside of the boundaries are counted as workers
 within an origin TAZ, while jobs in TAZ that are filled by workers who
@@
 1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
-3,081,885 jobs). As such, the origin-destination table contained in
-\{TTS2016R\} offers a perspective on all workers in the GGH and their
-home-based trips to places of GGH employment.
+3,081,900 jobs). As such, the OD table contained in \{TTS2016R\} offers
+a perspective on all workers in the GGH and their home-based trips to
+places of GGH employment.
+
+The count of links and trips made by the full-time working population
+and associated full-time place of employment per unique OD pair are
+quite variable. TAZ contain between 0 to 8,491 workers (median: 464, 3rd
+quantile: 1,378), 0 to 41,821 jobs (median: 244, 3rd quantile: 700), and
+generate between 0 to 241 trips (median: 15, 3rd quantile: 42).
 
@@
 
-Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers
-and jobs per TAZ. It can be observed that the spatial distribution of
-jobs and workers is unequal, which is indicative of a jobs-housing
-imbalance that can impact accessibility in a region
+Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
+people and associated jobs per TAZ. It can be observed that the spatial
+distribution of jobs and workers is unequal, which is indicative of a
+jobs -housing imbalance that can impact accessibility in a region
 \citep{Levine1998rethinking}. It can also be seen that there is a higher
@@
 
-As mentioned, \{TTS2016R\} also includes travel time data for each
-home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
-This travel time corresponds to a car commute calculated using the R
-package \{r5r\} (see descriptive statistics in Figure
-\ref{fig:TTS-16-survey-area}). It is important to note that travel times
-within this data set are calculated assuming only car travel and one
-departure time for all origins. These assumptions are not completely
-realistic since we know a small proportion of trips are taken by non-car
-modes and travel time departures varies. However, it is not possible
-from the data retrieval system to obtain higher order tabulations so we
-carry on with the assume that all trips are taken by one-time departure
-car trip.
+Also included in \{TTS2016R\} are the estimated travel times between OD
+as summarized in descriptive statistics table in Figure
+\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
+package \{r5r\}. \{r5r\} interfaces with the java-based \{r5\} routing
+engine developed separately by \href{https://conveyal.com/}{Conveyal}.
+The inputs to \{r5r\} for this data package were the desired mode, a
+maximum travel time threshold of 180 minutes, the geo-coded origin
+destination pairs based on the centroids of the TAZ, and the static Open
+Street Map road network of Ontario. A travel time threshold of 180
+minutes was selected since it captures almost all potential OD
+interactions.
+
+Additionally, the car mode was included since it is a critically
+important commute mode in the GGH. 2,598,379 of the trips are made using
+a car mode out of the total 3,282,611 work-related trips according to
+the TTS 2016 data (i.e., 79\% of trips are taken by car).
+
+These travel times are useful addition to \{TTS2016R\} since they are
+not included in the TTS Data Retrieval System but they are vitally
+important to estimate the cost of travel and associated impedance
+functions, among other possible applications. If the readership is
+interested in additional information regarding the travel time
+computation, please see the calculation notebook in the documentation of
+\{TTS2016R\} and details about \{r5r\} at the
+\href{https://ipeagit.github.io/r5r/index.html}{r5r package website}.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.}\label{fig:plot-tt-ttpertrip}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
+Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed.
+}\label{fig:plot-tt-ttpertrip}
 \end{figure}
-
 \newpage
@@
 \ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
-average travel time is distinct from other plots presented so far. We
-can see that in areas around the south-eastern border that make up the
-Greater Toronto and Hamilton Area (GTHA), Niagara and Waterloo, the
-average travel times are moderately low. Further from these areas,
-travel times are higher. Interestingly, even in eastern areas (e.g.,
-Peterborough) with high employment and high job concentration, average
-travel time is higher than within the GTHA.
+average travel time is distinct from other plots presented so far. For
+instance, we can see that in areas around the south-eastern border such
+as Niagara and Waterloo (purple and brown borders), the average travel
+times are moderately low. Additionally, travel times (by car) within the
+core of the Toronto census metropolitan area (CMA) (blue) is also
+moderately since traffic congestion is not reflected in the travel time
+estimations. Further from these areas, travel times are higher.
 
@@
 \end{array}
-\end{equation}
+\end{equation}\textbackslash end\{equation\}
 
@@
 
-The open data product introduced in this paper shares tables for
-home-to-work related data from the 2016 TTS. In addition, inter-centroid
-travel time tables are calculated, and the planning/municipality
-boundaries are included as a compliment. This open data product,
-\{TTS2016R\}, is freely available to explore in an \texttt{R}
-environment. One possible use of this data, as showcased in this paper,
-is the calibration of impedance functions which in turn can be used for
-accessibility analysis.
+\{TTS2016R\}, the open data package introduced in this paper fuses
+multiple sources of data. It includes an OD cross-tabulation by person
+and by trip mode table for home-to-work commute data from the 2016 TTS
+alongside complimentary boundaries and estimated travel times. The value
+of this data package is in its transparency, easy of access, and its
+open infrastructure for the addition of complimentary data sets in the
+future. Using \texttt{R} users can immediately and easily explore GGH
+commute flow trends as well as suggestion further amendments to the
+package by request. One possible use of this data, as showcased in this
+paper, is the calibration of impedance functions which in turn can be
+used for accessibility analysis.
 

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
Edits to draft manuscript - more to be done.

@@
 abstract: 
-  "This paper describes and visualises the data contained within the {TTS2016R} data-package created in `R`, the statistical computing and graphics language.
-In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS).
-Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, origin-destination trips, and calculated car travel time from TAZ origin-destination centroid pairs.
+  "This paper describes and visualises the data contained within the {TTS2016R} data package created in `R`, the statistical computing and graphics language.
+In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe (GGH) area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS).
+Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, the count of origin destination (OD) pairs and trips by mode per origin TAZ, calculated car travel time from TAZ OD centroid pairs, and associated spatial boundaries to link TAZ to the Canadian Census.
 To illustrate how this information can be analysed to understand patterns in commuting, we estimate a distance-decay curve (i.e., impedance function) for the region.
-{TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manipulation, and the final data products are detailed."
+The value in {TTS2016R} is that it is a growing open data product built on `R` infrastructure that allows for the immediate access of home-to-work commuting data alongside complimentary objects from different sources.
+The package will continue expanding with additions by the authors and the community at-large by requests in the future.
+{TTS2016R} can be freely explored and downloaded in the associated [Github repository](https://github.com/soukhova/TTS2016R) where the documentation and code involved in data creation (including this manuscript), manipulation, and all open data products are detailed."
 
-keywords: Jobs; population; travel time; impedance; Greater Toronto and Hamilton Area; Ontario, Canada; R 
+keywords: Jobs; population; work; commute; travel time; impedance; Greater Toronto and Hamilton Area; Ontario, Canada; R 
 classoption:
@@
 Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open].
-The product presented in this paper is a `R` data package which currently consists of a fusion of objects from a variety of sources: home-to-work flows and trips by mode are sourced from the 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018], estimated travel times are calculated, and boundary files from the TTS and from the Canadian Census.
+The product presented in this paper is a `R` data package which currently consists of a fusion of objects from a variety of sources: home-to-work flows sourced from the 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018], estimated travel times (calculated using {r5r} [@Pereira2021r5r]), and boundary files from the TTS [@datamanagementgroupSurveyBoundaryFiles2018] and from the Canadian Census [@statisticscanadaBoundaryFiles20162017].
 
@@
 {TTS2016R} is freely available on GitHub for all to install and freely use in the spirit of open and reproducible research.
-Currently, {TTS2016R} includes full-time home-based work-to-job origin destinations counts and mode-specific trip numbers retrieved from the 2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality, planning, and census metropolitan area boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada.
+Currently and in more detail, {TTS2016R} includes full-time home-based work-to-job origin destinations (OD) counts and mode-specific trip numbers retrieved from the 2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality, planning, and census metropolitan area boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada.
 In addition, the package includes TAZ centroid-to-centroid travel times by car, transit, cycling, and walking mode computed using package {r5r} [@Pereira2021r5r].
@@
 Though data from the TTS is freely available to the public through the [TTS Data Retrieval System](https://dmg.utoronto.ca/idrs/index), the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process.
-By pre-processing the data in a `R` environment and packaging it alongside complimentary data, {TTS2016R} offers a slice of the TTS data which can be immediately used by `R` users to analysis patterns of commuting to work in the region.
+By pre-processing the data, packaging it with complimentary data, and providing explicit documentation in a `R` environment, {TTS2016R} offers a slice of the TTS data that can be immediately used by `R` users to analysis patterns of commuting to work in the region.
 Anticipate this package to grow in the future: it currently provides an open infrastructure for additional TTS or complimentary data sets to be amended by the authors and the open-source community in the future by request.
@@
 Within the GGH boundaries, `r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` TAZ are specified and each TAZ is uniquely identified using the GTA06 Zoning System: the survey boundary is discussed in the 2016 TTS methodology and defined by the TTS [@data_management_group_tts_2018].
-The TAZ range between $\ge$`0.019` km^2 in spatial area to a maximum of `r max(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 (median: `r median(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2 and 3rd quantile: `r quantile(TTS2016R::ggh_taz$AREA, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")` km^2).
+The TAZ range between $\ge$ 0.019 km^2^ in spatial area to a maximum of `r max(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2^ (median: `r median(TTS2016R::ggh_taz$AREA) %>% prettyNum(digits =2, big.mark = ",")` km^2^ and 3rd quantile: `r quantile(TTS2016R::ggh_taz$AREA, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")` km^2^).
 
@@
                                    NA))
-
-Trips <- data.frame("Trips" = c(summary(od$sum_trips)[[1]] %>% round(), 
-                                   summary(od$sum_trips)[[2]] %>% round(),  
-                                   summary(od$sum_trips)[[3]] %>% round(), 
-                                   summary(od$sum_trips)[[4]] %>% round(), 
-                                   summary(od$sum_trips)[[5]] %>% round(),
-                                   summary(od$sum_trips)[[6]]%>% round(),
-                                   NA))
+# 
+# Trips <- data.frame("Trips" = c(summary(od$sum_trips)[[1]] %>% round(), 
+#                                    summary(od$sum_trips)[[2]] %>% round(),  
+#                                    summary(od$sum_trips)[[3]] %>% round(), 
+#                                    summary(od$sum_trips)[[4]] %>% round(), 
+#                                    summary(od$sum_trips)[[5]] %>% round(),
+#                                    summary(od$sum_trips)[[6]]%>% round(),
+#                                    NA))
 
@@
 
-desc_stats <- cbind(Statistics, Workers, Jobs, Links, Trips, Travel_time)
+desc_stats <- cbind(Statistics, Workers, Jobs, Links, Travel_time)
 
@@
 ```
-```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
-3,507 trips were not assigned travel times as they are longer than 180 mins.
-Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.", out.width="100%", fig.align = 'center', fig.pos='H', message=FALSE, warning=FALSE}
+```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
+3,507 trips were not assigned TT as they are longer than 180 mins.
+Spatial boundary files are retrieved from the TTS which define the survey area (Data Management Group, 2018a): the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.", out.width="100%", fig.align = 'center', fig.pos='H', message=FALSE, warning=FALSE}
 TTS_area_plot <- TTS_area_plot + annotate(geom = "table",
@@
 The count of links and trips made by the full-time working population and associated full-time place of employment per unique OD pair are quite variable.
-TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), and generate between `r 0` to `r 241` trips (median: `r 15`, 3rd quantile: `r 42`).
+TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), and generate between 0 to 241 trips (median: 15, 3rd quantile: 42).
 ```{r, include=FALSE, eval=FALSE, echo=FALSE}
@@
 
-```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.", fig.width=8, fig.height=5}
+```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ retrieved from the 2016 TTS (Data Management Group, 2018b).
+Spatial boundary files are retrieved from the TAZ defined by the TTS (Data Management Group, 2018a).", fig.width=8, fig.height=5}
 
@@
 It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs -housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
-It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
+It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs.
+The number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
 <!-- see the following cumulative probability distribution in Figure \ref{fig:ECD-plot}.
@@
 Also included in {TTS2016R} are the estimated travel times between OD as summarized in descriptive statistics table in Figure \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the package {r5r}.
-{r5r} interfaces with the java-based {r5} routing engine developed separately by [Conveyal](https://conveyal.com/).
-The inputs to {r5r} for this data package were the desired mode, a maximum travel time threshold of 180 minutes, the geo-coded origin destination pairs based on the centroids of the TAZ, and the static Open Street Map road network of Ontario.
+{r5r} interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+The inputs to {r5r} for this data package were the desired mode, a maximum travel time threshold of 180 minutes, the geo-coded origin destination pairs based on the centroids of the TAZ, and the static Open Street Map road network of Ontario (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022]).
 A travel time threshold of 180 minutes was selected since it captures almost all potential OD interactions.
@@
 ```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
-Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed.
+Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively.
 ", fig.width=8, fig.height=5, message=FALSE}

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.tex
Edits to draft manuscript - more to be done.

@@
 This paper describes and visualises the data contained within the
-\{TTS2016R\} data-package created in \texttt{R}, the statistical
+\{TTS2016R\} data package created in \texttt{R}, the statistical
 computing and graphics language. In addition to a synthetic example,
 \{TTS2016R\} contains home-to-work commute information for the Greater
-Golden Horseshoe area in Canada retrieved from the 2016 Transportation
-Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ),
-the number of people who are employed full-time per TAZ, the number of
-jobs per TAZ, origin-destination trips, and calculated car travel time
-from TAZ origin-destination centroid pairs. To illustrate how this
-information can be analysed to understand patterns in commuting, we
-estimate a distance-decay curve (i.e., impedance function) for the
-region. \{TTS2016R\} can be freely downloaded and explored at:
-\url{https://github.com/soukhova/TTS2016R} where the documentation and
-code involved in data creation, manipulation, and the final data
-products are detailed.
+Golden Horseshoe (GGH) area in Canada retrieved from the 2016
+Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis
+Zones (TAZ), the number of people who are employed full-time per TAZ,
+the number of jobs per TAZ, the count of origin destination (OD) pairs
+and trips by mode per origin TAZ, calculated car travel time from TAZ OD
+centroid pairs, and associated spatial boundaries to link TAZ to the
+Canadian Census. To illustrate how this information can be analysed to
+understand patterns in commuting, we estimate a distance-decay curve
+(i.e., impedance function) for the region. The value in \{TTS2016R\} is
+that it is a growing open data product built on \texttt{R}
+infrastructure that allows for the immediate access of home-to-work
+commuting data alongside complimentary objects from different sources.
+The package will continue expanding with additions by the authors and
+the community at-large by requests in the future. \{TTS2016R\} can be
+freely explored and downloaded in the associated
+\href{https://github.com/soukhova/TTS2016R}{Github repository} where the
+documentation and code involved in data creation (including this
+manuscript), manipulation, and all open data products are detailed.
 \end{abstract}
 
-\keywords{Jobs; population; travel time; impedance; Greater Toronto and
-Hamilton Area; Ontario, Canada; R}
+\keywords{Jobs; population; work; commute; travel time; impedance;
+Greater Toronto and Hamilton Area; Ontario, Canada; R}
 
@@
 \texttt{R} data package which currently consists of a fusion of objects
-from a variety of sources: home-to-work flows and trips by mode are
-sourced from the 2016 Transportation Tomorrow Survey (TTS)
-\citep{data_management_group_tts_2018}, estimated travel times are
-calculated, and boundary files from the TTS and from the Canadian
-Census.
+from a variety of sources: home-to-work flows sourced from the 2016
+Transportation Tomorrow Survey (TTS)
+\citep{data_management_group_tts_2018}, estimated travel times
+(calculated using \{r5r\} \citep{Pereira2021r5r}), and boundary files
+from the TTS \citep{datamanagementgroupSurveyBoundaryFiles2018} and from
+the Canadian Census \citep{statisticscanadaBoundaryFiles20162017}.
 
@@
 and freely use in the spirit of open and reproducible research.
-Currently, \{TTS2016R\} includes full-time home-based work-to-job origin
-destinations counts and mode-specific trip numbers retrieved from the
-2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality,
-planning, and census metropolitan area boundaries for the Greater Golden
-Horse area (GGH) located in southern Ontario, Canada. In addition, the
-package includes TAZ centroid-to-centroid travel times by car, transit,
-cycling, and walking mode computed using package \{r5r\}
-\citep{Pereira2021r5r}.
+Currently and in more detail, \{TTS2016R\} includes full-time home-based
+work-to-job origin destinations (OD) counts and mode-specific trip
+numbers retrieved from the 2016 TTS, traffic analysis zone (TAZ)
+boundaries, and municipality, planning, and census metropolitan area
+boundaries for the Greater Golden Horse area (GGH) located in southern
+Ontario, Canada. In addition, the package includes TAZ
+centroid-to-centroid travel times by car, transit, cycling, and walking
+mode computed using package \{r5r\} \citep{Pereira2021r5r}.
 
@@
 the raw data can be technically demanding, cumbersome to work with, and
-requires multiple software to process. By pre-processing the data in a
-\texttt{R} environment and packaging it alongside complimentary data,
-\{TTS2016R\} offers a slice of the TTS data which can be immediately
-used by \texttt{R} users to analysis patterns of commuting to work in
-the region. Anticipate this package to grow in the future: it currently
-provides an open infrastructure for additional TTS or complimentary data
-sets to be amended by the authors and the open-source community in the
-future by request.
+requires multiple software to process. By pre-processing the data,
+packaging it with complimentary data, and providing explicit
+documentation in a \texttt{R} environment, \{TTS2016R\} offers a slice
+of the TTS data that can be immediately used by \texttt{R} users to
+analysis patterns of commuting to work in the region. Anticipate this
+package to grow in the future: it currently provides an open
+infrastructure for additional TTS or complimentary data sets to be
+amended by the authors and the open-source community in the future by
+request.
 
@@
 the TTS \citep{data_management_group_tts_2018}. The TAZ range between
-\(\ge\)\texttt{0.019} km\^{}2 in spatial area to a maximum of 879
-km\^{}2 (median: 1.3 km\^{}2 and 3rd quantile: 2.8 km\^{}2).
+\(\ge\) 0.019 km\textsuperscript{2} in spatial area to a maximum of 879
+km\textsuperscript{2} (median: 1.3 km\textsuperscript{2} and 3rd
+quantile: 2.8 km\textsuperscript{2}).
 
@@
 
-\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
-3,507 trips were not assigned travel times as they are longer than 180 mins.
-Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
+\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
+3,507 trips were not assigned TT as they are longer than 180 mins.
+Spatial boundary files are retrieved from the TTS which define the survey area (Data Management Group, 2018a): the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ retrieved from the 2016 TTS (Data Management Group, 2018b).
+Spatial boundary files are retrieved from the TAZ defined by the TTS (Data Management Group, 2018a).}\label{fig:tts-workers-jobs-plot}
 \end{figure}
@@
 with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
-is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
-number TAZ with an extreme number of jobs at the highest and lowest
-percentiles is significantly higher than the number of workers.
+is higher than the mean of jobs. The number TAZ with an extreme number
+of jobs at the highest and lowest percentiles is significantly higher
+than the number of workers.
 
@@
 \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
-package \{r5r\}. \{r5r\} interfaces with the java-based \{r5\} routing
-engine developed separately by \href{https://conveyal.com/}{Conveyal}.
-The inputs to \{r5r\} for this data package were the desired mode, a
-maximum travel time threshold of 180 minutes, the geo-coded origin
-destination pairs based on the centroids of the TAZ, and the static Open
-Street Map road network of Ontario. A travel time threshold of 180
-minutes was selected since it captures almost all potential OD
-interactions.
+package \{r5r\}. \{r5r\} interfaces with the java-based R5 routing
+engine developed separately by Conveyal
+\citep{conveyalConveyalR5Routing2022}. The inputs to \{r5r\} for this
+data package were the desired mode, a maximum travel time threshold of
+180 minutes, the geo-coded origin destination pairs based on the
+centroids of the TAZ, and the static Open Street Map road network of
+Ontario (retrieved using Geofabrick
+\citep{geofabrikOntarioOpenStreetMapGeofabrik2022}). A travel time
+threshold of 180 minutes was selected since it captures almost all
+potential OD interactions.
 
@@
 \includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
-Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed.
+Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively.
 }\label{fig:plot-tt-ttpertrip}

### 2022-10-24 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.Rmd
submitted revision to journal. submission package in 'editorial-correspondance'.

@@
 ---
-title: "TTS2016R: A dataset to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area, Canada"
+title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe Area, Ontario, Canada"
 author:
@@
 abstract: 
-  "This paper describes and visualises the data contained within the {TTS2016R} data-package created in `R`, the statistical computing and graphics language.
-In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS).
-Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, origin-destination trips, and calculated car travel time from TAZ origin-destination centroid pairs.
+  "This paper describes and visualises the data contained within the {TTS2016R} data package created in `R`, the statistical computing and graphics language.
+In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe (GGH) area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS).
+Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, the count of origin destination (OD) pairs and trips by mode per origin TAZ, calculated car travel time from TAZ OD centroid pairs, and associated spatial boundaries to link TAZ to the Canadian Census.
 To illustrate how this information can be analysed to understand patterns in commuting, we estimate a distance-decay curve (i.e., impedance function) for the region.
-{TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manipulation, and the final data products are detailed."
+{TTS2016R} is a growing open data product built on `R` infrastructure that allows for the immediate access of home-to-work commuting data alongside complimentary objects from different sources.
+The package will continue expanding with additions by the authors and the community at-large by requests in the future.
+{TTS2016R} can be freely explored and downloaded in the associated [Github repository](https://github.com/soukhova/TTS2016R) where the documentation and code involved in data creation, manipulation, and all open data products are detailed."
 
-keywords: Jobs; population; travel time; impedance; Greater Toronto and Hamilton Area; Ontario, Canada; R 
+keywords: Jobs; population; work; commute; travel time; impedance; Greater Toronto and Hamilton Area; Greater Golden Horshoe Area, Ontario, Canada; R 
 classoption:

### 2022-10-24 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.tex
submitted this version to EPB
submitted revision to journal. submission package in 'editorial-correspondance'.

@@
 
-\title{TTS2016R: A dataset to study population and employment patterns
-from the 2016 Transportation Tomorrow Survey (TTS) in the Greater
-Toronto and Hamilton Area, Canada}
+\title{TTS2016R: A data set to study population and employment patterns
+from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden
+Horseshoe Area, Ontario, Canada}
 
@@
 This paper describes and visualises the data contained within the
-\{TTS2016R\} data-package created in \texttt{R}, the statistical
+\{TTS2016R\} data package created in \texttt{R}, the statistical
 computing and graphics language. In addition to a synthetic example,
 \{TTS2016R\} contains home-to-work commute information for the Greater
-Golden Horseshoe area in Canada retrieved from the 2016 Transportation
-Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ),
-the number of people who are employed full-time per TAZ, the number of
-jobs per TAZ, origin-destination trips, and calculated car travel time
-from TAZ origin-destination centroid pairs. To illustrate how this
-information can be analysed to understand patterns in commuting, we
-estimate a distance-decay curve (i.e., impedance function) for the
-region. \{TTS2016R\} can be freely downloaded and explored at:
-\url{https://github.com/soukhova/TTS2016R} where the documentation and
-code involved in data creation, manipulation, and the final data
-products are detailed.
+Golden Horseshoe (GGH) area in Canada retrieved from the 2016
+Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis
+Zones (TAZ), the number of people who are employed full-time per TAZ,
+the number of jobs per TAZ, the count of origin destination (OD) pairs
+and trips by mode per origin TAZ, calculated car travel time from TAZ OD
+centroid pairs, and associated spatial boundaries to link TAZ to the
+Canadian Census. To illustrate how this information can be analysed to
+understand patterns in commuting, we estimate a distance-decay curve
+(i.e., impedance function) for the region. \{TTS2016R\} is a growing
+open data product built on \texttt{R} infrastructure that allows for the
+immediate access of home-to-work commuting data alongside complimentary
+objects from different sources. The package will continue expanding with
+additions by the authors and the community at-large by requests in the
+future. \{TTS2016R\} can be freely explored and downloaded in the
+associated \href{https://github.com/soukhova/TTS2016R}{Github
+repository} where the documentation and code involved in data creation,
+manipulation, and all open data products are detailed.
 \end{abstract}
 
-\keywords{Jobs; population; travel time; impedance; Greater Toronto and
-Hamilton Area; Ontario, Canada; R}
+\keywords{Jobs; population; work; commute; travel time; impedance;
+Greater Toronto and Hamilton Area; Greater Golden Horshoe Area, Ontario,
+Canada; R}
 
@@
 
-\bibliographystyle{sageh}
-\bibliography{bibfile}
-
 \end{document}

### 2022-10-24 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
submitted revision to journal. submission package in 'editorial-correspondance'.

@@
 ---
-title: "TTS2016R: A dataset to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area, Canada"
+title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe Area, Ontario, Canada"
 author:
-  - name: "Anastasia Soukhov"
-    email: "soukhoa@mcmaster.ca"
-    affiliation: SEES
-  - name: "Antonio Páez"
-    email: "paezha@mcmaster.ca"
-    affiliation: SEES
+  - name: "Anon1"
+    email: "Anon1@anon.org"
+    affiliation: CODE
+  - name: "Anon2"
+    email: "Anon2@anon.org"
+    affiliation: CODE
     footnote: 1
 address:
-  - code: SEES
-    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
+  - code: CODE
+    address: "Some school"
 footnote:
@@
 To illustrate how this information can be analysed to understand patterns in commuting, we estimate a distance-decay curve (i.e., impedance function) for the region.
-The value in {TTS2016R} is that it is a growing open data product built on `R` infrastructure that allows for the immediate access of home-to-work commuting data alongside complimentary objects from different sources.
+{TTS2016R} is a growing open data product built on `R` infrastructure that allows for the immediate access of home-to-work commuting data alongside complimentary objects from different sources.
 The package will continue expanding with additions by the authors and the community at-large by requests in the future.
-{TTS2016R} can be freely explored and downloaded in the associated [Github repository](https://github.com/soukhova/TTS2016R) where the documentation and code involved in data creation (including this manuscript), manipulation, and all open data products are detailed."
+{TTS2016R} can be freely explored and downloaded in the associated [Github repository](https://github.com/soukhova/TTS2016R) where the documentation and code involved in data creation, manipulation, and all open data products are detailed."
 
-keywords: Jobs; population; work; commute; travel time; impedance; Greater Toronto and Hamilton Area; Ontario, Canada; R 
+keywords: Jobs; population; work; commute; travel time; impedance; Greater Toronto and Hamilton Area; Greater Golden Horshoe Area, Ontario, Canada; R 
 classoption:
@@
 The values are organized within the origin destination (OD) table in the {TTS2016R} package and are derived from the cross-tabulation by person and by trip for the full-time employed population and associated places of employment.
-The TTS is a proportionally representative survey, hence the values included in {TTS2016R} are adjusted to reflect the GGH population.
 
-It is important to note that the total number of full-time workers and jobs in the TTS 2016 region are not equal.
+Additionally, the total number of full-time workers and jobs in the TTS 2016 region are not equal.
 Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but have workplaces that are outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
-This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs).
-As such, the OD table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.
+This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs.
+The TTS is a proportionally representative survey, hence the values included in {TTS2016R} are adjusted to reflect the GGH working population and their home-based trips to places of GGH employment.
 
@@
 It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs -housing imbalance that can impact accessibility in a region [@Levine1998rethinking].
-It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs.
+Also, there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers and `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs.
 The number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers.
@@
 
-Also included in {TTS2016R} are the estimated travel times between OD as summarized in descriptive statistics table in Figure \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the package {r5r}.
+Also included in {TTS2016R} are the estimated travel times between OD as summarized in the descriptive statistics table in Figure \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the package {r5r}.
 {r5r} interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
-The inputs to {r5r} for this data package were the desired mode, a maximum travel time threshold of 180 minutes, the geo-coded origin destination pairs based on the centroids of the TAZ, and the static Open Street Map road network of Ontario (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022]).
+The inputs to {r5r} for this data package were: the desired mode, a maximum travel time threshold of 180 minutes, the geo-coded origin destination pairs based on the centroids of the TAZ, and the static OpenStreetMap road network of Ontario (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022]).
 A travel time threshold of 180 minutes was selected since it captures almost all potential OD interactions.
 
-Additionally, the car mode was included since it is a critically important commute mode in the GGH.
-`r sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) %>% prettyNum(big.mark = ",")` of the trips are made using a car mode out of the total `r sum(od$sum_trips) %>% prettyNum(big.mark = ",")` work-related trips according to the TTS 2016 data  (i.e., `r (sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) / sum(od$sum_trips)) %>% percent()` of trips are taken by car).
+Additionally, car travel is included in this data package because it is a critically important commute mode in the GGH.
+`r sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) %>% prettyNum(big.mark = ",")` of the trips are made using a car out of the total `r sum(od$sum_trips) %>% prettyNum(big.mark = ",")` work-related trips according to the TTS 2016 data  (i.e., `r (sum(od$C_DPC, od$C_PPC, od$C_RS, od$C_SB, od$C_TA) / sum(od$sum_trips)) %>% percent()` of trips are taken by car).
 
-These travel times are useful addition to {TTS2016R} since they are not included in the TTS Data Retrieval System but they are vitally important to estimate the cost of travel and associated impedance functions, among other possible applications.
+These travel times are a useful addition to {TTS2016R} since they are not included in the [TTS Data Retrieval System](https://dmg.utoronto.ca/idrs/index) but they are vitally important to estimate the cost of travel and associated impedance functions, among other possible applications.
 If the readership is interested in additional information regarding the travel time computation, please see the calculation notebook in the documentation of {TTS2016R} and details about {r5r} at the [r5r package website](https://ipeagit.github.io/r5r/index.html).
@@
 
-```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
+```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Calculated total worker travel time by car (left) and average worker travel time by car (right) for each TAZ in the 2016 TTS.
+Car travel times calculated using {r5r} (Pereira et al., 2021).
 Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively.
@@
 For instance, we can see that in areas around the south-eastern border such as Niagara and Waterloo (purple and brown borders), the average travel times are moderately low.
-Additionally, travel times (by car) within the core of the Toronto census metropolitan area (CMA) (blue) is also moderately since traffic congestion is not reflected in the travel time estimations.
+Additionally, travel times (by car) within the core of the Toronto census metropolitan area (CMA) (blue) is also moderately low since traffic congestion is not reflected in the travel time calculations.
 Further from these areas, travel times are higher.
@@
 
+An application of the {TTS2016R} package is the calculation of an impedance function.
 Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@talen_assessing_1998;@paez_jobs_2013;@barboza_balancing_2021].
@@
 
-\begin{equation}
-\label{gamma-dist}
-\begin{array}{l}\ 
-f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for }	0 \leq x \leq \infty\\
-\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
-\end{array}
-\end{equation}
+\renewcommand{\[}{\begin{equation}}
+\renewcommand{\]}{\end{equation}}
+
+$$
+\begin{aligned}
+f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for }	0 \leq x \leq \infty \label{gamma-dist} \\
+\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
+\end{aligned}
+$$
 
@@
 <!-- # fitdistrplus::descdist(data=all_tt)  -->
-<!-- # ``` -->\newpage
+<!-- # ``` -->
 
@@
 
-{TTS2016R}, the open data package introduced in this paper fuses multiple sources of data.
+This paper introduces {TTS2016R}, an open data product in the form of a `R` data package.
+This package is a fusion of data from multiple sources and we demonstrate the spatial and numeric extent of the data contained within.
 It includes an OD cross-tabulation by person and by trip mode table for home-to-work commute data from the 2016 TTS alongside complimentary boundaries and estimated travel times.
 The value of this data package is in its transparency, easy of access, and its open infrastructure for the addition of complimentary data sets in the future.
-Using `R` users can immediately and easily explore GGH commute flow trends as well as suggestion further amendments to the package by request.
+`R` users can immediately and easily explore GGH commute flow trends as well as suggestion further amendments to the package by request.
 One possible use of this data, as showcased in this paper, is the calibration of impedance functions which in turn can be used for accessibility analysis.

### 2022-10-24 Manuscript-Data-Package/Manuscript-Data-Package.tex
submitted revision to journal. submission package in 'editorial-correspondance'.

@@
 
-\author{Anastasia Soukhov\affilnum{}, Antonio Páez\affilnum{}}
+\author{Anon1\affilnum{}, Anon2\affilnum{}}
 
@@
 understand patterns in commuting, we estimate a distance-decay curve
-(i.e., impedance function) for the region. The value in \{TTS2016R\} is
-that it is a growing open data product built on \texttt{R}
-infrastructure that allows for the immediate access of home-to-work
-commuting data alongside complimentary objects from different sources.
-The package will continue expanding with additions by the authors and
-the community at-large by requests in the future. \{TTS2016R\} can be
-freely explored and downloaded in the associated
-\href{https://github.com/soukhova/TTS2016R}{Github repository} where the
-documentation and code involved in data creation (including this
-manuscript), manipulation, and all open data products are detailed.
+(i.e., impedance function) for the region. \{TTS2016R\} is a growing
+open data product built on \texttt{R} infrastructure that allows for the
+immediate access of home-to-work commuting data alongside complimentary
+objects from different sources. The package will continue expanding with
+additions by the authors and the community at-large by requests in the
+future. \{TTS2016R\} can be freely explored and downloaded in the
+associated \href{https://github.com/soukhova/TTS2016R}{Github
+repository} where the documentation and code involved in data creation,
+manipulation, and all open data products are detailed.
 \end{abstract}
@@
 \keywords{Jobs; population; work; commute; travel time; impedance;
-Greater Toronto and Hamilton Area; Ontario, Canada; R}
+Greater Toronto and Hamilton Area; Greater Golden Horshoe Area, Ontario,
+Canada; R}
 
@@
 for the full-time employed population and associated places of
-employment. The TTS is a proportionally representative survey, hence the
-values included in \{TTS2016R\} are adjusted to reflect the GGH
-population.
+employment.
 
-It is important to note that the total number of full-time workers and
-jobs in the TTS 2016 region are not equal. Since the outer boundaries of
-the TTS are permeable, workers who reside within the boundaries but have
-workplaces that are outside of the boundaries are counted as workers
-within an origin TAZ, while jobs in TAZ that are filled by workers who
-reside outside the GGH boundaries are \emph{unknown} since they were not
+Additionally, the total number of full-time workers and jobs in the TTS
+2016 region are not equal. Since the outer boundaries of the TTS are
+permeable, workers who reside within the boundaries but have workplaces
+that are outside of the boundaries are counted as workers within an
+origin TAZ, while jobs in TAZ that are filled by workers who reside
+outside the GGH boundaries are \emph{unknown} since they were not
 surveyed. This mismatch results in the total number of workers being
-1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
-3,081,900 jobs). As such, the OD table contained in \{TTS2016R\} offers
-a perspective on all workers in the GGH and their home-based trips to
-places of GGH employment.
+1.12 times larger than the number of jobs. The TTS is a proportionally
+representative survey, hence the values included in \{TTS2016R\} are
+adjusted to reflect the GGH working population and their home-based
+trips to places of GGH employment.
 
@@
 jobs -housing imbalance that can impact accessibility in a region
-\citep{Levine1998rethinking}. It can also be seen that there is a higher
-number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
-with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
-is higher than the mean of jobs. The number TAZ with an extreme number
-of jobs at the highest and lowest percentiles is significantly higher
-than the number of workers.
+\citep{Levine1998rethinking}. Also, there is a higher number of TAZ with
+no workers than zones with no jobs (i.e., 791 TAZ with no workers and
+396 TAZ with no jobs) and the mean of workers per TAZ is higher than the
+mean of jobs. The number TAZ with an extreme number of jobs at the
+highest and lowest percentiles is significantly higher than the number
+of workers.
 
@@
 Also included in \{TTS2016R\} are the estimated travel times between OD
-as summarized in descriptive statistics table in Figure
+as summarized in the descriptive statistics table in Figure
 \ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
@@
 \citep{conveyalConveyalR5Routing2022}. The inputs to \{r5r\} for this
-data package were the desired mode, a maximum travel time threshold of
+data package were: the desired mode, a maximum travel time threshold of
 180 minutes, the geo-coded origin destination pairs based on the
-centroids of the TAZ, and the static Open Street Map road network of
+centroids of the TAZ, and the static OpenStreetMap road network of
 Ontario (retrieved using Geofabrick
@@
 
-Additionally, the car mode was included since it is a critically
-important commute mode in the GGH. 2,598,379 of the trips are made using
-a car mode out of the total 3,282,611 work-related trips according to
-the TTS 2016 data (i.e., 79\% of trips are taken by car).
+Additionally, car travel is included in this data package because it is
+a critically important commute mode in the GGH. 2,598,379 of the trips
+are made using a car out of the total 3,282,611 work-related trips
+according to the TTS 2016 data (i.e., 79\% of trips are taken by car).
 
-These travel times are useful addition to \{TTS2016R\} since they are
-not included in the TTS Data Retrieval System but they are vitally
-important to estimate the cost of travel and associated impedance
-functions, among other possible applications. If the readership is
-interested in additional information regarding the travel time
-computation, please see the calculation notebook in the documentation of
-\{TTS2016R\} and details about \{r5r\} at the
-\href{https://ipeagit.github.io/r5r/index.html}{r5r package website}.
+These travel times are a useful addition to \{TTS2016R\} since they are
+not included in the \href{https://dmg.utoronto.ca/idrs/index}{TTS Data
+Retrieval System} but they are vitally important to estimate the cost of
+travel and associated impedance functions, among other possible
+applications. If the readership is interested in additional information
+regarding the travel time computation, please see the calculation
+notebook in the documentation of \{TTS2016R\} and details about \{r5r\}
+at the \href{https://ipeagit.github.io/r5r/index.html}{r5r package
+website}.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.
+\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time by car (left) and average worker travel time by car (right) for each TAZ in the 2016 TTS.
+Car travel times calculated using {r5r} (Pereira et al., 2021).
 Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively.
@@
 core of the Toronto census metropolitan area (CMA) (blue) is also
-moderately since traffic congestion is not reflected in the travel time
-estimations. Further from these areas, travel times are higher.
+moderately low since traffic congestion is not reflected in the travel
+time calculations. Further from these areas, travel times are higher.
 
@@
 
-Impedance functions are useful to understand mobility behaviour and are
-used to estimate gravity models of spatial interaction
-\citep{wilson1971, haynes_gravity_1985} and applied in accessibility
-analysis
+An application of the \{TTS2016R\} package is the calculation of an
+impedance function. Impedance functions are useful to understand
+mobility behaviour and are used to estimate gravity models of spatial
+interaction \citep{wilson1971, haynes_gravity_1985} and applied in
+accessibility analysis
 \citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
@@
 
-\begin{equation}
-\label{gamma-dist}
-\begin{array}{l}\ 
-f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
-\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
-\end{array}
-\end{equation}\textbackslash end\{equation\}
+\renewcommand{\[}{\begin{equation}}
+\renewcommand{\]}{\end{equation}}
 
-\newpage
+\[
+\begin{aligned}
+f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty \label{gamma-dist} \\
+\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
+\end{aligned}
+\]
 
@@
 
-\{TTS2016R\}, the open data package introduced in this paper fuses
-multiple sources of data. It includes an OD cross-tabulation by person
+This paper introduces \{TTS2016R\}, an open data product in the form of
+a \texttt{R} data package. This package is a fusion of data from
+multiple sources and we demonstrate the spatial and numeric extent of
+the data contained within. It includes an OD cross-tabulation by person
 and by trip mode table for home-to-work commute data from the 2016 TTS
@@
 open infrastructure for the addition of complimentary data sets in the
-future. Using \texttt{R} users can immediately and easily explore GGH
-commute flow trends as well as suggestion further amendments to the
-package by request. One possible use of this data, as showcased in this
-paper, is the calibration of impedance functions which in turn can be
-used for accessibility analysis.
+future. \texttt{R} users can immediately and easily explore GGH commute
+flow trends as well as suggestion further amendments to the package by
+request. One possible use of this data, as showcased in this paper, is
+the calibration of impedance functions which in turn can be used for
+accessibility analysis.
 
@@
 \bibliographystyle{sageh}
-\bibliography{bibfile}
+\bibliography{bibfile.bib}
 

### 2022-10-28 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.Rmd
submitted this version after minor journal requirment additions - revsion 1

@@
 author:
-  - name: "Anastasia Soukhov"
-    email: "soukhoa@mcmaster.ca"
-    affiliation: SEES
-  - name: "Antonio Páez"
-    email: "paezha@mcmaster.ca"
-    affiliation: SEES
+  - name: Anastasia Soukhov*
+    num: 1
+  - name: Antonio Páez
+    num: 1
     footnote: 1
 address:
-  - code: SEES
-    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+  - num: 1
+    org: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
+corrauth: "Anastasia Soukhov, soukhoa@mcmaster.ca"
 abstract: 
@@
 ---
+
+**Anastasia Soukhov** is a PhD student in the School of Earth, Environment & Society at McMaster University.
+She has a masters and bachelors in Civil Engineering with a transportation specialization.
+She is enthusiastic about sustainable and equitable housing and transportation and is a researcher within the Mobilizing Justice Partnership working to establish data driven equity standards in Canadian cities.
+Her recent work includes: studies on sustainable passenger vehicle energy consumption and wheel-to-wheel emission simulation, sustainable mobility policy optimization, and competitive accessibility measures.
+
+**Antonio Páez** is full professor in the School of Earth, Environment & Society at McMaster University.
+Recent work includes studies on accessibility, spatial analysis of qualitative variables, aging and mobility, transportation and social exclusion, the influence of the built and social environments on travel behavior, social networks and decision making, telework adoption, and healthcare provision and trends.
+Paez has published widely, and is author or co-author of over 150 papers, many appearing in leading international journals.
+He currently serves as Editor-in-Chief of the Journal of Geographical Systems, and sits on the editorial boards of Transportation, Journal of Transport Geography, Geographical Analysis, and International Journal of Geographical Information Science, among others.

### 2022-10-28 Manuscript-Data-Package/Manuscript-Data-Package-cover-page.tex
submitted this version after minor journal requirment additions - revsion 1

@@
 
-\author{Anastasia Soukhov\affilnum{}, Antonio Páez\affilnum{}}
+\author{Anastasia Soukhov*\affilnum{1}, Antonio Páez\affilnum{1}}
 
-\affiliation{\affilnum{}{}}
+\affiliation{\affilnum{1}{School of Earth, Environment and Society,
+McMaster University, Hamilton, ON, L8S 4K1, Canada}}
+
+\corrauth{Anastasia Soukhov,
+\href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}}
 
@@
 
+\textbf{Anastasia Soukhov} is a PhD student in the School of Earth,
+Environment \& Society at McMaster University. She has a masters and
+bachelors in Civil Engineering with a transportation specialization. She
+is enthusiastic about sustainable and equitable housing and
+transportation and is a researcher within the Mobilizing Justice
+Partnership working to establish data driven equity standards in
+Canadian cities. Her recent work includes: studies on sustainable
+passenger vehicle energy consumption and wheel-to-wheel emission
+simulation, sustainable mobility policy optimization, and competitive
+accessibility measures.
+
+\textbf{Antonio Páez} is full professor in the School of Earth,
+Environment \& Society at McMaster University. Recent work includes
+studies on accessibility, spatial analysis of qualitative variables,
+aging and mobility, transportation and social exclusion, the influence
+of the built and social environments on travel behavior, social networks
+and decision making, telework adoption, and healthcare provision and
+trends. Paez has published widely, and is author or co-author of over
+150 papers, many appearing in leading international journals. He
+currently serves as Editor-in-Chief of the Journal of Geographical
+Systems, and sits on the editorial boards of Transportation, Journal of
+Transport Geography, Geographical Analysis, and International Journal of
+Geographical Information Science, among others.
+
 \end{document}

### 2022-10-28 Manuscript-Data-Package/Manuscript-Data-Package_R0.tex
submitted this version after minor journal requirment additions - revsion 1

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\setcitestyle{aysep={,}}

\title{TTS2016R: A dataset to study population and employment patterns
from the 2016 Transportation Tomorrow Survey (TTS) in the Greater
Toronto and Hamilton Area, Canada}

\runninghead{}

\author{Anastasia Soukhov\affilnum{}, Antonio Páez\affilnum{}}

\affiliation{\affilnum{}{}}

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data-package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe area in Canada retrieved from the 2016 Transportation
Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ),
the number of people who are employed full-time per TAZ, the number of
jobs per TAZ, origin-destination trips, and calculated car travel time
from TAZ origin-destination centroid pairs. To illustrate how this
information can be analysed to understand patterns in commuting, we
estimate a distance-decay curve (i.e., impedance function) for the
region. \{TTS2016R\} can be freely downloaded and explored at:
\url{https://github.com/soukhova/TTS2016R} where the documentation and
code involved in data creation, manipulation, and the final data
products are detailed.
\end{abstract}

\keywords{Jobs; population; travel time; impedance; Greater Toronto and
Hamilton Area; Ontario, Canada; R}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is
an \texttt{R} data package which currently consists of three objects
which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or
are curated to facilitate the use and analysis of TTS data. This package
includes person-to-jobs origin-destinations, traffic analysis zone (TAZ)
boundaries, and planning/municipality boundaries for the Greater Golden
Horse area (GGH) located in southern Ontario, Canada
\citep{data_management_group_tts_2018}. In addition, the package
includes TAZ centroid-to-centroid travel times by car computed using
package \{r5r\} \citep{Pereira2021r5r}. The aim of this paper is to walk
readers through the empirical home-based work commute data set,
illustrate the calculation of an impedance function that can be used to
calculate accessibility to employment, and invite its use in other
applications.

\hypertarget{home-to-work-commute-data}{%
\section{Home-to-work commute data}\label{home-to-work-commute-data}}

\{TTS2016R\} includes counts of fully-employed population by place of
residence (origin) and counts of full-time jobs by place of work
(destination) aggregated by TAZ (n=3,764 within the survey boundaries).
TAZ typically are defined based on land-use and population demographics
in order to estimate the number of trips produced and attracted to each
zone \citep{meyer_urban_2001}. As such, each TAZ is uniquely identified
using the GTA06 Zoning System which can be used to join to the
origin-destination table (i.e., trips taken).

The number of jobs (3,081,885) and workers (3,446,957) in this package
are organized in the form of an origin-destination table which is
indicative of home-to-work commute patterns (there are 3,446,957
potential interactions). These data were retrieved from the TTS Data
Retrieval System on October 28, 2021 and reflect the potential
interaction of full-time employed people and jobs within the GGH survey
boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by
the 2016 TTS methodology \citep{data_management_group_tts_2018}.

Also included in \{TTS2016R\} are travel times and cost of travel from
origin to destination by car; travel times are calculated using the
\texttt{R} package \{r5r\}. These travel times are useful to estimate
the cost of travel and to calculate impedance functions, among other
possible applications. For simplicity, all interactions within
\{TTS2016R\} are assumed to be taken by car, and the travel time is
calculated from an origin TAZ centroid to a destination TAZ centroid.
The centroid is snapped to the nearest street line by \texttt{r5r} and
the travel time is calculated for all trips assuming a car travel mode.
Additionally, only travel times less than or equal to 180 mins (3 hrs)
are calculated; this threshold represents 99\% of trip's travel times
which are summarized in Figure \ref{fig:TTS-16-survey-area}.

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{images/TTS16-survey-area}

}

\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ. Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).}\label{fig:TTS-16-survey-area}
\end{figure}

\hypertarget{employed-individuals-and-jobs}{%
\subsection{Employed individuals and
jobs}\label{employed-individuals-and-jobs}}

The origin-destination table (i.e., trips) consists of a
cross-tabulation of people who are employed full-time by place of GGH
residence (origin) and places of GGH employment (destination) using the
GTA06 TAZ zoning system. It is important to note that the total number
of workers and jobs is the TTS 2016 region are not equal but the number
of trips taken are equal to the number of workers. Since the outer
boundaries of the TTS are permeable, workers who reside within the
boundaries but travel outside of the boundaries are counted as workers
within an origin TAZ, while jobs in TAZ that are filled by workers who
reside outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
3,081,885 jobs). As such, the origin-destination table contained in
\{TTS2016R\} offers a perspective on all workers in the GGH and their
home-based trips to places of GGH employment.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
\end{figure}

Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers
and jobs per TAZ. It can be observed that the spatial distribution of
jobs and workers is unequal, which is indicative of a jobs-housing
imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
number TAZ with an extreme number of jobs at the highest and lowest
percentiles is significantly higher than the number of workers.

\hypertarget{calculated-travel-time}{%
\subsection{Calculated travel time}\label{calculated-travel-time}}

As mentioned, \{TTS2016R\} also includes travel time data for each
home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}.
This travel time corresponds to a car commute calculated using the R
package \{r5r\} (see descriptive statistics in Figure
\ref{fig:TTS-16-survey-area}). It is important to note that travel times
within this data set are calculated assuming only car travel and one
departure time for all origins. These assumptions are not completely
realistic since we know a small proportion of trips are taken by non-car
modes and travel time departures varies. However, it is not possible
from the data retrieval system to obtain higher order tabulations so we
carry on with the assume that all trips are taken by one-time departure
car trip.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS.}\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time resembles the spatial trend distribution in the number of
employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
average travel time is distinct from other plots presented so far. We
can see that in areas around the south-eastern border that make up the
Greater Toronto and Hamilton Area (GTHA), Niagara and Waterloo, the
average travel times are moderately low. Further from these areas,
travel times are higher. Interestingly, even in eastern areas (e.g.,
Peterborough) with high employment and high job concentration, average
travel time is higher than within the GTHA.

\hypertarget{calibrating-an-impedance-function}{%
\subsection{Calibrating an impedance
function}\label{calibrating-an-impedance-function}}

Impedance functions are useful to understand mobility behaviour and are
used to estimate gravity models of spatial interaction
\citep{wilson1971, haynes_gravity_1985} and applied in accessibility
analysis
\citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
An impedance function \(f(\cdot)\) depends on the cost of travel
\(c_{ij}\) between locations \(i\) and \(j\) (all which is supplied in
the travel time and origin-destination table within \{TTS2016R\}).

A useful technique to calibrate an impedance function is to use the trip
length distribution (TLD) as measured from origin-destination data
\citep{horbachov_theoretical_2018, batista_estimation_2019}. The TLD is
the representation of the likelihood that a proportion of trips are
taken at a specific travel cost. In our data set, where we assume cost
is travel time, the impedance function maps low travel times to higher
proportions of trips, and high travel times are mapped to low proportion
of trips.

Using the data contained in \{TTS2016R\}, we fit the empirical TLD to a
density distribution using maximum likelihood techniques and the
Nelder-Mead method for direct optimization available within the
\texttt{R} package \{fitdistrplus\} \citep{fitdistrplus_2015}. Based on
goodness-of-fit criteria and diagnostics seen in Figure
\ref{fig:TLD-Gamma-plot}, the gamma distribution is selected. The
`shape' parameter is \(\alpha\) = 2.019, the estimated `rate' is
\(\beta\) = 0.094 , and \(\Gamma(\alpha)\) is defined in Equation
(\ref{gamma-dist}).

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{images/impedance_function}

}

\caption{\label{fig:TLD-Gamma-plot}Empirical TTS 2016 home-based car TLD (black) and calibrated gamma distribution impedance function (red) with associated Q-Q and P-P plots}\label{fig:TLD-Gamma-plot}
\end{figure}

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}

\newpage

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

The open data product introduced in this paper shares tables for
home-to-work related data from the 2016 TTS. In addition, inter-centroid
travel time tables are calculated, and the planning/municipality
boundaries are included as a compliment. This open data product,
\{TTS2016R\}, is freely available to explore in an \texttt{R}
environment. One possible use of this data, as showcased in this paper,
is the calibration of impedance functions which in turn can be used for
accessibility analysis.

In the spirit of novel and original research, we hope readers value the
efforts made to detail the data in order to improve transparency in our
work and encourage others to replicate and, hopefully, inspire research
of their own. We see this product as providing open infrastructure for
additional TTS or complimentary data sets to be amended by the authors
or wider open-source community in the future.

\bibliographystyle{sageh}
\bibliography{bibfile.bib}

\end{document}

### 2023-01-23 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
Update to vignette after journal article published.

@@
 ---
+
 ```{r knitr-setup, include=FALSE}
@@
 # Home-to-work commute data
+
 <!-- NOTE: When extracting trip vs.
@@
 -->
+
 Currently, {TTS2016R} includes counts of full-time employed population by place of residence (origin), counts of full-time usual place of work (destination), number of trips to work by mode, and the calculated potential travel time of the trips in the GGH.
@@
 ```
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
@@
 ```
+
 ```{r creating-desc-stats-table-edits}
@@
 ```
+
 ```{r creating-desc-stats-table}
@@
 ```
+
 ```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, and calculated OD car travel time (TT) per origin TAZ.
@@
 Additionally, the total number of full-time workers and jobs in the TTS 2016 region are not equal.
-Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but have workplaces that are outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed.
+Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but have workplaces that are outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are *unknown* since they were not surveyed.
 This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs.
@@
 TAZ contain between `r min(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")` workers (median: `r median(TTS2016R::ggh_taz$workers) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$workers, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), `r min(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` to `r max(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")` jobs (median: `r median(TTS2016R::ggh_taz$jobs) %>% prettyNum(digits =2, big.mark = ",")`, 3rd quantile: `r quantile(TTS2016R::ggh_taz$jobs, probs=0.75) %>% unname() %>% prettyNum(digits =2, big.mark = ",")`), and generate between 0 to 241 trips (median: 15, 3rd quantile: 42).
+
 ```{r, include=FALSE, eval=FALSE, echo=FALSE}
@@
 ```
+
+````{=html}
 <!--
@@
 -->
+````
+
 Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed people and associated jobs per TAZ.
@@
 If the readership is interested in additional information regarding the travel time computation, please see the calculation notebook in the documentation of {TTS2016R} and details about {r5r} at the [r5r package website](https://ipeagit.github.io/r5r/index.html).
+
 ```{r data-for-impedance}
@@
 ```
+
 \newpage
@@
 An application of the {TTS2016R} package is the calculation of an impedance function.
-Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@talen_assessing_1998;@paez_jobs_2013;@barboza_balancing_2021].
+Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959; @talen_assessing_1998; @paez_jobs_2013; @barboza_balancing_2021].
 An impedance function $f(\cdot)$ depends on the cost of travel $c_{ij}$ between locations $i$ and $j$ (all which is supplied in the travel time and origin-destination table within {TTS2016R}).
@@
 The 'shape' parameter is $\alpha$ = `r round(gamma_$estimate[1], 3)`, the estimated 'rate' is $\beta$ = `r round(gamma_$estimate[2], 3)` , and $\Gamma(\alpha)$ is defined in Equation (\ref{gamma-dist}).
+
 ```{r save-impedance-plot, include=FALSE}
@@
 
+```{=tex}
 \renewcommand{\[}{\begin{equation}}
 \renewcommand{\]}{\end{equation}}
-
+```
 $$
@@
 ```
+
 <!-- # ```{r plot-cullen-frey, message = FALSE, fig.height=6, fig.cap="\\label{fig:plot-cullen-frey}Cullen and frey graphy for the 2016 TTS calculated travel times."} -->
+
 <!-- # fitdistrplus::descdist(data=all_tt)  -->
+
 <!-- # ``` -->
@@
 With this motivation, we invite all who are interested to use {TTS2016R} to explore the worker-employment patterns contained in the {TTS2016R} package.
---> In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own.
+-->
+
+In the spirit of novel and original research, we hope readers value the efforts made to detail the data in order to improve transparency in our work and encourage others to replicate and, hopefully, inspire research of their own.
 We see this product as providing open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.

### 2023-01-23 R/TTS2016R-package.R

@@ -9,14 +9,14 @@ NULL
 
 #' TTS2016R
 #'
-#' A data package which consists of objects sourced from the [2016
-#' Transportation Tomorrow Survey (TTS)](http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction)
+#' A data package which consists of objects sourced from the
+#' [2016 Transportation Tomorrow Survey (TTS)](http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction)
 #' or are curated to facilitate the use and analysis of TTS data. This package includes worker-to-jobs origin
 #' -destinations and traffic analysis zone (TAZ) boundaries for the Greater Golden Horse area (GGH) located in
 #' southern Ontario, Canada. The package also includes associated spatial boundary files.
 #'
 #' @docType package
-#' @name TTS2016R-package
+#' @name TTS2016R
 #' @author Antonio Paez, School of Earth, Environment and Society, McMaster University \email{paezha@@mcmaster.ca}
 #' @author Anastasia Soukhov, School of Earth, Environment and Society, McMaster University \email{soukhoa@@mcmaster.ca}
 #' @references \url{https://github.com/soukhova/TTS2016R}

### 2023-01-23 R/TTS2016R-package.R
building package site

@@ -16,7 +16,7 @@ NULL
 #' southern Ontario, Canada. The package also includes associated spatial boundary files.
 #'
 #' @docType package
-#' @name TTS2016R
+#' @name TTS2016R-package
 #' @author Antonio Paez, School of Earth, Environment and Society, McMaster University \email{paezha@@mcmaster.ca}
 #' @author Anastasia Soukhov, School of Earth, Environment and Society, McMaster University \email{soukhoa@@mcmaster.ca}
 #' @references \url{https://github.com/soukhova/TTS2016R}

### 2023-10-12 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
Update to vignette after journal article published.
New table with households by taz by income range

@@
 ```{r load-packages}
-library(readxl)
-library(sf)
-library(tidyverse)
-library(units)
-library(tmap)
-library(ggplot2)
+library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(logspline) # not installed on this machine
+library(readxl) # not installed on this machine
+library(sf) # Simple Features for R
+library(tidyverse) # not installed on this machine
+library(tmap) # Thematic Maps
+library(units) # Measurement Units for R Vectors
 ```

### 2023-10-12 R/TTS2016R-package.R

@@ -22,6 +22,45 @@ NULL
 #' @references \url{https://github.com/soukhova/TTS2016R}
 NULL
 
+
+
+#' census metropolitan area (CMA)/census agglomeration (CA) boundaries in the Greater Golden Horseshoe (GGH) area and associated data.
+#'
+#' This object contains the census metropolitan area (CMA)/census agglomeration (CA) for the spatial region  associated with the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.
+#'
+#' @format A simple feature class (sf) polygon object containing 16 rows and 4 variables; each row represents a unique census metropolitan area (CMA) of census agglomeration (CA) with associated geo-referenced geometry. CMA and CA consist of one or more adjacent municipalities around a core. The CMA/CA boundaries are defined for the purpose of the Canadian Census based on population. As such, the core of each CMA must have a population of at least 50,000 and the entire CMA must have a population of at least 100,000. CA are smaller in population, thir course must have a population of at least 10,000.
+#' \describe{
+#'   \item{CMAUID}{A unique identifier for each census metropolitan area (CMA)/census agglomeration (CA).}
+#'   \item{CMANAME}{The name of the census metropolitan area (CMA)/census agglomeration (CA).}
+#'   \item{CMATYPE}{A one character field indicating whether the unit is a census metropolitan area, a tracted census agglomeration or a non-tracted census agglomeration. "B" is a CMA and "D" is a CA with no census tracts, and "K" is a CA with census tracts.}
+#'   \item{geometry}{The sfc polygon geometry (boundaries).}
+#'}
+#' @docType data
+#' @keywords planning districts regions municipalities boundary
+#' @name ggh_cma
+#' @usage data(ggh_cma)
+#' @source "The 2016 census metropolitan areas and census agglomeration boundary files as created by the Canadian Census available [here](https://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/bound-limit-2016-eng.cfm) accessed October 21st 2022 (Boundary Files, 2016 Census. Statistics Canada Catalogue no. 92-160-X.). All variable definitions are based on the definitions included in the census year 2016 boundary file reference guide (Boundary Files, Reference Guide, Second edition, 2016 Census. Statistics Canada Catalogue no. 92-160-G.).
+"ggh_cma"
+
[…]

### 2023-11-08 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
Working on development branch to ensure reproducibility of data processing and adding new big od table

--- 2022-03-11 data-raw/00-TTS16-and-Bound-Data-Preparation-OD.Rmd
@@
 ```{r load-packages}
-library(readxl)
-library(sf)
-library(tidyverse)
-library(units)
+library(dplyr) # A Grammar of Data Manipulation
+library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+#library(logspline) # not installed on this machine
+library(readr) # not installed on this machine
+library(sf) # Simple Features for R
+library(stringr)
+library(tidyr)
+#library(tidyverse) # not installed on this machine
+#library(tmap) # Thematic Maps
+library(units) # Measurement Units for R Vectors
 ```
@@
 ```{r}
-ggh_taz <- ggh_taz %>%
+ggh_taz <- ggh_taz |>
   st_transform(crs = 32617)
@@
 ```{r}
-ggh_taz <- ggh_taz %>%
+ggh_taz <- ggh_taz |>
   transmute(GTA06 = as.character(GTA06),
-            AREA = st_area(geometry) %>% 
-              set_units(km^2) %>% 
+            AREA = st_area(geometry) |> 
+              set_units(km^2) |> 
               drop_units())
@@
 ```{r}
-ggh_pd <- ggh_pd %>%
+ggh_pd <- ggh_pd |>
   st_transform(crs = 32617)
@@
 ```{r}
-ggh_pd <- ggh_pd %>%
-  mutate(REGION_name = 
-           ifelse(REGION == 1, "Toronto", 
-                  ifelse(REGION == 2, "Durham", 
-                         ifelse(REGION == 3, "York", 
-                                ifelse(REGION == 4, "Peel", 
-                                       ifelse(REGION == 5, "Halton", 
-                                              ifelse(REGION == 6, "Hamilton", 
-                                                     ifelse(REGION == 11, "Niagara", 
-                                                            ifelse(REGION == 12, "Waterloo", 
-                                                                   ifelse(REGION == 13, "Guelph", 
-                                                                          ifelse(REGION == 14, "Wellington", 
-                                                                                 ifelse(REGION == 15, "Orangeville", 
-                                                                                        ifelse(REGION == 16, "Barrie", 
-                                                                                               ifelse(REGION == 17, "Simcoe", 
-                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
-                                                                                                             ifelse(REGION == 19, "Peterborough City", 
-                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
-                                                                                                                           ifelse(REGION == 21, "Orillia", 
-                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
-                                                                                                                                         ifelse(REGION == 23, "Brantford", 
-                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
+ggh_pd <- ggh_pd |>
+  mutate(REGION_name = case_when(REGION == 1 ~ "Toronto",
+                                  REGION == 2 ~ "Durham",
+                                  REGION == 3 ~ "York",
+                                  REGION == 4 ~ "Peel",
+                                  REGION == 5 ~ "Halton",
+                                  REGION == 6 ~ "Hamilton",
+                                  REGION == 11 ~ "Niagara",
+                                  REGION == 12 ~ "Waterloo",
+                                  REGION == 13 ~ "Guelph",
+                                  REGION == 14 ~ "Wellington",
+                                  REGION == 15 ~ "Orangeville",
+                                  REGION == 16 ~ "Barrie",
+                                  REGION == 17 ~ "Simcoe",
+                                  REGION == 18 ~ "Kawartha Lakes",
+                                  REGION == 19 ~ "Peterborough City",
+                                  REGION == 20 ~ "Peterborough County", 
+                                  REGION == 21 ~ "Orillia", 
+                                  REGION == 22 ~ "Dufferin",
+                                  REGION == 23 ~ "Brantford",
+                                  REGION == 24 ~ "Brant"))
 ```
 
-## Origin destination matrices for work and shop
+add 'REGION' from the planning distract object to the `ggh_taz`. 
+```{r}
+#buffering to clean up the edges of these objects -> there wre issues using st_join. 
+ggh_taz <- st_buffer(ggh_taz, 
+                     dist = 0)
 
-Read trip tables. The query was for "trips" by zone of residence, zone of work, and trip purpose (Work):
+ggh_pd <- st_buffer(ggh_pd, 
+                    dist = 0)
+
+ggh_taz <- st_join(ggh_taz, 
+                   ggh_pd |>
+                     transmute(REGION),
+                   left = TRUE, 
+                   largest = TRUE)
+```
+
+## CMA and CA boundaries
+
+If there's interest in tying the TAZ to the Canadian Census, the boundaries and IDs for census metropolitan areas (CMA) and census ares (CA) are provided in this package as well.
+These areas are determined by Statistics Canada and
+
+Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
+
+<!-- FILE NOT IN FOLDER -->
+```{r}
+ggh_cma <- st_read("TTS16-data-inputs/Boundaries/lcma000b16a_e.shp")
+```
+
+Project object and retain specific columns:
+```{r}
+ggh_cma <- ggh_cma |>
+  st_transform(crs = 32617) |>
+  transmute(CMAUID,
+            CMANAME,
+            CMATYPE)
+```
+
+add 'CMAUID' from the cma object to the `ggh_taz`. 
+```{r}
+#doing a st_join doesn't provide a clean solution, so I find the centroids of each TAZ, st_join them to the CMA/CA, and then join the resulting CMAUID property to the ggh_taz object
+test_pts <-st_centroid(ggh_taz)
+test<-st_join(test_pts, ggh_cma |> transmute(CMAUID), left = TRUE, join=st_within)
+ggh_taz <-merge(ggh_taz, test |> transmute(GTA06, CMAUID)|> st_drop_geometry(), by=c("GTA06"))
+```
+
+## Origin destination matrices for place of residence and typical place work (by person)
+Read exported TTS 2016 tables.
+The query was a cross-tabulation of "persons" by zone of residence and place of employment.
+The data was filtered to only include full-time not-remote workers and was retrieved from the TTS 2016:
 ```{r}
@@
 ```{r}
-od_ft <- od_ft %>% 
+od_ft <- od_ft |> 
   slice((idx+2):n())
@@
 ```{r}
-od_ft <- od_ft %>%
-  separate(X1, into = c("Zones", "Persons"), sep = " (?=[^ ]+$)") %>%
-  mutate(Zones = str_trim(Zones, side = "both")) %>%
-  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
+od_ft <- od_ft |>
+  separate(X1, into = c("Zones", "Persons"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
   mutate(Origin = str_trim(Origin),
@@
 
+## Trips taken (by mode) by worker from place of residence to job at place of work
+
+Read exported TTs 2016 tables.
+The query was a cross-tablulation of "trips" by zone of residence and place of employment.
+The data was filtered to only include full-time not-remote working trip purposes by primary mode and was retrieved from the TTS 2016:
+
+<!-- FILE DOES NOT EXIST IN DIRECTORY -->
+```{r}
+od_ft_trips <- read_delim(file = "TTS16-data-inputs/OD-by-FT-Employment/tts-2016-OD-FT-Employment-by_W_R_trip-byprimmode.txt", 
+                      delim = "\t", 
+                      col_names = FALSE)
+```
+
+Find the position in the table where the information on work trips begins:
+```{r}
+id_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Transit excluding GO rail)")
+id_cycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Cycle)")
+id_car_driver <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto driver)")
+id_GO <- which(od_ft_trips$X1 == "TABLE    : mode_prime (GO rail only)")
+id_GO_transit <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Joint GO rail and local transit)")
+id_motorcycle <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Motorcycle)")
+id_other <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Other)")
+id_car_pass <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Auto passenger)")
+id_schoolbus <- which(od_ft_trips$X1 == "TABLE    : mode_prime (School bus)")
+id_taxi <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Taxi passenger)")
+id_rideshare <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Paid rideshare)")
+id_walk <- which(od_ft_trips$X1 == "TABLE    : mode_prime (Walk)")
+id_end <- od_ft_trips$X1 |> length()
+```
+
+Slice the table to obtain the work trips by primary mode (there are 12 possible modes, so 12 new tables) (remove heading information and other modes):
+```{r}
+od_ft_trips_transit <- od_ft_trips |> 
+  slice((id_transit+3):id_cycle-1)
+
+od_ft_trips_cycle <- od_ft_trips |> 
+  slice((id_cycle+3):id_car_driver-1)
+
+od_ft_trips_car_driver <- od_ft_trips |> 
+  slice((id_car_driver+3):id_GO-1)
+
+od_ft_trips_GO <- od_ft_trips |> 
+  slice((id_GO+3):id_GO_transit-1)
+
+od_ft_trips_GO_transit <- od_ft_trips |> 
+  slice((id_GO_transit+3):id_motorcycle-1)
+
+od_ft_trips_motorcycle <- od_ft_trips |> 
+  slice((id_motorcycle+3):id_other-1)
+
+od_ft_trips_other <- od_ft_trips |> 
+  slice((id_other+3):id_car_pass-1)
+
+od_ft_trips_car_pass <- od_ft_trips |> 
+  slice((id_car_pass+3):id_schoolbus-1)
+
+od_ft_trips_schoolbus <- od_ft_trips |> 
+  slice((id_schoolbus+3):id_taxi-1)
+
+od_ft_trips_taxi<- od_ft_trips |> 
+  slice((id_taxi+3):id_rideshare-1)
+
+od_ft_trips_rideshare <- od_ft_trips |> 
+  slice((id_rideshare+3):id_walk-1)
+
+od_ft_trips_walk <- od_ft_trips |> 
+  slice((id_walk+2):id_end)
+```
+
+Let's consolidate the 12 tables into 1 table, based on origin - destination and clean up the table. If there's no trip using that mode, a 0 is added. The following are the codes:
+
+- T_LT - Local transit excluding GO rail
+- T_LTGO - Local transit and GO
+- T_GO - GO rail excluding local transit
+```{r}
+od_ft_trips_transit <- od_ft_trips_transit |>
+  separate(X1, into = c("Zones", "T_LT"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_LT = as.numeric(T_LT))
+
+od_ft_trips_GO_transit <- od_ft_trips_GO_transit |>
+  separate(X1, into = c("Zones", "T_LTGO"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_LTGO = as.numeric(T_LTGO))
+
+od_ft_trips_GO <- od_ft_trips_GO |>
+  separate(X1, into = c("Zones", "T_GO"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         T_GO = as.numeric(T_GO))
+```
+
+- BC - Cycle
+- W - Walk
+```{r}
+od_ft_trips_cycle <- od_ft_trips_cycle |>
+  separate(X1, into = c("Zones", "BC"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         BC = as.numeric(BC))
+
+od_ft_trips_walk <- od_ft_trips_walk |>
+  separate(X1, into = c("Zones", "W"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         W = as.numeric(W))
+```
+
+- C_DPC - Driver private car 
+- C_PPC - Passenger private car
+- C_TA - Taxi
+- C_RS - Rideshare
+- C_SB - School bus
+```{r}
+od_ft_trips_car_driver <- od_ft_trips_car_driver |>
+  separate(X1, into = c("Zones", "C_DPC"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_DPC = as.numeric(C_DPC))
+
+od_ft_trips_car_pass <- od_ft_trips_car_pass |>
+  separate(X1, into = c("Zones", "C_PPC"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_PPC = as.numeric(C_PPC))
+
+od_ft_trips_taxi <- od_ft_trips_taxi |>
+  separate(X1, into = c("Zones", "C_TA"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_TA = as.numeric(C_TA))
+
+od_ft_trips_rideshare <- od_ft_trips_rideshare |>
+  separate(X1, into = c("Zones", "C_RS"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_RS = as.numeric(C_RS))
+
+od_ft_trips_schoolbus <- od_ft_trips_schoolbus |>
+  separate(X1, into = c("Zones", "C_SB"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         C_SB = as.numeric(C_SB))
+```
+
+- M - Motorcycle
+- O - Other
+```{r}
+od_ft_trips_motorcycle<- od_ft_trips_motorcycle |>
+  separate(X1, into = c("Zones", "M"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         M = as.numeric(M))
+
+od_ft_trips_other <- od_ft_trips_other |>
+  separate(X1, into = c("Zones", "O"), sep = " (?=[^ ]+$)") |>
+  mutate(Zones = str_trim(Zones, side = "both")) |>
+  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") |>
+  mutate(Origin = str_trim(Origin),
+         Destination = str_trim(Destination),
+         O = as.numeric(O))
+```
+
+And now, consolidate all 12 tables into 1 table! with 12 columns for mode trips numbers + the origin and destination columns (those are GTA06 TAZ IDs):
+```{r}
+od_ft_trips <- full_join(od_ft_trips_transit, od_ft_trips_GO_transit, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_GO, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_cycle, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_walk, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_car_driver, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_car_pass, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_taxi, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_rideshare, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_schoolbus, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_motorcycle, by=c("Origin", "Destination")) |>
+  full_join(od_ft_trips_other, by=c("Origin", "Destination")) 
+```
+
+Replace all NAs with 0s:
+```{r}
+od_ft_trips[is.na(od_ft_trips)] <- 0
+```
+
+Combine the `od_ft` person data frame with the `od_ft_trips` data frame.
+Notice that there are more OD pairs in the `od_ft` data frame (`r od_ft$Origin |> length()`) than `od_ft_trips` (`r od_ft_trips$Origin |> length()`) because some OD pairs which have workers and/or are a location for jobs do not have any trips taken that survey day.
+The TTS 2016 is a travel survey after all.
+```{r}
+od_ft_p_t <- full_join(od_ft, od_ft_trips, by=c("Origin", "Destination")) 
+```
+
+Check to see if the total number of trips adds up to the total available from TTS 2016 (to ensure we separated and merged all the tables correctly)
+```{r}
+od_ft_p_t <- od_ft_p_t |> rowwise() |>
+  mutate(sum_trips = sum(T_LT,T_LTGO,T_GO,BC,W,C_DPC,C_PPC,C_TA,C_RS,C_SB,M,O, na.rm=T))
+
+#the number of total trips.
+This value is slightly higher than the value for total trips (not disagg.
+by mode, which is equal to 3,282,539 - I checked in excel) as available from the TTS 2016.
+However, I checked the summed up total trip, disaggregated by mode, in excel, and it yields this number of 3,282,611 as well.
+So there must just be some rounding up that occurred in the TTS 2016 when disaggregating by mode...
+regardless the numerical difference is insignificant.
+od_ft_p_t$sum_trips |> sum()
+```
+
+Replace all NAs with 0s, again:
+```{r}
+od_ft_p_t[is.na(od_ft_p_t)] <- 0
+```
+
 ## Workers and Jobs
 
-The od table can be used to calculate the workers and jobs by traffic analysis zone:
+The od table can be used to calculate the count of workers, count of jobs, and count of % transit, bike, and walk slit, by traffic analysis zone:
 ```{r}
-workers_ft <- od_ft %>%
-  group_by(Origin) %>%
+workers_ft <- od_ft_p_t |>
+  group_by(Origin) |>
   summarize(workers = sum(Persons),
-            .groups = "drop") %>%
+            .groups = "drop") |>
   transmute(GTA06 = Origin,
@@
 
-jobs_ft <- od_ft %>%
-  group_by(Destination) %>%
+jobs_ft <- od_ft_p_t |>
+  group_by(Destination) |>
   summarize(jobs = sum(Persons),
-            .groups = "drop") %>%
+            .groups = "drop") |>
   transmute(GTA06 = Destination,
             jobs)
+```
 
+Active travel modal splits:
+```{r}
+LTGO_split_ft <- od_ft_p_t |>
+  group_by(Origin) |>
+  summarize(LTGO_split = sum(T_LT, T_LTGO, T_GO)/sum(sum_trips)) |> 
+  transmute(GTA06 = Origin,
+            LTGO_split)
+
+W_split_ft <- od_ft_p_t |>
+  group_by(Origin) |>
+  summarize(W_split = sum(W)/sum(sum_trips)) |>
+  transmute(GTA06 = Origin,
+            W_split)
+
+BC_split_ft <- od_ft_p_t |>
+  group_by(Origin) |>
+  summarize(BC_split = sum(BC)/sum(sum_trips)) |>
+  transmute(GTA06 = Origin,
+            BC_split)
 ```
@@
 ```{r}
-ggh_taz <- ggh_taz %>%
+ggh_taz <- ggh_taz |>
   # Join workers
   left_join(workers_ft,
-            by = "GTA06") %>%
+            by = "GTA06") |>
   # Join jobs
   left_join(jobs_ft,
-            by = "GTA06") %>%
-  # Replace NAs with zeros
-  mutate(workers = replace_na(workers, 0),
-         jobs = replace_na(jobs, 0))
+            by = "GTA06") |>
+  left_join(LTGO_split_ft,
+            by = "GTA06") |>
+  left_join(W_split_ft,
+            by = "GTA06") |>
+  left_join(BC_split_ft,
+            by = "GTA06")
 ```
+
+Replace NAs with 0s for the new columns:
+```{r}
+ggh_taz <- ggh_taz |>
+  mutate_at(vars(workers, jobs, LTGO_split, W_split, BC_split), ~replace_na(., 0))
+```
+
+## Visualize
 
@@
 
+Plot transit split (0 to 1, 1 being 100% of trips made using transit) distribution in the ggh. Aggregating by CMA could be interesting:
+```{r}
+ggplot() +
+  geom_sf(data = ggh_taz,
+          aes(fill = LTGO_split),
+          color = NA) +
+  scale_fill_distiller(direction = 1)
+```
+Let's zoom into Toronto CMA, and plot the transit split
+```{r}
+#rotating the WGS84 crs so it displays Toronto aesthetically on a page; 17degree shift using +gamma
+# lat_0 and lonc are the approximate centroid of Toronto
+crs_string <- "+proj=omerc +lat_0=43.6532 +lonc=-79.3832 +alpha=0 +k_0=.7 +datum=WGS84 +units=m +no_defs +gamma=17"
+
+#the bounding box, so we focus on Toronto
+toronto_bb <- st_bbox(ggh_cma |> filter(CMANAME %in% c("Toronto")))
+
+tm_shape(ggh_taz, bbox = toronto_bb, projection=crs_string) +
+  tm_polygons(col = "LTGO_split", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_cma |> filter(CMANAME %in% c("Toronto")), bbox = toronto_bb, projection=crs_string) +
+  tm_borders()
+```
+The CMA/CAs overlapping all the TAZ:
+```{r}
+tm_shape(ggh_taz, projection=crs_string) +
+  tm_polygons(col = "CMAUID", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_cma, projection=crs_string) +
+  tm_borders()
+```
+
+The planning boundaries overlapping all the TAZ:
+```{r}
+tm_shape(ggh_taz, projection=crs_string) +
+  tm_polygons(col = "REGION", border.alpha = 0.1, style = "jenks") +
+  tm_shape(ggh_pd, projection=crs_string) +
+  tm_borders()
+```
+
 ## Creating input for travel-time OD calculations
@@
 #workers (origins)
-workers_ft_cent <- workers_ft %>% merge(ggh_taz_cent %>% dplyr::select(GTA06), by="GTA06") %>% st_sf()
+workers_ft_cent <- workers_ft |> merge(ggh_taz_cent |> dplyr::select(GTA06), by="GTA06") |> st_sf()
 
-work_origins <- cbind(workers_ft_cent, st_coordinates(st_transform(workers_ft_cent, crs = 4326))) %>%
-  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
-  dplyr::select(id, lon, lat) %>% as.data.frame()
+work_origins <- cbind(workers_ft_cent, st_coordinates(st_transform(workers_ft_cent, crs = 4326))) |>
+  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") |> as.data.frame()  |>
+  dplyr::select(id, lon, lat) |> as.data.frame()
 
@@
 #jobs (destinations)
-jobs_ft_cent <- jobs_ft %>% merge(ggh_taz_cent %>% dplyr::select(GTA06), by="GTA06") %>% st_sf
+jobs_ft_cent <- jobs_ft |> merge(ggh_taz_cent |> dplyr::select(GTA06), by="GTA06") |> st_sf
 
-job_destinations <- cbind(jobs_ft_cent, st_coordinates(st_transform(jobs_ft_cent, crs = 4326))) %>%
-  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") %>% as.data.frame()  %>%
+job_destinations <- cbind(jobs_ft_cent, st_coordinates(st_transform(jobs_ft_cent, crs = 4326))) |>
+  dplyr::rename(lon = "X", lat = "Y", id = "GTA06") |> as.data.frame()  |>
   dplyr::select(id, lon, lat)

### 2023-11-08 data-raw/01-TTS16-Calculating-Travel-Time-OD.Rmd
Working on development branch to ensure reproducibility of data processing and adding new big od table

@@
 ```{r setup, include=FALSE}
-library(disk.frame)
-library(dplyr)
-library(ggplot2)
-library(gridExtra)
-library(httr)
-library(knitr)
+library(disk.frame) # not installed on this machine
+library(dplyr) # A Grammar of Data Manipulation
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
+library(httr) # Tools for Working with URLs and HTTP
+library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
 #install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
-library(kableExtra)
+library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
 # install.packages("devtools")
@@
 # devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
-library(lodes)
-library(progress)
-library(purrr)
-library(r5r)
-library(sf)
-library(stplanr)
-library(tidycensus)
-library(tidyr)
-library(tmap)
+#library(lodes) # not installed on this machine
+library(progress) # Terminal Progress Bars
+library(purrr) # Functional Programming Tools
+library(sf) # Simple Features for R
+library(stplanr) # not installed on this machine
+library(tidycensus) # not installed on this machine
+library(tidyr) # Tidy Messy Data
+library(tmap) # Thematic Maps
 library(zoo) # for rollapplyr
@@
 options(future.globals.maxSize = Inf) # allow disk.frame threads to talk to eachother without limits
-options(java.parameters = "-Xmx6G") # assign memory for r5r to work with
 tmap_mode("plot")
+```
+
+```{r}
+options(java.parameters = "-Xmx30G") # assign memory for r5r to work with
+library(r5r) # Rapid Realistic Routing with 'R5' 
 ```

### 2023-11-08 R/TTS2016R-package.R

@@ -26,7 +26,7 @@ NULL
 
 #' census metropolitan area (CMA)/census agglomeration (CA) boundaries in the Greater Golden Horseshoe (GGH) area and associated data.
 #'
-#' This object contains the census metropolitan area (CMA)/census agglomeration (CA) for the spatial region  associated with the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.
+#' This object contains the census metropolitan area (CMA)/census agglomeration (CA) for the spatial region  associated with the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.  The object is projected as WGS 84 / UTM zone 17N (epsg: 32617).
 #'
 #' @format A simple feature class (sf) polygon object containing 16 rows and 4 variables; each row represents a unique census metropolitan area (CMA) of census agglomeration (CA) with associated geo-referenced geometry. CMA and CA consist of one or more adjacent municipalities around a core. The CMA/CA boundaries are defined for the purpose of the Canadian Census based on population. As such, the core of each CMA must have a population of at least 50,000 and the entire CMA must have a population of at least 100,000. CA are smaller in population, thir course must have a population of at least 10,000.
 #' \describe{
@@ -44,7 +44,7 @@ NULL
 
 #' Planning boundaries in the Greater Golden Horseshoe (GGH) area and associated data.
 #'
-#' This object contains the planning boundaries (PB) associated with the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.
+#' This object contains the planning boundaries (PB) associated with the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.  The object is projected as WGS 84 / UTM zone 17N (epsg: 32617).
 #'
 #' @format A simple feature class (sf) polygon object containing 105 rows and 5 variables; each row represents a unique planning region with associated geo-referenced geometry.
 #' \describe{
@@ -63,7 +63,7 @@ NULL
 
 #' Traffic analysis zones boundaries in the Greater Golden Horseshoe (GGH) area and associated data.
 #'
-#' This object contains traffic analysis zones (TAZ) sourced from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe (GGH) area.
[…]

### 2023-11-09 R/TTS2016R-package.R
Revision to version 0.1-1

@@ -137,7 +137,7 @@ NULL
 #'}
 #' @docType data
 #' @keywords Jobs Workers TTS 2016 in the Greater Golden Horseshoe (GGH) area.
-#' @name ggh_taz
-#' @usage data(ggh_taz)
+#' @name hh_sed_taz
+#' @usage data(hh_sed_taz)
 #' @source "2016 Transportation Tomorrow Survey" from [Data Management Group](http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction) accessed November 14 2021.
 "hh_sed_taz"

### 2023-11-09 data-raw/00-TTS16-and-Household-Statistics.Rmd
Adding preprocessing files for newest table

---
title: "TTS Origin-Destination Data by Full-Time Employment"
output: 
---

In this notebook we process socio-economic and demographic information about households in their places of residence.

Household income ranges at the level of GTS2006 traffic analysis zones. This is the query:

```
Cross Tabulation Query Form - Person - 2016 v1.1

Row: 2006 GTA zone of household - gta06_hhld
Column: Income range of household - income

Filters:
No Filters
```
The data are in this table:

- `tts-2016-household-income-by-residence.txt`, retrieved on October 12, 2023

## Preliminaries

Load packages:
```{r load-packages}
library(dplyr) # A Grammar of Data Manipulation
library(readr) # Read Rectangular Text Data
library(tidyr) # Tidy Messy Data
```

## Number of household by income range by traffic analysis zones

Read the raw data table
```{r}
hh_sed_taz <- read_delim(file = "TTS16-data-inputs/Household-Income/tts-2016-household-income-by-taz.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

Find the position in the table where the information on work trips begins:
```{r}
idx <- which(hh_sed_taz$X1 == "Table: ")
```

Slice the table to obtain the work trips (remove heading information):
```{r}
hh_sed_taz <- hh_sed_taz |> 
  slice((idx+1):n())
```

Separate the columns and convert number of households (initially read as a character) to numeric:
```{r}
hh_sed_taz <- hh_sed_taz |>
  separate(X1, into = c("GTA06",
                        "$0 to $14999",
                        "$15000 to $39999",
                        "$40000 to $59999",
                        "$60000 to $99999",
                        "$100000 to $124999",
                        "$125000 and above",
                        "Decline / do not know"),
           sep = ",") |>
  slice_tail(n = nrow(hh_sed_taz) - 1) |>
  mutate(`$0 to $14999` = as.numeric(`$0 to $14999`),
         `$15000 to $39999` = as.numeric(`$15000 to $39999`),
         `$40000 to $59999` = as.numeric(`$40000 to $59999`),
         `$60000 to $99999` = as.numeric(`$60000 to $99999`),
         `$100000 to $124999` = as.numeric(`$100000 to $124999`),
         `$125000 and above` = as.numeric(`$125000 and above`),
         `Decline / do not know` = as.numeric(`Decline / do not know`))
```

Save to data folder:
```{r}
usethis::use_data(hh_sed_taz)
```

### 2023-11-09 data-raw/03-TTS16-GGH4-Bound-Data-Preparation-OD.Rmd
Adding preprocessing files for newest table

---
title: "R Notebook"
output: html_notebook
---

This notebook is to read a shapefile. The file is a zoning system used by Metrolinx and possibly compatible with TTS. I got this file from Matt Pittlock on November 2, 2023. The file is `OriginalV5_10202023`.

Load packages:
```{r}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(sf) # Simple Features for R
```

Read file:
```{r}
ggh_taz_v4 <- st_read("TTS16-data-inputs/Boundaries/OriginalV5_10202023.shp")
```

```{r}
ggh_taz_v4 <- st_make_valid(ggh_taz_v4)
```

Plot:
```{r}
ggplot() +
  geom_sf(data = st_make_valid(ggh_taz_v4))
```

Summary of object:
```{r}
summary(ggh_taz_v4)
```

Keep only useful columns:
```{r}
ggh_taz_v4 <- ggh_taz_v4 |>
  transmute(GGHV4 = taz, 
            CSDUID = csduid, 
            CSDNAME = csdname, 
            CMAUID = cmauid, 
            CMANAME = cmaname, 
            CSDUID = csduid, 
            PD = pd)
```

Transform the object so that the projection is consistent with other geographical objects.

Read a simple features object:
```{r}
load(glue::glue(here::here(), "/data/ggh_taz.rda"))
```

The object is projected as NAD83 / UTM zone 17N (epsg: 26917).

```{r}
ggh_taz_v4 <- ggh_taz_v4 |>
  st_transform(crs = st_crs(ggh_taz))
```

Plot:
```{r}
ggplot() +
  geom_sf(data = st_make_valid(ggh_taz_v4))
```

Save as simple features:
```{r}
usethis::use_data(ggh_taz_v4, 
                  overwrite = TRUE)
```

### 2023-11-09 data-raw/04-GGH4-Extract-OD-by-Purpose-Mode-Time-Day-from-Raw-Files.Rmd
Adding preprocessing files for newest table

---
title: "TTS OD by Purpose by Start Time/Day of Week"
output: 
---

In this notebook I process _trips_ by origin and destination, cross-tabulated by start time/day of week and purpose.

All tables were retrieved from TTS on November 2-3, 2023. The type of the table is _Trips_ (other table types in TTS are _Household_, _Person_, and _Transit_).

The naming convention for the tables is `ttsCross-OD-GGH4-{Purpose}-{Day}-{Mode}.txt`.

- ttsCross indicate that it is a cross-tabulation.
- OD indicates that the cross-tabulation is by origin and destination.
- GGH4 indicate the zoning system, which is GGHM V4.
- {Purpose} is the purpose of the trip.
- {Day} is the name of the trip.
- {Mode} is the primary mode of the trip.

The possible values of {Purpose} and their definitions when not self-explanatory are:

- Work
- Work2: Subsequent work
- School
- School2: Subsequent school
- Home
- Daycare
- Facilitate: Facilitate passenger
- Market: Market/Shop
- Other
- (Unkown: not downloaded)

The possible values for {Day} are (TTS does not collect data for weekends):

- Monday
- Tuesday
- Wednesday
- Thursday
- Friday
- (Unknown: not downloaded)

The possible values for {Mode} and their definitions (when not self-explanatory) are:

- Transit: Transit excluding GO rail
- Cycle 
- Driver: Auto driver 
- GO: GO rail only
- GO+Transit: Joint GO rail and local transit 
- Motorcycle 
- Other
- School: School bus
- Passenger: Auto passenger
- Taxi: Taxi passenger
- Rideshare: Paid rideshare
- Walk
- (Unknown: not downloaded)

The Data Guide can be found [here](http://dmg.utoronto.ca/wp-content/uploads/2022/06/2016TTS_DataGuide.pdf).

## Preliminaries

Load packages:
```{r load-packages}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
library(lubridate) # Make Dealing with Dates a Little Easier
library(readr) # Read Rectangular Text Data
library(sf) # Simple Features for R
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(tidyr) # Tidy Messy Data
```

# Checks

We want to make sure that the tables were downloaded correctly and there are no mismatches between the name of the table and the contents.

Define the parameters for the names of the tables:
```{r}
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
purpose <- c("Work", "Work2", "School", "School2", "Home", "Daycare", "Facilitate", "Market", "Other")
mode <- c("Transit", "Cycle", "Driver", "GO", "GO+Transit", "Motorcycle", "Other", "School", "Passenger", "Taxi", "Rideshare", "Walk")
```

Loop through the tables and check the filters recorded in each:
```{r}
# Parameters for testing, commented out
d <- 1
p <- 2
m <- 6

# Initialize concordance with the values in file names and the values in filters inside the files
concordance <- data.frame(prps_file = NULL, day_file = NULL, md_file = NULL,
                          prps = NULL, day = NULL, md = NULL)

for(p in 1:length(purpose)){
  # Loop through modes
  for(m in 1:length(mode)){
    # Loop through weekdays
    for(d in 1:length(weekday)){
      # Read the files in turn
      raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
                             delim = "\t", 
                             col_names = FALSE,
                             show_col_types = FALSE)

      # Identify the locations in the table where the filters are recorded and store them in a temporary data frame
      # Record values of the file being read
      concordance_temp <- data.frame(prps_file = purpose[p],
                                     day_file = weekday[d],
                                     md_file = mode[m],
                                     # Extract the values of the filters as found in the file
                                     prps = raw_data |> 
                                       slice(which(str_detect(raw_data$X1, "Trip purpose of destination"))) |>
                                       transmute(prps = str_sub(X1, -3, -3)) |>
                                       pull(prps),
                                     day = raw_data |> 
                                       slice(which(str_detect(raw_data$X1, "Day of week trip data were collected"))) |>
                                       transmute(day = str_sub(X1, -3, -3)) |>
                                       pull(day),
                                     md = raw_data |> 
                                       slice(which(str_detect(raw_data$X1, "Primary travel mode of trip"))) |>
                                       transmute(md = str_sub(X1, -3, -3)) |>
                                       pull(md)) |>
        mutate(prps = case_when(prps == "W" ~ "Work",
                                prps == "R" ~ "Work2",
                                prps == "S" ~ "School",
                                prps == "C" ~ "School2",
                                prps == "H" ~ "Home",
                                prps == "D" ~ "Daycare",
                                prps == "F" ~ "Facilitate",
                                prps == "M" ~ "Market",
                                prps == "O" ~ "Other"),
               day = case_when(day == "1" ~ "Monday",
                               day == "2" ~ "Tuesday",
                               day == "3" ~ "Wednesday",
                               day == "4" ~ "Thursday",
                               day == "5" ~ "Friday",
                               day == "B" ~ "Transit",
                               day == "C" ~ "Monday"),
               md = case_when(md == "B" ~ "Transit",
                              md == "C" ~ "Cycle",
                              md == "D" ~ "Driver",
                              md == "G" ~ "GO",
                              md == "J" ~ "GO+Transit",
                              md == "M" ~ "Motorcycle",
                              md == "O" ~ "Other",
                              md == "P" ~ "Passenger",
                              md == "S" ~ "School",
                              md == "T" ~ "Taxi",
                              md == "U" ~ "Rideshare",
                              md == "W" ~ "Walk"))

      # Bind the latest check to the concordance
      concordance <- rbind(concordance,
                           concordance_temp)
    }
  }
}

concordance <- concordance |>
  mutate(prps_check = prps_file == prps,
         day_check = day_file == day,
         md_check = md_file == md)
```

Review the checks:
```{r}
concordance |>
  summarize(prps_check = sum(prps_check),
            day_check = sum(day_check),
            md_check = sum(md_check))
```

All the checks pass.

# OD tables by purpose, time of day and day of week

Slice the table to obtain the trips (remove heading information) and add day:
```{r}
# Define the parameters for reading the files and organizing the data: weekday, purpose and mode
# weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
# purpose <- c("Work", "Work2", "School", "School2", "Home", "Daycare", "Facilitate", "Market", "Other")
# mode <- c("Transit", "Cycle", "Drive", "GO", "GO+Transit", "Motorcycle", "Other", "School", "Passenger", "Taxi", "Rideshare", "Walk")

weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
purpose <- c("Work", "Work2", "School", "School2", "Home", "Daycare", "Facilitate", "Market", "Other")
mode <- c("Transit", "Cycle", "Driver", "GO", "GO+Transit", "Motorcycle", "Other", "School", "Passenger", "Taxi", "Rideshare", "Walk")

# Initialize data frame for the processed data. There will be six columns, for origin/destination ids, number of trips, start time of trip, day of week, and purpose of trip at destination.
od_purpose_time <- data.frame(ggh4_orig = NULL, 
                              ggh4_dest = NULL, 
                              Trips = NULL, 
                              start_time = NULL, 
                              day = NULL, 
                              purpose = NULL,
                              mode = NULL)

# Initialize data frame to tabulate empty tables
empty_tables <- expand.grid(day = weekday, prps = purpose, md = mode) |>
  mutate(status = "Readable")

# Loop through purposes
for(p in 1:length(purpose)){
  # Loop through modes
  for(m in 1:length(mode)){
    # Loop through weekdays
    for(d in 1:length(weekday)){
      # Read data file for purpose/weekday
      raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
                             delim = "\t", 
                             col_names = FALSE,
                             show_col_types = FALSE)

      # Initialize temporary data frame to process the data
      od_purpose_temp <- data.frame(X1 = NULL, purpose = NULL)

      # Identify the locations of the time partitions in the table
      idx <- which(str_detect(raw_data$X1, "TABLE    :"))
      idx <- c(idx, nrow(raw_data) + 1)

      # Retrieve the information about the start time of trips
      start_time <- raw_data |>
        slice(idx) |>
        transmute(start_time = str_extract(X1, "\\d+"))

      # Run only if the length of idx is not indicative of an empty table 
      if(length(idx) > 1){

        # Loop through times to extract all useful information, discard rest
        for(i in 1:(length(idx) - 1)){
          od_purpose_temp <- rbind(od_purpose_temp,
                                   raw_data |>
                                     slice((idx[i]+2):(idx[i+1]-1)) |>
                                     mutate(start_time = start_time$start_time[i]))
        }

        # Separate the columns and convert variables

        od_purpose_temp <- od_purpose_temp |>
          separate(X1, into = c("ggh4_orig", "Trips"), sep = " (?=[^ ]+$)") |>
          mutate(ggh4_orig = str_trim(ggh4_orig, side = "both")) |>
          separate(ggh4_orig, into = c("ggh4_orig", "ggh4_dest"), sep = " (?=[^ ]+$)") |>
          mutate(ggh4_orig = str_trim(ggh4_orig),
                 Trips = as.numeric(Trips),
                 start_time = ifelse(str_length(start_time)<4, 
                                     paste0("0", start_time, ""), 
                                     start_time),
                 start_time = paste0(str_sub(start_time, 1, 2), ":", str_sub(start_time, -2, -1)),
                 start_time = hm(start_time),
                 day = weekday[d],
                 purpose = purpose[p],
                 mode = mode[m])

        # Bind the data frames
        od_purpose_time <- rbind(od_purpose_time,
                                 od_purpose_temp)
      }else{
        # Print warning for manual check
        print(glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/Table ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt is empty: check"))
        empty_tables <- empty_tables |>
          mutate(status = ifelse(day == weekday[d] & prps == purpose[p] & md == mode[m],
                                 "Not readable", status))
      }
    }  
  }
}

od_purpose_time <- od_purpose_time |>
  mutate(day = factor(day,
                      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                      ordered = TRUE),
         purpose = factor(purpose,
                          levels = c("Work", "Work2", "School", "School2", "Home", "Daycare", "Facilitate", "Market", "Other"),
                          labels = c("Work", "Subsequent work", "School", "Subsequent school", "Home", "Daycare", "Facilitate passenger", "Marketing or shopping", "Other")),
         mode = factor(mode,
                       levels = c("Transit", "Cycle", "Driver", "GO", "GO+Transit", "Motorcycle", "Other", "Passenger", "School", "Taxi", "Rideshare", "Walk"),
                       labels = c("Transit excluding GO rail", "Cycle", "Auto driver", "GO rail only", "Joint GO rail and local transit", "Motorcycle", "Other", "Auto passenger", "School bus", "Taxi passenger", "Paid rideshare", "Walk")))
```

Check empty tables:
```{r}
ggplot(data = empty_tables,
       aes(x = prps, y = md)) +
  geom_tile(aes(fill = status)) +
  facet_wrap(~ day) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

After checking the tables, I figures out that they do contain information but the format is funky. They contain a header and then a rectangular array of values like a tiny origin-destination table, instead of sensible origin-destination-value tuples.

We can use the record of unreadable tables above to deal with this tables in a different way.

```{r}
# Find the tables that were not readable
empty_tables <- empty_tables |> 
  filter(status == "Not readable")

# Initialize a data frame to deposit the data
od_purpose_temp <- data.frame(ggh4_orig = NULL, 
                              ggh4_dest = NULL, 
                              Trips = NULL, 
                              start_time = NULL, 
                              day = NULL, 
                              purpose = NULL,
                              mode = NULL)

# Loop through tables that were not successfully read using the previous protocol
for(t in 1:nrow(empty_tables)){
  # Read data file for purpose/weekday
  raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"), 
                         delim = "\t", 
                         col_names = FALSE,
                         show_col_types = FALSE,
                         skip_empty_rows = FALSE)

  # Identify the locations of the time partitions in the table
  idx <- which(str_detect(raw_data$X1, "Table: \\d+"))
  idx <- c(idx, nrow(raw_data) + 2)

  # Retrieve the information about the start time of trips
  start_time <- raw_data |>
    slice(idx) |>
    transmute(start_time = str_extract(X1, "\\d+"))

  # Run only if the length of idx is not indicative of an empty table 
  if(length(idx) > 1){
    # Loop through segments of table
    for(i in 1:(length(idx) - 1)){

      od_purpose_temp2 <- read_csv(glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"),
                                   skip = idx[i] + 1,
                                   n_max = (idx[i+1]-3) - (idx[i]+2),
                                   show_col_types = FALSE,
                                   name_repair = "minimal") |>
        rename(ggh4_orig = 1) |>
        pivot_longer(cols = -ggh4_orig,
                     names_to = "ggh4_dest",
                     values_to = "Trips") |>
        mutate(start_time = start_time$start_time[i],
               day = empty_tables$day[t],
               purpose = empty_tables$prps[t],
               mode = empty_tables$md[t])

      # Bind to table
      od_purpose_temp <- rbind(od_purpose_temp,
                               od_purpose_temp2) |>
        filter(Trips > 0)
    }
  }
}

# Finish preparing the variables
od_purpose_temp <- od_purpose_temp |>
  mutate(start_time = ifelse(str_length(start_time) < 4, 
                             paste0("0", start_time, ""), 
                             start_time),
         start_time = paste0(str_sub(start_time, 1, 2), ":", str_sub(start_time, -2, -1)),
         start_time = hm(start_time),
         day = factor(day,
                      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                      ordered = TRUE),
         purpose = factor(purpose,
                          levels = c("Work", "Work2", "School", "School2", "Home", "Daycare", "Facilitate", "Market", "Other"),
                          labels = c("Work", "Subsequent work", "School", "Subsequent school", "Home", "Daycare", "Facilitate passenger", "Marketing or shopping", "Other")),
         mode = factor(mode,
                       levels = c("Transit", "Cycle", "Driver", "GO", "GO+Transit", "Motorcycle", "Other", "Passenger", "School", "Taxi", "Rideshare", "Walk"),
                       labels = c("Transit excluding GO rail", "Cycle", "Auto driver", "GO rail only", "Joint GO rail and local transit", "Motorcycle", "Other", "Auto passenger", "School bus", "Taxi passenger", "Paid rideshare", "Walk")))
```

Bind the tables:
```{r}
od_ggh4_mode_purpose <- rbind(od_purpose_time,
                         od_purpose_temp) 
```

Summary of table:
```{r}
summary(od_ggh4_mode_purpose)
```

Make start times of trips consistent with day. Some trips are recorded on a certain day, say Monday, but with a start time that exceed midnight (i.e., 24:00). Those trips should be recorded as having been made on the following day. This code corrects this quirk of the data:
```{r}
od_ggh4_mode_purpose <- od_ggh4_mode_purpose |>
  # Convert days to numeric; the convention is that Sunday is 1
  mutate(day = case_when(day == "Monday" ~ 2,
                         day == "Tuesday" ~ 3,
                         day == "Wednesday" ~ 4,
                         day == "Thursday" ~ 5,
                         day == "Friday" ~ 6),
         # Check whether the starting time is greater than 24:00; if true, calculate the time as if it was on the following day, otherwise keep as is
         start_time = if_else(start_time > hm("24:00"), 
                              start_time - hm("24:00"), 
                              start_time),
         # Check whether the starting time is greater than 24:00; if true, change the day of the trip to the following day, otherwise keep as is
         day = if_else(start_time >= hm("24:00"), 
                       day + 1, 
                       day),
         # Convert days from numeric to weekdays
         day = wday(day,
                    label = TRUE))
```

Summary of table:
```{r}
summary(od_ggh4_mode_purpose)
```

Some trips are now recorded as having taken place on Saturday. In general, a small number of trips are changed to the following day, but now the maximum start time is 24:00.

Total number of trips in table:
```{r}
sum(od_ggh4_mode_purpose$Trips)
```

There is a small number of trips ($n = 7,368$) that where taken for an unknown purpose by any mode and on any day. There are zero trips that where undertaken by an unkown mode for any purpose on any day.

In other words, the table above is incomplete, but the number of trips that are missing is very small.

Rename columns to make it consistent with other data objects in package:
```{r}
od_ggh4_mode_purpose <- od_ggh4_mode_purpose |>
  rename(Origin = ggh4_orig,
         Destination = ggh4_dest)
```

Save the data:
```{r}
usethis::use_data(od_ggh4_mode_purpose, 
                  overwrite = TRUE)
```

<!---

-->

Load data object:
```{r}
library(TTS2016R)
data("od_ggh4_mode_purpose")
```

# Analysis

Summary of table:
```{r}
summary(od_ggh4_mode_purpose)
```

Check the total number of trips in the survey:
```{r}
sum(od_ggh4_mode_purpose$Trips)
```

Plot frequency by start time of trips:
```{r}
ggplot(data = od_ggh4_mode_purpose |>
         filter(day != "Sat") |>
         mutate(start_time = as.numeric(start_time, "hours")),
       aes(x = start_time,
           weight = Trips,
           color = day)) +
  geom_density(bw = 0.5) +
  ggtitle("Trip start time for all purposes")# +
#facet_wrap(~ purpose)
```

Plot frequency by start time of trips:
```{r}
ggplot(data = od_ggh4_mode_purpose |>
         filter(day != "Sat") |>
         mutate(start_time = as.numeric(start_time, "hours")),
       aes(x = start_time,
           weight = Trips,
           color = day)) +
  geom_freqpoly(bins = 50) +
  ggtitle("Trip start time for all purposes")# +
#facet_wrap(~ purpose)
```

Summary of trips:
```{r}
od_ggh4_mode_purpose |>
  group_by(mode, day) |>
  summarize(Trips = sum(Trips),
            .groups = "drop")
```

```{r}
ggplot(data = od_ggh4_mode_purpose |>
         filter(day != "Sat", purpose == "Home") |>
         mutate(start_time = as.numeric(start_time, "hours")),
       aes(x = start_time,
           weight = Trips,
           color = day)) +
  geom_density(bw = 0.5) +
  ggtitle("Trip start time for home")# +
#facet_wrap(~ purpose)
```

Read geography file:
```{r}
load(file = glue(here::here(), "/Geography Files/gghv5.rda"))
```

Calculate trip production for WORK:
```{r}
trip_production <- od_purpose_time |>
  filter(purpose == "Work") |>
  group_by(ggh4_orig, day) |>
  summarize(trip_production = sum(Trips),
            .groups = "drop") |>
  pivot_wider(names_from = day,
              values_from = trip_production,
              names_prefix = "trip_production_")
```

Calculate trip attraction:
```{r}
trip_attraction <- od_purpose_time |>
  filter(purpose == "Work") |>
  group_by(ggh4_dest, day) |>
  summarize(trip_attraction = sum(Trips),
            .groups = "drop") |>
  pivot_wider(names_from = day,
              values_from = trip_attraction,
              names_prefix = "trip_attraction_")
```

Join to sf object:
```{r}
gghv5 <- gghv5 |>
  left_join(trip_production |>
              mutate(ggh4_orig = as.numeric(ggh4_orig)),
            by = c("taz_v4" = "ggh4_orig")) |>
  left_join(trip_attraction |>
              mutate(ggh4_dest = as.numeric(ggh4_dest)),
            by = c("taz_v4" = "ggh4_dest"))
```

Plot trip production:
```{r}
ggplot() +
  geom_sf(data = gghv5 |>
            filter(cmaname == "Hamilton"),
          aes(fill = trip_production_Monday),
          color = NA) +
  scale_fill_distiller(name = "Trips", direction = 1, trans = "log") +
  ggtitle("Trip production for work - Monday")
theme_minimal()
```

Plot trip attraction:
```{r}
ggplot() +
  geom_sf(data = gghv5 |>
            filter(cmaname == "Hamilton"),
          aes(fill = trip_attraction_Monday),
          color = NA) +
  scale_fill_distiller(name = "Trips", direction = 1, trans = "log") +
  ggtitle("Trip attraction for work - Monday") +
  theme_minimal()
```
--->

### 2023-11-09 data-raw/TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/Extract-OD-by-Purpose-Mode-Time-Day-from-Raw-Files.Rmd
Adding preprocessing files for newest table

--- 2023-11-09 data-raw/04-GGH4-Extract-OD-by-Purpose-Mode-Time-Day-from-Raw-Files.Rmd
@@
 ```{r}
-# Parameters for testing, commented out
-d <- 1
-p <- 2
-m <- 6
+  # d <- 1
+# p <- 2
+# m <- 6
 
@@
       # Read the files in turn
-      raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
+      raw_data <- read_delim(file = glue("ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
                              delim = "\t", 
@@
       # Read data file for purpose/weekday
-      raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
+      raw_data <- read_delim(file = glue("ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt"), 
                              delim = "\t", 
@@
         # Print warning for manual check
-        print(glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/Table ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt is empty: check"))
+        print(glue("Table ttsCross-OD-GGH4-{purpose[p]}-{weekday[d]}-{mode[m]}.txt is empty: check"))
         empty_tables <- empty_tables |>
@@
   # Read data file for purpose/weekday
-  raw_data <- read_delim(file = glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"), 
+  raw_data <- read_delim(file = glue("ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"), 
                          delim = "\t", 
@@
 
-      od_purpose_temp2 <- read_csv(glue("TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"),
+      od_purpose_temp2 <- read_csv(glue("ttsCross-OD-GGH4-{empty_tables$prps[t]}-{empty_tables$day[t]}-{empty_tables$md[t]}.txt"),
                                    skip = idx[i] + 1,
@@
 ```{r}
-od_ggh4_mode_purpose <- rbind(od_purpose_time,
+od_purpose_time <- rbind(od_purpose_time,
                          od_purpose_temp) 
-```
-
-Summary of table:
-```{r}
-summary(od_ggh4_mode_purpose)
-```
-
-Make start times of trips consistent with day.
-Some trips are recorded on a certain day, say Monday, but with a start time that exceed midnight (i.e., 24:00).
-Those trips should be recorded as having been made on the following day.
-This code corrects this quirk of the data:
-```{r}
-od_ggh4_mode_purpose <- od_ggh4_mode_purpose |>
-  # Convert days to numeric; the convention is that Sunday is 1
-  mutate(day = case_when(day == "Monday" ~ 2,
-                         day == "Tuesday" ~ 3,
-                         day == "Wednesday" ~ 4,
-                         day == "Thursday" ~ 5,
-                         day == "Friday" ~ 6),
-         # Check whether the starting time is greater than 24:00; if true, calculate the time as if it was on the following day, otherwise keep as is
-         start_time = if_else(start_time > hm("24:00"), 
-                              start_time - hm("24:00"), 
-                              start_time),
-         # Check whether the starting time is greater than 24:00; if true, change the day of the trip to the following day, otherwise keep as is
-         day = if_else(start_time >= hm("24:00"), 
-                       day + 1, 
-                       day),
-         # Convert days from numeric to weekdays
-         day = wday(day,
-                    label = TRUE))
-```
-
-Summary of table:
-```{r}
-summary(od_ggh4_mode_purpose)
-```
-
-Some trips are now recorded as having taken place on Saturday. In general, a small number of trips are changed to the following day, but now the maximum start time is 24:00.
-
-Total number of trips in table:
-```{r}
-sum(od_ggh4_mode_purpose$Trips)
-```
-
-There is a small number of trips ($n = 7,368$) that where taken for an unknown purpose by any mode and on any day.
-There are zero trips that where undertaken by an unkown mode for any purpose on any day.
-
-In other words, the table above is incomplete, but the number of trips that are missing is very small.
-
-Rename columns to make it consistent with other data objects in package:
-```{r}
-od_ggh4_mode_purpose <- od_ggh4_mode_purpose |>
-  rename(Origin = ggh4_orig,
-         Destination = ggh4_dest)
-```
-
-Save the data:
-```{r}
-usethis::use_data(od_ggh4_mode_purpose, 
-                  overwrite = TRUE)
-```
-
-<!---
-
--->
-
-Load data object:
-```{r}
-library(TTS2016R)
-data("od_ggh4_mode_purpose")
 ```
@@
 ```{r}
-summary(od_ggh4_mode_purpose)
+summary(od_purpose_time)
 ```
@@
 ```{r}
-sum(od_ggh4_mode_purpose$Trips)
+sum(od_purpose_time$Trips)
 ```
@@
 ```{r}
-ggplot(data = od_ggh4_mode_purpose |>
-         filter(day != "Sat") |>
+ggplot(data = od_purpose_time |>
          mutate(start_time = as.numeric(start_time, "hours")),
@@
            color = day)) +
-  geom_density(bw = 0.5) +
+  geom_density(adjust = 1) +
   ggtitle("Trip start time for all purposes")# +
@@
 
-Plot frequency by start time of trips:
 ```{r}
-ggplot(data = od_ggh4_mode_purpose |>
-         filter(day != "Sat") |>
+ggplot(data = od_purpose_time |>
+         filter(purpose == "Home") |>
          mutate(start_time = as.numeric(start_time, "hours")),
@@
            color = day)) +
-  geom_freqpoly(bins = 50) +
-  ggtitle("Trip start time for all purposes")# +
-#facet_wrap(~ purpose)
-```
-
-Summary of trips:
-```{r}
-od_ggh4_mode_purpose |>
-  group_by(mode, day) |>
-  summarize(Trips = sum(Trips),
-            .groups = "drop")
-```
-
-```{r}
-ggplot(data = od_ggh4_mode_purpose |>
-         filter(day != "Sat", purpose == "Home") |>
-         mutate(start_time = as.numeric(start_time, "hours")),
-       aes(x = start_time,
-           weight = Trips,
-           color = day)) +
-  geom_density(bw = 0.5) +
+  geom_density(adjust = 2) +
   ggtitle("Trip start time for home")# +
@@
 ```
-
 Read geography file:
@@
 ```
---->

### 2023-11-24 data-raw/05-TTS16-Calculating-Travel-Time-OD-GGH4.Rmd
Updated with code to write od_ggh4_mode_purpose to arrow parquet

--- 2022-10-22 data-raw/01-TTS16-Calculating-Travel-Time-OD.Rmd
@@
 ```{r setup, include=FALSE}
-library(disk.frame)
-library(dplyr)
-library(ggplot2)
-library(gridExtra)
-library(httr)
-library(knitr)
-#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
-library(kableExtra)
-# install.packages("devtools")
-# install.packages("Rtools")
-# devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
-library(lodes)
-library(progress)
-library(purrr)
-library(r5r)
-library(sf)
-library(stplanr)
-library(tidycensus)
-library(tidyr)
-library(tmap)
+library(disk.frame) # not installed on this machine
+library(dplyr) # A Grammar of Data Manipulation
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
+library(httr) # Tools for Working with URLs and HTTP
+library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
+library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
+library(progress) # Terminal Progress Bars
+library(purrr) # Functional Programming Tools
+library(sf) # Simple Features for R
+library(stplanr) # not installed on this machine
+library(tidycensus) # not installed on this machine
+library(tidyr) # Tidy Messy Data
+library(tmap) # Thematic Maps
+library(TTS2016R)
 library(zoo) # for rollapplyr
@@
 setup_disk.frame() # setup disk.frame parallel computing
-options(future.globals.maxSize = Inf) # allow disk.frame threads to talk to eachother without limits
-options(java.parameters = "-Xmx6G") # assign memory for r5r to work with
+options(future.globals.maxSize = Inf) # allow disk.frame threads to talk to each other without limits
 tmap_mode("plot")
@@
 
+```{r}
+options(java.parameters = "-Xmx30G") # assign memory for r5r to work with
+library(r5r) # Rapid Realistic Routing with 'R5' 
+```
+
 ##--- Prepare Input Data
-loading the destination points and origins points:
+
+Prepare destination points and origins points:
 ```{r}
-load(file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/work_origins.Rdata")
-load(file = "TTS16-data-inputs/Travel-Time-Calculations/inputs/job_destinations.Rdata")
+data("ggh_taz_v4")
+
+coords <- ggh_taz_v4 |>
+  st_transform(crs = 4326) |>
+  st_centroid() |>
+  st_coordinates()
+
+work_origins <- ggh_taz_v4 |>
+  st_drop_geometry() |>
+  transmute(id = GGHV4,
+            lon = coords[,1],
+            lat = coords[,2])
+
+job_destinations <- work_origins
 ```
@@
 
+Version of data downloaded:
+
 ```{r download data, include=FALSE, eval=FALSE}
 #downloading ontario osm in the correct format
-download.file(url = paste0("https://download.geofabrik.de/north-america/canada/ontario-latest.osm.pbf"),
+download.file(url = paste0("https://download.openstreetmap.fr/extracts/north-america/canada/ontario/golden_horseshoe-latest.osm.pbf"),
               destfile = file.path(r5_path, "osm.pbf"),
@@
 ##--- Set Up R5 Routing
+
+<!-- QUESTION FOR RAFA ET AL. IS IT POSSIBLE TO HAVE MULTIPLE osm.pbf FILES IN THE SAME DIRECTORY AND CALL THEM BY NAME? -->
 

### 2024-11-18 data-raw/02-TTS16-Imp-Function-Trip-Length-Distribution.Rmd
Commit to move from Windows

@@
 ```{r}
-library(ggplot2)
-library(dplyr)
-library(MASS)
-
-# install.packages("fitdistrplus")
-# install.packages("logspline")
-library(fitdistrplus)
-library(logspline)
+library(dplyr) # A Grammar of Data Manipulation
+library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(logspline) # Routines for Logspline Density Estimation
+library(MASS) # Support Functions and Datasets for Venables and Ripley's MASS
 
@@
 ##-- Exploring the OD travel time data
+
+<!--FILE NOT THERE!!-->
 

### 2024-11-18 data-raw/04-GGH4-Extract-OD-by-Purpose-Mode-Time-Day-from-Raw-Files (SFConflict paezha@mcmaster.ca 2024-01-20-14-51-34).Rmd
Commit to move from Windows

--- 2023-11-09 data-raw/04-GGH4-Extract-OD-by-Purpose-Mode-Time-Day-from-Raw-Files.Rmd
@@
 
+Save to parquet for sharing:
+```{r}
+library(arrow)
+# using compression
+if (codec_is_available("gzip")) {
+  tf <- tempfile(tmpdir = glue::glue(stringr::str_replace_all(here::here(), "/", "\\\\"),
+                                     "\\data-raw\\TTS16-data-inputs\\OD-GGH4-by-Time-Day-Mode-Purpose\\"),
+                 fileext = ".gz.parquet")
+  write_parquet(od_ggh4_mode_purpose,
+                tf, 
+                compression = "gzip", 
+                compression_level = 5)
+}
+```
+
+Rename file:
+```{r}
+arrow_file <- list.files(path = glue::glue(here::here(), 
+                             "/data-raw/TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose"),
+           pattern = "^(.*)\\.gz.parquet$")
+# arrow_file <- stringr::str_replace(arrow_file,
+#                                    stringr::str_extract(arrow_file, "[^.]+"),
+#                                    glue::glue(, ""))
+
+file.rename(from = glue::glue(here::here(), 
+                             "/data-raw/TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/{arrow_file}"), 
+            to = glue::glue(here::here(), 
+                             "/data-raw/TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/od_ggh4_mode_purpose.gz.parquet"))
+```
+
+Remove data object:
+```{r}
+rm(od_ggh4_mode_purpose)
+```
+
+Check that parquet file can be read:
+```{r}
+od_ggh4_mode_purpose <- read_parquet(glue::glue(here::here(), 
+                             "/data-raw/TTS16-data-inputs/OD-GGH4-by-Time-Day-Mode-Purpose/od_ggh4_mode_purpose.gz.parquet"),
+             as_data_frame = FALSE)
+```
+
 <!---
-
--->
 
